---
title: "Rearrange EU decay rate for plotting"
output: html_document
author: "YM"
date: "2025-10-01"
---

```{r setup}
rm(list=ls())
library(broom)
library(tidyverse)
library(nls.multstart)
library(ggplot2)
library(patchwork)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) #assumes we are in scripts folder
#knitr::opts_knit$set(root.dir = "/Users/s2009817/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Swaffer lab documents/NGS_metadata/summary_figures/EU_combined_analysis/scripts")
load("plotting_functions.rda")
load('fitting_function.rda')
#clipr::write_clip()
```

## Plotting functions 

```{r}
plot_correlation <- function(data, x, y, z=NULL, z_lim=c(), corr_type="spearman", x_lab=NULL, y_lab=NULL, color_lab=NULL, log2_x=FALSE, log2_y=FALSE, log2_z=FALSE)
  #z: the coloring scheme; optional, if not specified the scatter is all in royalblue3
  #corr_type: "spearman" or "pearson"
  #x_lab, y_lab, color_lab: axis labels; take a string
  #log2_x, log2_y, log2_z: plot x, y or colour axis orin log2 scale or not; boolean value
  {load_pkg("viridis")
  
  if (grepl("\\$", deparse(substitute(x)))) {x_lab <- strsplit(deparse(substitute(x)), "\\$")[[1]][2]}
  if (grepl("\\$", deparse(substitute(y)))) {y_lab <- strsplit(deparse(substitute(y)), "\\$")[[1]][2]}
  if (grepl("\\$", deparse(substitute(z)))) {color_lab <- strsplit(deparse(substitute(z)), "\\$")[[1]][2]}
  x_lab_suffix = NULL
  if (log2_x==TRUE) {x[x <= 0 & !is.na(x)] <- NA
  x <- log2(x)
  x_lab_suffix = "(log2 scale)"}
  y_lab_suffix = NULL
  if (log2_y==TRUE) {y[y <= 0 & !is.na(y)] <- NA
  y <- log2(y)
  y_lab_suffix = "(log2 scale)"}
  color_lab_suffix = NULL
  if (log2_z==TRUE) {z[z <= 0 & !is.na(y)] <- NA
  z <- log2(z)
  color_lab_suffix = "(log2 scale)"}
  labs <- labs(x = paste0(gsub("_", " ", x_lab),"\n", x_lab_suffix), 
               y = paste0(gsub("_", " ", y_lab),"\n", y_lab_suffix),
               color = paste0(str_wrap(gsub("_", " ", color_lab), width = 9),"\n", color_lab_suffix))

  bin_sum <- Hmisc::cut2(x, g=50) %>% bind_cols(., x, y) %>% dplyr::rename("bins"=1,"x"=2,"y"=3) %>% drop_na() %>% dplyr::group_by(bins) %>% summarise(sd=sd(y, na.rm = TRUE), x=mean(x), y=mean(y), n = n())
  
  binned_line <- bin_sum %>% geom_line(mapping = aes(x=(x), y=(y)))

  annotate_r <- annotate(geom = 'text', label =  paste("R =", (corrr::correlate(x=x, y=y, method=corr_type)$x %>% round(digits = 3))), x=-Inf, y=Inf, hjust=-0.1, vjust=2)

  annotate_n <- geom_text(aes(label = paste("n =", sum(!is.na(x) & !is.na(y))), x=Inf, y=-Inf), hjust = 1.2, vjust = -1.2)
  
  if (missing(z)) {
    points <- data %>% geom_point(mapping = aes(x=(x), y=(y)), size=1.8, alpha=0.2, color = "royalblue3")
    p <- ggplot() + points + annotate_r + annotate_n + binned_line + geom_abline(slope=1, intercept = 0, colour = "grey") + labs}
    else {
      points <- data %>% geom_point(mapping = aes(x=(x), y=(y), color=(z)),  size=1.8, alpha=0.3)
      color <- viridis::scale_color_viridis(begin=0.2, end=1, limits=z_lim, oob=scales::squish)
      p <- ggplot() + points + annotate_r + annotate_n + color + binned_line + geom_abline(slope=1, intercept = 0, colour = "grey") + labs}
   
return(p)
}

batch_plot <- function(dataframe, x, y, z = NULL, z_lim = NULL, corr_type, log2_x = NULL, log2_y = NULL, log2_z = NULL) 
#x and y: x and y axis of scatter plot; a vector where elements must be column names in "dataframe"
#z: the coloring scheme; optional, if not specified the scatter is all in royalblue3; if specified needs to be a vector (elements must be column names in "dataframe") of the same length as x and y, put NA if coloring is not needed for individual plots
#corr_type: a vector of the same length as x and y, where elements must be "spearman" or "pearson"
#log2_x, log2_y, log2_z:  optional, if not specified all plots are with linear axis; if specified must be a vector of the same length as x and y, where elements must be TRUE or FALSE
  {
  if (is.null(log2_x)) {log2_x <- rep("FALSE", length(x))}
  if (is.null(log2_y)) {log2_y <- rep("FALSE", length(y))}
  if (is.null(log2_z)) {log2_z <- rep("FALSE", length(z))}

  #the case with no z input
  if (is.null(z)) {pairs <- tibble::tibble(x = x, y = y, corr_type = corr_type, log2_x, log2_y)
    pmap(.l = pairs, .f = function(x, y, corr_type, log2_x, log2_y) 
    {plot_correlation(data = dataframe, x = dataframe[[x]], y = dataframe[[y]], corr_type = corr_type, 
                        x_lab = x, y_lab = y, log2_x = log2_x, log2_y = log2_y)})}
  #the case with z input but no z_lim specified
  else if (is.null(z_lim)) {pairs <- tibble::tibble(x = x, y = y, z = z, corr_type = corr_type, log2_x, log2_y, log2_z)
    pmap(.l = pairs, .f = function(x, y, z, corr_type, log2_x, log2_y, log2_z) {
      if (is.na(z)) {plot_correlation(data = dataframe, x = dataframe[[x]], y = dataframe[[y]], corr_type = corr_type, x_lab = x, y_lab = y, log2_x = log2_x, log2_y = log2_y)}
      else {plot_correlation(data = dataframe, x = dataframe[[x]], y = dataframe[[y]], z = dataframe[[z]], corr_type = corr_type, x_lab = x, y_lab = y, color_lab = z, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)}
    })}
  #the case with z input and also z_lim specified
  else {pairs <- tibble::tibble(x = x, y = y, z = z, z_lim = z_lim, corr_type = corr_type, log2_x, log2_y, log2_z)
    pmap(.l = pairs, .f = function(x, y, z, z_lim, corr_type, log2_x, log2_y, log2_z) {
      if (is.na(z)) {plot_correlation(data = dataframe, x = dataframe[[x]], y = dataframe[[y]], corr_type = corr_type, x_lab = x, y_lab = y, log2_x = log2_x, log2_y = log2_y)}
      else {plot_correlation(data = dataframe, x = dataframe[[x]], y = dataframe[[y]], z = dataframe[[z]], z_lim = z_lim, corr_type = corr_type, x_lab = x, y_lab = y, color_lab = z, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)}
    })}
}
```

## Combine all current expression level datasets and half-life datasets in long format, and attach sample info (e.g. experiment_date, strain, size, condition) for filtering

```{r}
sample_info <- read_csv("../tables/sample_info_combined.csv", show_col_types=FALSE)
verified_ORF <- read_tsv("../tables/2230128_verified_ORF.txt", show_col_types=FALSE) %>% filter(species == "S_cerevisiae")

#translation efficiency from literature
df_TE <- read_csv("../tables/translation_efficiency.csv", show_col_types=FALSE) %>% 
  mutate(TE_MandP_YM_fc = if_else(TE_MandP_small_YM > 0 & TE_MandP_large_YM > 0, 
                                  TE_MandP_large_YM / TE_MandP_small_YM, NA_real_),
         TE_P_YM_fc = if_else(TE_P_large_YM > 0 & TE_P_small_YM > 0, TE_P_large_YM / TE_P_small_YM, NA_real_))

#gene fraction and tail length from Nanopore seq
df_ONT <-  read_csv("../tables/gene_fraction&tail_length.csv", show_col_types=FALSE)

#take mean input TPM as expression level / mRNA abundance
df_input_long <- read_tsv("../tables/df_sc_combined_EU_rescaled.txt", show_col_types=FALSE) %>%
  dplyr::select(gene_id, experiment_strain_condition, TPM_input) %>%
  dplyr::distinct() %>%
  left_join(., sample_info, join_by(experiment_strain_condition))

#take mRNA half lives
hl_files <- dir(path="../output_tables/", pattern=".*output.*\\.txt", full.names=TRUE)
df_half_lives_long <- hl_files %>% map(~read_tsv(., show_col_types=FALSE)) %>%
  set_names(substr(basename(hl_files), 
                   nchar(basename(hl_files))-5,
                   nchar(basename(hl_files))-4)) %>% 
  bind_rows(., .id = "fitting_strategy") %>%
  dplyr::select(2,3,1,estimate_half_life) %>%
  left_join(., sample_info, join_by(experiment_strain_condition))
```

## Transform to wide format and calculate some averages between replicates

```{r}
df_input <- df_input_long %>%
  #filter(condition=="WT") %>%
  pivot_wider(id_cols = gene_id,
              names_from = label,
              values_from = TPM_input,
              names_glue = "{label}_{.value}") %>%
  mutate(WT_small_TPM_input = rowMeans(across(c(MS96_small_TPM_input,MS1039_small_TPM_input)), na.rm = TRUE),
         WT_large_TPM_input = rowMeans(across(c(MS96_large_TPM_input,MS1039_large_TPM_input)), na.rm = TRUE),
         MS908_small_eIF4G_dep_1hr_ave_TPM_input = 
           rowMeans(across(c(MS908_small_eIF4G_dep_1hr_rep1_TPM_input,
                             MS908_small_eIF4G_dep_1hr_rep2_TPM_input)), na.rm = TRUE),
         MS908_large_eIF4G_dep_1hr_ave_TPM_input = 
           rowMeans(across(c(MS908_large_eIF4G_dep_1hr_rep1_TPM_input,
                             MS908_large_eIF4G_dep_1hr_rep2_TPM_input)), na.rm = TRUE),
         eIF4G_dep_1hr_small_TPM_input = 
           rowMeans(across(c(MS908_small_eIF4G_dep_1hr_rep1_TPM_input,
                             MS908_small_eIF4G_dep_1hr_rep2_TPM_input,
                             MS1039_small_eIF4G_dep_1hr_TPM_input)), na.rm = TRUE),
         eIF4G_dep_1hr_large_TPM_input = 
           rowMeans(across(c(MS908_large_eIF4G_dep_1hr_rep1_TPM_input,
                             MS908_large_eIF4G_dep_1hr_rep2_TPM_input,
                             MS1039_large_eIF4G_dep_1hr_TPM_input)), na.rm = TRUE))
df_half_lives <- df_half_lives_long %>% 
  #filter(condition=="WT") %>%
  filter(fitting_strategy=="c1") %>%
  pivot_wider(id_cols = gene_id,
              names_from = c(label, fitting_strategy), 
              values_from = estimate_half_life,
              names_glue = "{label}_{.value}_by_{fitting_strategy}") %>%
  mutate(WT_small_estimate_half_life_by_c1 = 
           rowMeans(across(c(MS96_small_estimate_half_life_by_c1, 
                             MS1039_small_estimate_half_life_by_c1)), na.rm = TRUE),
         WT_large_estimate_half_life_by_c1 = 
           rowMeans(across(c(MS96_large_estimate_half_life_by_c1, 
                             MS1039_large_estimate_half_life_by_c1)), na.rm = TRUE),
         MS908_small_eIF4G_dep_1hr_ave_estimate_half_life_by_c1 =
           rowMeans(across(c(MS908_small_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1,
                             MS908_small_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1)),na.rm = TRUE),
         MS908_large_eIF4G_dep_1hr_ave_estimate_half_life_by_c1 = 
           rowMeans(across(c(MS908_large_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1,
                             MS908_large_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1)), na.rm = TRUE),
         eIF4G_dep_1hr_small_estimate_half_life_by_c1=
           rowMeans(across(c(MS908_small_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1,
                             MS908_small_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1,
                             MS1039_small_eIF4G_dep_1hr_estimate_half_life_by_c1)), na.rm = TRUE),
         eIF4G_dep_1hr_large_estimate_half_life_by_c1=
           rowMeans(across(c(MS908_large_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1,
                             MS908_large_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1,
                             MS1039_large_eIF4G_dep_1hr_estimate_half_life_by_c1)), na.rm = TRUE))
```

## Calculate half-life fold change large to small

```{r}
sample_info_fc <- sample_info %>% 
  dplyr::select(experiment_date, strain, condition) %>% 
  dplyr::distinct() %>% 
  bind_rows(., tibble(experiment_date = rep(NA, 2), strain = rep("Average", 2), 
                      condition = c("WT","eIF4G_dep_1hr"))) %>% 
  mutate(label = paste0(strain, "_", condition),
         label = if_else(replace_na(experiment_date == "241204", FALSE), "MS908_eIF4G_dep_1hr_rep1", label),
         label = if_else(replace_na(experiment_date == "250423", FALSE), "MS908_eIF4G_dep_1hr_rep2", label))

df_half_lives_sb <- df_half_lives_long %>% 
  filter(fitting_strategy=="c1") %>%
  pivot_wider(id_cols = c(gene_id, strain, condition, experiment_date, fitting_strategy),
              names_from = size, 
              values_from = estimate_half_life,
              names_glue = "{size}_{.value}") %>% 
  #Combine averages of replicates
  bind_rows(., df_half_lives %>% 
              dplyr::select(gene_id, WT_small_estimate_half_life_by_c1, WT_large_estimate_half_life_by_c1,
                            eIF4G_dep_1hr_small_estimate_half_life_by_c1,
                            eIF4G_dep_1hr_large_estimate_half_life_by_c1) %>% 
              pivot_longer(cols = -gene_id, 
                           names_to = c("condition", "size", "fitting_strategy"),
                           names_pattern = "^(.+?)_(small|large)_estimate_half_life_by_(c\\d+)$",
                           values_to = "estimate_half_life") %>% 
              pivot_wider(id_cols = c(gene_id, condition, fitting_strategy),
                          names_from = size, 
                          values_from = estimate_half_life,
                          names_glue = "{size}_{.value}") %>% 
              mutate(strain = "Average", experiment_date = NA)) %>% 
  mutate(estimate_half_life_fc = large_estimate_half_life / small_estimate_half_life) 
```

##Calculate half-life fold change 1hr eIF4G depletion to WT

```{r}
df_half_lives_fc <- df_half_lives %>% 
  mutate(eIF4G_dep_1hr_ave_to_WT_small_half_lives =  eIF4G_dep_1hr_small_estimate_half_life_by_c1 /
                                           WT_small_estimate_half_life_by_c1,
         eIF4G_dep_1hr_ave_to_WT_large_half_lives = eIF4G_dep_1hr_large_estimate_half_life_by_c1 /
                                           WT_large_estimate_half_life_by_c1,
         eIF4G_dep_1hr_MS1039_to_WT_small_half_lives = MS1039_small_eIF4G_dep_1hr_estimate_half_life_by_c1 / 
                                           WT_small_estimate_half_life_by_c1,
         eIF4G_dep_1hr_MS1039_to_WT_large_half_lives = MS1039_large_eIF4G_dep_1hr_estimate_half_life_by_c1 / 
                                           WT_large_estimate_half_life_by_c1) %>% 
  dplyr::select(gene_id, eIF4G_dep_1hr_ave_to_WT_small_half_lives, eIF4G_dep_1hr_ave_to_WT_large_half_lives, 
                eIF4G_dep_1hr_MS1039_to_WT_small_half_lives, eIF4G_dep_1hr_MS1039_to_WT_large_half_lives) %>% 
  left_join(., df_input %>% mutate(eIF4G_dep_1hr_to_WT_small_input = eIF4G_dep_1hr_small_TPM_input / 
                                     WT_small_TPM_input, 
                                   eIF4G_dep_1hr_to_WT_large_input = eIF4G_dep_1hr_large_TPM_input / 
                                     WT_large_TPM_input) %>%
              dplyr::select(gene_id, 
                            eIF4G_dep_1hr_to_WT_small_input, eIF4G_dep_1hr_to_WT_large_input), join_by(gene_id))
```

## Combine all datasets

```{r}
df_combined <- left_join(df_half_lives, df_input, join_by(gene_id)) %>% #join input to half-lives, since half lives from c1 are from genes with input TPM>5, extremely lowly expressed genes are excluded from following analysis
  left_join(., df_half_lives_sb %>% 
              left_join(., sample_info_fc, join_by(experiment_date, strain, condition)) %>% 
              pivot_wider(id_cols = gene_id,
                          names_from = c(label, fitting_strategy), 
                          values_from = estimate_half_life_fc,
                          names_glue = "{label}_{.value}_by_{fitting_strategy}"), join_by(gene_id)) %>% #join fold-change large to small
  left_join(., df_half_lives_fc, join_by(gene_id)) %>% #join fold-change 1hr eIF4G depletion to WT
  left_join(., df_TE %>% 
              dplyr::select(c(gene_id, contains("mean_norm"), TE_WT_PP, TE_WT_RS, contains("YM"))), 
            join_by(gene_id)) %>% #join translation efficiency
  left_join(., df_ONT, join_by(gene_id))
```


## Check correlation between WT replicates (MS96 & MS1039 DMSO)

```{r}
x = c("MS96_small_estimate_half_life_by_c1", 
      "MS96_large_estimate_half_life_by_c1", 
      "MS96_small_TPM_input", 
      "MS96_large_TPM_input")
y = c("MS1039_small_estimate_half_life_by_c1", 
      "MS1039_large_estimate_half_life_by_c1", 
      "MS1039_small_TPM_input", 
      "MS1039_large_TPM_input")
corr_type = rep("spearman", 4)
log2_x = c("FALSE", "FALSE", "TRUE", "TRUE")
log2_y = c("FALSE", "FALSE", "TRUE", "TRUE")

batch_p <- batch_plot(df_combined, x, y, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y)
wrap_plots(batch_p, ncol = 2) %>% ggsave("../figures/WT_correlation.png", ., width = 6.25, height = 6, dpi = 300) 
wrap_plots(batch_p, ncol = 2)
```

## Check correlation between 1hr eIF4G depletion replicates (MS908 & MS1039)

```{r}
x = c("MS908_small_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1",
      "MS1039_small_eIF4G_dep_1hr_estimate_half_life_by_c1", 
      "MS1039_small_eIF4G_dep_1hr_estimate_half_life_by_c1", 
      "MS1039_small_eIF4G_dep_1hr_estimate_half_life_by_c1",
      "MS908_large_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1",
      "MS1039_large_eIF4G_dep_1hr_estimate_half_life_by_c1", 
      "MS1039_large_eIF4G_dep_1hr_estimate_half_life_by_c1", 
      "MS1039_large_eIF4G_dep_1hr_estimate_half_life_by_c1", 
      "MS908_small_eIF4G_dep_1hr_rep1_TPM_input",
      "MS1039_small_eIF4G_dep_1hr_TPM_input",
      "MS1039_small_eIF4G_dep_1hr_TPM_input",
      "MS1039_small_eIF4G_dep_1hr_TPM_input",
      "MS908_large_eIF4G_dep_1hr_rep1_TPM_input",
      "MS1039_large_eIF4G_dep_1hr_TPM_input",
      "MS1039_large_eIF4G_dep_1hr_TPM_input",
      "MS1039_large_eIF4G_dep_1hr_TPM_input")
y = c("MS908_small_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1",
      "MS908_small_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1",
      "MS908_small_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1",
      "MS908_small_eIF4G_dep_1hr_ave_estimate_half_life_by_c1",
      "MS908_large_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1",
      "MS908_large_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1",
      "MS908_large_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1",
      "MS908_large_eIF4G_dep_1hr_ave_estimate_half_life_by_c1",
      "MS908_small_eIF4G_dep_1hr_rep2_TPM_input",
      "MS908_small_eIF4G_dep_1hr_rep1_TPM_input",
      "MS908_small_eIF4G_dep_1hr_rep2_TPM_input",
      "MS908_small_eIF4G_dep_1hr_ave_TPM_input",
      "MS908_large_eIF4G_dep_1hr_rep2_TPM_input",
      "MS908_large_eIF4G_dep_1hr_rep1_TPM_input",
      "MS908_large_eIF4G_dep_1hr_rep2_TPM_input",
      "MS908_large_eIF4G_dep_1hr_ave_TPM_input")
z = c(rep("WT_small_TPM_input", 4), rep("WT_large_TPM_input", 4), rep(NA, 8))
corr_type = rep("spearman", 16)
log2_x = c(rep("FALSE", 8), rep("TRUE", 8))
log2_y = c(rep("FALSE", 8), rep("TRUE", 8))
log2_z = c(rep("TRUE", 8), rep("FALSE", 8))

batch_p <- batch_plot(df_combined, x, y, z, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)
wrap_plots(batch_p, ncol = 4) %>% ggsave("../figures/eIF4G_1hr_dep_correlation.png", ., width = 22.5, height = 17.5, dpi = 300)
wrap_plots(batch_p, ncol = 4)
```

## Correlation between WT and 30min Xrn1 depletion

```{r}
x = c("WT_small_estimate_half_life_by_c1", 
      "WT_large_estimate_half_life_by_c1", 
      "WT_small_TPM_input", 
      "WT_large_TPM_input")
y = c("MS917_small_Xrn1_dep_30min_estimate_half_life_by_c1", 
      "MS917_large_Xrn1_dep_30min_estimate_half_life_by_c1", 
      "MS917_small_Xrn1_dep_30min_TPM_input", 
      "MS917_large_Xrn1_dep_30min_TPM_input")
z = c("WT_small_TPM_input", "WT_large_TPM_input", NA, NA)
corr_type = rep("spearman", 4)
log2_x = c("FALSE", "FALSE", "TRUE", "TRUE")
log2_y = c("FALSE", "FALSE", "TRUE", "TRUE")
log2_z = c("TRUE", "TRUE", "FALSE", "FALSE")

batch_p <- batch_plot(df_combined, x, y, z, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)
wrap_plots(batch_p, ncol = 2) %>% ggsave("../figures/WT_vs_Xrn1.png", ., width = 10.4, height = 8.3, dpi = 300)
wrap_plots(batch_p, ncol = 2)
```

## Correlation between WT and eIF4G depletion

```{r}
x = c("WT_small_estimate_half_life_by_c1",
      "WT_small_estimate_half_life_by_c1", 
      "WT_small_estimate_half_life_by_c1", 
      "MS1039_small_eIF4G_dep_30min_estimate_half_life_by_c1",
      "MS1039_small_eIF4G_dep_30min_estimate_half_life_by_c1",
      "WT_large_estimate_half_life_by_c1",
      "WT_large_estimate_half_life_by_c1", 
      "WT_large_estimate_half_life_by_c1", 
      "MS1039_large_eIF4G_dep_30min_estimate_half_life_by_c1",
      "MS1039_large_eIF4G_dep_30min_estimate_half_life_by_c1",
      "WT_small_TPM_input",
      "WT_small_TPM_input",
      "WT_small_TPM_input",
      "MS1039_small_eIF4G_dep_30min_TPM_input",
      "MS1039_small_eIF4G_dep_30min_TPM_input",
      "WT_large_TPM_input",
      "WT_large_TPM_input",
      "WT_large_TPM_input",
      "MS1039_large_eIF4G_dep_30min_TPM_input",
      "MS1039_large_eIF4G_dep_30min_TPM_input")
y = c("MS1039_small_eIF4G_dep_30min_estimate_half_life_by_c1",
      "MS1039_small_eIF4G_dep_1hr_estimate_half_life_by_c1",
      "eIF4G_dep_1hr_small_estimate_half_life_by_c1",
      "MS1039_small_eIF4G_dep_1hr_estimate_half_life_by_c1",
      "eIF4G_dep_1hr_small_estimate_half_life_by_c1",
      "MS1039_large_eIF4G_dep_30min_estimate_half_life_by_c1",
      "MS1039_large_eIF4G_dep_1hr_estimate_half_life_by_c1",
      "eIF4G_dep_1hr_large_estimate_half_life_by_c1",
      "MS1039_large_eIF4G_dep_1hr_estimate_half_life_by_c1",
      "eIF4G_dep_1hr_large_estimate_half_life_by_c1",
      "MS1039_small_eIF4G_dep_30min_TPM_input",
      "MS1039_small_eIF4G_dep_1hr_TPM_input",
      "eIF4G_dep_1hr_small_TPM_input",
      "MS1039_small_eIF4G_dep_1hr_TPM_input",
      "eIF4G_dep_1hr_small_TPM_input",
      "MS1039_large_eIF4G_dep_30min_TPM_input",
      "MS1039_large_eIF4G_dep_1hr_TPM_input",
      "eIF4G_dep_1hr_large_TPM_input",
      "MS1039_large_eIF4G_dep_1hr_TPM_input",
      "eIF4G_dep_1hr_large_TPM_input")
z = c(rep("WT_small_TPM_input", 5), rep("WT_large_TPM_input", 5), rep(NA, 10))
corr_type = rep("spearman", 20)
log2_x = c(rep("FALSE", 10), rep("TRUE", 10))
log2_y = c(rep("FALSE", 10), rep("TRUE", 10))
log2_z = c(rep("TRUE", 10), rep("FALSE", 10))

batch_p <- batch_plot(df_combined, x, y, z, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)
wrap_plots(batch_p, ncol = 5) %>%  ggsave("../figures/WT_vs_eIF4G.png", ., width = 27, height = 17, dpi = 300)
wrap_plots(batch_p, ncol = 5)
```

## Correlation between small and large

```{r}
x = c("MS96_small_estimate_half_life_by_c1",
      "MS1039_small_estimate_half_life_by_c1",
      "WT_small_estimate_half_life_by_c1",
      "MS917_small_Xrn1_dep_30min_estimate_half_life_by_c1",
      "MS1039_small_eIF4G_dep_30min_estimate_half_life_by_c1", 
      "MS908_small_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1",
      "MS908_small_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1",
      "MS908_small_eIF4G_dep_1hr_ave_estimate_half_life_by_c1",
      "MS1039_small_eIF4G_dep_1hr_estimate_half_life_by_c1",
      "eIF4G_dep_1hr_small_estimate_half_life_by_c1", 
      "MS96_small_TPM_input", 
      "MS1039_small_TPM_input", 
      "WT_small_TPM_input",
      "MS917_small_Xrn1_dep_30min_TPM_input", 
      "MS1039_small_eIF4G_dep_30min_TPM_input", 
      "MS908_small_eIF4G_dep_1hr_rep1_TPM_input",
      "MS908_small_eIF4G_dep_1hr_rep2_TPM_input",
      "MS908_small_eIF4G_dep_1hr_ave_TPM_input", 
      "MS1039_small_eIF4G_dep_1hr_TPM_input", 
      "eIF4G_dep_1hr_small_TPM_input")
y = c("MS96_large_estimate_half_life_by_c1", 
      "MS1039_large_estimate_half_life_by_c1", 
      "WT_large_estimate_half_life_by_c1", 
      "MS917_large_Xrn1_dep_30min_estimate_half_life_by_c1", 
      "MS1039_large_eIF4G_dep_30min_estimate_half_life_by_c1", 
      "MS908_large_eIF4G_dep_1hr_rep1_estimate_half_life_by_c1", 
      "MS908_large_eIF4G_dep_1hr_rep2_estimate_half_life_by_c1", 
      "MS908_large_eIF4G_dep_1hr_ave_estimate_half_life_by_c1",
      "MS1039_large_eIF4G_dep_1hr_estimate_half_life_by_c1", 
      "eIF4G_dep_1hr_large_estimate_half_life_by_c1", 
      "MS96_large_TPM_input",
      "MS1039_large_TPM_input", 
      "WT_large_TPM_input", 
      "MS917_large_Xrn1_dep_30min_TPM_input", 
      "MS1039_large_eIF4G_dep_30min_TPM_input", 
      "MS908_large_eIF4G_dep_1hr_rep1_TPM_input", 
      "MS908_large_eIF4G_dep_1hr_rep2_TPM_input", 
      "MS908_large_eIF4G_dep_1hr_ave_TPM_input", 
      "MS1039_large_eIF4G_dep_1hr_TPM_input", 
      "eIF4G_dep_1hr_large_TPM_input")
z = c(rep("WT_small_TPM_input", 10), rep(NA, 10))
corr_type = rep("spearman", 20)
log2_x = c(rep("FALSE", 10), rep("TRUE", 10))
log2_y = c(rep("FALSE", 10), rep("TRUE", 10))
log2_z = c(rep("TRUE", 10), rep("FALSE", 10))

batch_p <- batch_plot(df_combined, x, y, z, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)
wrap_plots(batch_p, ncol = 5) %>%  ggsave("../figures/small_vs_large.png", ., width = 27, height = 17, dpi = 300)
wrap_plots(batch_p, ncol = 5)
```

## Correlation between fold-change of 1hr eIF4G depletion to WT and expression level

```{r}
x = c("eIF4G_dep_1hr_ave_to_WT_small_half_lives", 
      "eIF4G_dep_1hr_ave_to_WT_large_half_lives")
y = c("WT_small_TPM_input", 
      "WT_large_TPM_input")
corr_type = rep("spearman", 2)
log2_x = rep("TRUE", 2)
log2_y = rep("TRUE", 2)
batch_p <- batch_plot(df_combined, x, y, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y)
wrap_plots(batch_p, ncol = 2) %>%  ggsave("../figures/eIF4G_to_WT_fc_vs_WT_input.png", ., width = 7.5, height = 3.75, dpi = 300)
wrap_plots(batch_p, ncol = 2)
```

## Correlation between fold-change of large to small and expression level

```{r}
x = c("Average_WT_estimate_half_life_fc_by_c1")
y = c("WT_small_TPM_input")
corr_type = c("spearman")
log2_x = c("TRUE")
log2_y = c("TRUE")

batch_p <- batch_plot(df_combined, x, y, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y)
wrap_plots(batch_p, ncol = 1) %>%  ggsave("../figures/large_to_small_half_life_fc_vs_WT_small_input.png", ., width = 3.75, height = 3.5, dpi = 300)
wrap_plots(batch_p, ncol = 1)
```

## Correlation between fold-change of 1hr eIF4G depletion to WT and translation efficiency

```{r}
#Abundance (input) vs TE
plot_corr2(df_combined %>% dplyr::select(contains("YM_mean_norm"), TE_WT_RS, TE_WT_PP, WT_small_TPM_input, WT_large_TPM_input)) %>% ggsave("../figures/WT_input_vs_WT_TE_matrix.png", ., width = 7.5, height = 7.5, dpi = 300)

x = c(rep("WT_small_TPM_input", 2), 
      rep("WT_large_TPM_input", 2))
y = c("TE_small_MandP_YM_mean_norm", "TE_small_P_YM_mean_norm", "TE_large_MandP_YM_mean_norm", "TE_large_P_YM_mean_norm")
corr_type = rep("spearman", 4)
log2_x = rep("TRUE", 4)
log2_y = rep("TRUE", 4)
batch_p <- batch_plot(df_combined, x, y, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y)
wrap_plots(batch_p, ncol = 2) %>%  ggsave("../figures/WT_input_vs_WT_TE_YM.png", ., width = 7.5, height = 7.5, dpi = 300)
wrap_plots(batch_p, ncol = 2)

#Fold change vs TE
plot_corr2(df_combined %>% dplyr::select(contains("YM_mean_norm"), TE_WT_RS, TE_WT_PP, contains("eIF4G_dep_1hr_ave_to_WT"))) %>% ggsave("../figures/eIF4G_to_WT_fc_vs_WT_TE_matrix.png", ., width = 7.5, height = 7.5, dpi = 300)

x = c(rep("eIF4G_dep_1hr_ave_to_WT_small_half_lives", 2), rep("eIF4G_dep_1hr_ave_to_WT_large_half_lives", 2))
y = c("TE_small_P_1_YM_mean_norm", "TE_small_P_10_YM_mean_norm", "TE_large_P_1_YM_mean_norm", "TE_large_P_10_YM_mean_norm")
z = rep(c("WT_small_TPM_input", "WT_large_TPM_input"), 2)
z_lim = replicate(4, c(-3, 13), simplify = FALSE)
corr_type = rep("spearman", 4)
log2_x = rep("TRUE", 4)
log2_y = rep("TRUE", 4)
log2_z = rep("TRUE", 4)
batch_p <- batch_plot(df_combined, x, y, z, corr_type = corr_type, z_lim = z_lim, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)
wrap_plots(batch_p, ncol = 2) %>% ggsave("../figures/eIF4G_to_WT_fc_vs_WT_TE_YM.png", ., width = 9.5, height = 7.5, dpi = 300)
wrap_plots(batch_p, ncol = 2)
```

## Correlation between fold-change of large to small and translation efficiency

```{r}
plot_corr2(df_combined %>% dplyr::select(contains("YM_mean_norm"), TE_WT_RS, TE_WT_PP, Average_WT_estimate_half_life_fc_by_c1, Average_eIF4G_dep_1hr_estimate_half_life_fc_by_c1)) %>% ggsave("../figures/large_to_small_half_life_fc_vs_WT_TE_matrix.png", ., width = 8, height = 8, dpi = 300)

x = rep("Average_WT_estimate_half_life_fc_by_c1", 4)
y = c("TE_small_MandP_YM_mean_norm", "TE_small_P_YM_mean_norm", "TE_large_MandP_YM_mean_norm", "TE_large_P_YM_mean_norm")
corr_type = rep("spearman", 4)
log2_x = rep("TRUE", 4)
log2_y = rep("TRUE", 4)

batch_p <- batch_plot(df_combined, x, y, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y)
wrap_plots(batch_p, ncol = 2) %>% ggsave("../figures/large_to_small_half_life_fc_vs_WT_TE_YM.png", ., width = 7.5, height = 7.5, dpi = 300)
wrap_plots(batch_p, ncol = 2)
```

## Correlation between fold-change of large to small and fold-change of translation efficiency

```{r}
x = rep("Average_WT_estimate_half_life_fc_by_c1", 2)
y = c("TE_MandP_YM_fc", "TE_P_YM_fc")
corr_type = rep("spearman", 2)
log2_x = rep("TRUE", 2)
log2_y = rep("TRUE", 2)

batch_p <- batch_plot(df_combined, x, y, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y)
wrap_plots(batch_p, ncol = 2) %>% ggsave("../figures/large_to_small_half_life_fc_vs_WT_TE_YM_fc.png", ., width = 7.5, height = 3.75, dpi = 300)
wrap_plots(batch_p, ncol = 2)
```

## Additive or combinatory effect of input and TE on half-life fold-change

```{r}
df_model <- df_combined %>% dplyr::select(gene_id, eIF4G_dep_1hr_ave_to_WT_small_half_lives, eIF4G_dep_1hr_ave_to_WT_large_half_lives, Average_WT_estimate_half_life_fc_by_c1, WT_small_TPM_input, WT_large_TPM_input, TE_WT_PP, TE_WT_RS) %>% drop_na() #make sure all model from the same dataset for comparison

model_input <- lm(Average_WT_estimate_half_life_fc_by_c1 ~ WT_small_TPM_input, data = df_model)
model_TE <- lm(Average_WT_estimate_half_life_fc_by_c1 ~ TE_WT_PP, data = df_model)
model_additive <- lm(Average_WT_estimate_half_life_fc_by_c1 ~ WT_small_TPM_input + TE_WT_PP, data = df_model)
model_interaction <- lm(Average_WT_estimate_half_life_fc_by_c1 ~ WT_small_TPM_input * TE_WT_PP, data = df_model)

#paste("just input adj.r.squred:",summary(model_input)$adj.r.squared)
#paste("just TE adj.r.squred:",summary(model_TE)$adj.r.squared)
#paste("input + TE adj.r.squred:",summary(model_additive)$adj.r.squared)
#paste("input * TE adj.r.squred:",summary(model_interaction)$adj.r.squared)

AIC(model_input, model_TE, model_additive, model_interaction) %>% ungroup() %>% dplyr::select(AIC)
```

## Correlation between WT half-life and translation efficiency
```{r}
plot_corr2(df_combined %>% dplyr::select(contains("YM_mean_norm"), TE_WT_RS, TE_WT_PP, WT_small_estimate_half_life_by_c1, WT_large_estimate_half_life_by_c1, eIF4G_dep_1hr_small_estimate_half_life_by_c1, eIF4G_dep_1hr_large_estimate_half_life_by_c1)) %>% ggsave("../figures/large_and_small_half_life_vs_WT_TE_matrix.png", ., width = 8.5, height = 8.5, dpi = 300)

x = c(rep("WT_small_estimate_half_life_by_c1", 2),
      rep("WT_large_estimate_half_life_by_c1", 2),
      rep("eIF4G_dep_1hr_small_estimate_half_life_by_c1", 2),
      rep("eIF4G_dep_1hr_large_estimate_half_life_by_c1", 2))
y = rep(c("TE_WT_PP", "TE_WT_RS"), 4)
corr_type = rep("spearman", 8)
log2_x = rep("TRUE", 8)
log2_y = rep("TRUE", 8)

batch_p <- batch_plot(df_combined, x, y, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y)
wrap_plots(batch_p, ncol = 4) %>% ggsave("../figures/large_and_small_half_life_vs_WT_TE.png", ., width = 14, height = 7, dpi = 300)
wrap_plots(batch_p, ncol = 4)
```

## Correlation between fold-change of WT large to small and fold-change of 1hr eIF4G depletion to WT

```{r}
x = rep("Average_WT_estimate_half_life_fc_by_c1", 2)
y = c("eIF4G_dep_1hr_ave_to_WT_small_half_lives", "eIF4G_dep_1hr_ave_to_WT_large_half_lives" )
z = c("WT_small_TPM_input", "WT_large_TPM_input")
corr_type = rep("spearman", 2)
log2_x = rep("TRUE", 2)
log2_y = rep("TRUE", 2)
log2_z = rep("TRUE", 2)

batch_p <- batch_plot(df_combined, x, y, z, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)
wrap_plots(batch_p, ncol = 2) %>% ggsave("../figures/eIF4G_to_WT_fc_vs_large_to_small_half_life_fc.png", ., width = 9.25, height = 3.75, dpi = 300)
wrap_plots(batch_p, ncol = 2)
```

```{r}
x = rep(c("WT_small_estimate_half_life_by_c1", 
      "WT_large_estimate_half_life_by_c1",
      "eIF4G_dep_1hr_small_estimate_half_life_by_c1",
      "eIF4G_dep_1hr_large_estimate_half_life_by_c1"), 2)
y = c("MS69_small_mrl100_read_fraction",
      "MS69_large_mrl100_read_fraction",
      "MS798_small_eIF4F_dep_1hr_mrl100_read_fraction",
      "MS798_large_eIF4F_dep_1hr_mrl100_read_fraction",
      "MS69_small_mrl100_mean_tail_length",
      "MS69_large_mrl100_mean_tail_length",
      "MS798_small_eIF4F_dep_1hr_mrl100_mean_tail_length",
      "MS798_large_eIF4F_dep_1hr_mrl100_mean_tail_length")
corr_type = rep("pearson", 8)
log2_x = rep("TRUE", 8)
log2_y = rep("TRUE", 8)

batch_p <- batch_plot(df_combined, x, y, corr_type = corr_type, log2_x = log2_x, log2_y = log2_y)
wrap_plots(batch_p, ncol = 4) %>% ggsave("../figures/half_life_vs_gene_fraction&tail_length.png", ., width = 16, height = 8, dpi = 300)
wrap_plots(batch_p, ncol = 4)
```

#temporary
```{r}
x = c(rep("eIF4G_dep_1hr_ave_to_WT_small_half_lives", 1), rep("eIF4G_dep_1hr_ave_to_WT_large_half_lives", 1))
y = c("TE_P_small_YM", "TE_P_large_YM")
z = rep(c("WT_small_TPM_input", "WT_large_TPM_input"), 1)
z_lim = replicate(2, c(-3, 13), simplify = FALSE)
corr_type = rep("spearman", 2)
log2_x = rep("TRUE", 2)
log2_y = rep("TRUE", 2)
log2_z = rep("TRUE", 2)
batch_p0 <- batch_plot(df_combined %>% filter(TPMsc_input_small_YM > 0, TPMsc_input_large_YM > 0),
                      x, y, z, corr_type = corr_type, z_lim = z_lim, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)
batch_p10_s <- batch_plot(df_combined %>% filter(TPMsc_input_small_YM > 10),
                      c("eIF4G_dep_1hr_ave_to_WT_small_half_lives"),
                      c("TE_P_small_YM"), c("WT_small_TPM_input"), corr_type = rep("spearman", 1), z_lim = replicate(1, c(-3, 13), simplify = FALSE), log2_x = rep("TRUE", 1), log2_y = rep("TRUE", 1), log2_z = rep("TRUE", 1))
batch_p10_b <- batch_plot(df_combined %>% filter(TPMsc_input_large_YM > 10),
                      c("eIF4G_dep_1hr_ave_to_WT_large_half_lives"),
                      c("TE_P_large_YM"), c("WT_large_TPM_input"), corr_type = rep("spearman", 1), z_lim = replicate(1, c(-3, 13), simplify = FALSE), log2_x = rep("TRUE", 1), log2_y = rep("TRUE", 1), log2_z = rep("TRUE", 1))
batch_p <- c(batch_p0, batch_p10_s, batch_p10_b)
wrap_plots(batch_p, ncol = 2) %>% ggsave("x.png", ., width = 9.5, height = 7.5, dpi = 300)


x = c(rep("eIF4G_dep_1hr_ave_to_WT_small_half_lives", 1), rep("eIF4G_dep_1hr_ave_to_WT_large_half_lives", 1))
y = rep("TE_WT_PP", 2)
z = rep(c("WT_small_TPM_input", "WT_large_TPM_input"), 1)
z_lim = replicate(2, c(-3, 13), simplify = FALSE)
corr_type = rep("spearman", 2)
log2_x = rep("TRUE", 2)
log2_y = rep("TRUE", 2)
log2_z = rep("TRUE", 2)
batch_p <- batch_plot(df_combined %>% filter(TE_WT_PP > 0.45),
                      x, y, z, corr_type = corr_type, z_lim = z_lim, log2_x = log2_x, log2_y = log2_y, log2_z = log2_z)
wrap_plots(batch_p, ncol = 1) %>% ggsave("x.png", ., width = 4.75, height = 7.5, dpi = 300)
```

