---
title: "Functional Category Annotation Across GO Enrichment Plots (PC1/PC2)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this file's location
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)
library(patchwork)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Reconstructed from an earlier session after the categorised version of
this script was lost. Adds functional category annotation (e.g.
mitochondrial, amino acid biosynthesis) to GO enrichment plots for
biological context at a glance.

Reads the four saved output tables from 01_go_clusterprofiler.Rmd,
applies a keyword classifier to each GO term's Description text, and
re-plots with category as colour. Keyword matching on free text is
imperfect - category assignments are printed for manual review below.

```{r shared-category-function}
# GO term -> broad category classifier
categorise_go_term <- function(description) {
  d <- str_to_lower(description)
  case_when(
    str_detect(d, "ribosom|translation|rrna|preribosome|rna processing|spliceosom|ribonucleoprotein|nucleol") ~ "Ribosome biogenesis / translation",
    str_detect(d, "mitochondri|respirat|electron transport|oxidative phosphorylation|tricarboxylic|citrate cycle|cytochrome|proton.transport") ~ "Mitochondrial / respiration",
    str_detect(d, "amino acid|sulfur|methionine|arginine|histidine|lysine biosynthe") ~ "Amino acid / sulfur metabolism",
    str_detect(d, "carbohydrate|glucose|glycolysis|gluconeogenesis|hexose|lactate") ~ "Carbohydrate / glucose metabolism",
    str_detect(d, "cell cycle|budding|cytokinesis|dna replication|chromosome segregation") ~ "Cell cycle",
    str_detect(d, "stress|starvation|heat shock|osmotic|oxidative stress|autophagy") ~ "Stress response",
    str_detect(d, "chromatin|transcription elongation|transcription initiation|histone|epigenetic") ~ "Chromatin / transcription regulation",
    str_detect(d, "vacuol|lysosom|proteolysis|protein catabolic") ~ "Vacuolar / protein degradation",
    str_detect(d, "lipid|fatty acid|steroid|ergosterol") ~ "Lipid / sterol metabolism",
    str_detect(d, "carboxylic acid|organic acid|oxoacid|precursor metabolite|energy derivation") ~ "General / central metabolism",
    TRUE ~ "Other / unclassified"
  )
}

category_colours <- c(
  "Ribosome biogenesis / translation"      = project_colours[["blue"]],
  "Mitochondrial / respiration"            = project_colours[["red"]],
  "Amino acid / sulfur metabolism"         = project_colours[["purple"]],
  "Carbohydrate / glucose metabolism"      = "#e8a33d",
  "Cell cycle"                             = "#3d9e5a",
  "Stress response"                        = "#c94f7c",
  "Chromatin / transcription regulation"   = "#5a5a9e",
  "Vacuolar / protein degradation"         = "#8c6d31",
  "Lipid / sterol metabolism"              = "#31a1a3",
  "General / central metabolism"           = "#b0a19a",
  "Other / unclassified"                   = "grey50"
)
```

```{r load-saved-tables}
go_pc1_top    <- read_tsv(file.path(output_dir, "go_clusterprofiler_pc1_top10pct.txt"), show_col_types = FALSE)
go_pc1_bottom <- read_tsv(file.path(output_dir, "go_clusterprofiler_pc1_bottom10pct.txt"), show_col_types = FALSE)
go_pc2_top    <- read_tsv(file.path(output_dir, "go_clusterprofiler_pc2_top10pct.txt"), show_col_types = FALSE)
go_pc2_bottom <- read_tsv(file.path(output_dir, "go_clusterprofiler_pc2_bottom10pct.txt"), show_col_types = FALSE)

cat("Loaded saved GO enrichment tables:",
    nrow(go_pc1_top), "/", nrow(go_pc1_bottom), "/",
    nrow(go_pc2_top), "/", nrow(go_pc2_bottom), "rows (PC1 top/bottom, PC2 top/bottom)\n")
```

```{r plot-function}
# Single categorised panel: top 20 terms by adjusted p-value, coloured
# by functional category, sized by gene count. Each panel keeps its
# own self-contained legend (patchwork's guides = "collect" produced
# broken/clipped legends here).
plot_categorised_panel <- function(df, panel_label, direction_label, shared_categories, size_limits) {
  df_top20 <- df %>%
    arrange(p.adjust) %>%
    slice_head(n = 20) %>%
    mutate(
      category           = categorise_go_term(Description),
      category           = factor(category, levels = shared_categories),
      Description_short  = truncate_wrap_description(Description),
      Description_short  = fct_reorder(Description_short, -log10(p.adjust)),
      neg_log10_padj     = -log10(p.adjust)
    )

  cat("\nCategory assignment for panel", panel_label, "-", direction_label, "(REVIEW):\n")
  print(df_top20 %>% select(Description, category), n = Inf)

  ggplot(df_top20, aes(x = neg_log10_padj, y = Description_short,
                       size = Count, colour = category)) +
    geom_point() +
    scale_colour_manual(values = category_colours, name = "Functional category", drop = FALSE) +
    scale_size_continuous(limits = size_limits, range = c(1, 8), name = "Count") +
    labs(
      tag = panel_label,
      x   = expression(-log[10]("adjusted p-value")),
      y   = direction_label
    ) +
    guides(colour = guide_legend(ncol = 1, order = 1),
           size   = guide_legend(ncol = 1, order = 2)) +
    theme_bw(base_size = 32) +
    theme(
      axis.text.y     = element_text(size = 28, colour = "black"),
      axis.text.x     = element_text(size = 26, colour = "black"),
      axis.title      = element_text(size = 30, colour = "black"),
      axis.title.y    = element_text(size = 27, face = "italic", color = "grey40"),
      plot.tag        = element_text(face = "bold", size = 36),
      legend.text     = element_text(size = 26),
      legend.title    = element_text(size = 28),
      legend.position = "right"
    )
}

make_combined_categorised_figure <- function(df_top, df_bottom, label_top, label_bottom) {
  shared_size_limits <- range(c(df_top$Count, df_bottom$Count), na.rm = TRUE)

  # Union of categories present in either panel's top 20, in
  # category_colours order, so colour mapping stays consistent.
  top20_categories <- function(df) {
    df %>% arrange(p.adjust) %>% slice_head(n = 20) %>%
      pull(Description) %>% categorise_go_term() %>% unique()
  }
  present_categories <- union(top20_categories(df_top), top20_categories(df_bottom))
  shared_categories  <- names(category_colours)[names(category_colours) %in% present_categories]

  p_top    <- plot_categorised_panel(df_top,    "A", label_top,    shared_categories, shared_size_limits)
  p_bottom <- plot_categorised_panel(df_bottom, "B", label_bottom, shared_categories, shared_size_limits)

  p_top / p_bottom
}
```

```{r figure-9-pc1-categorised, fig.width=28, fig.height=32}
fig9_categorised <- make_combined_categorised_figure(
  go_pc1_top, go_pc1_bottom,
  label_top    = "Top 10% PC1 (n = 462 genes) — active TOR/PKA signalling state",
  label_bottom = "Bottom 10% PC1 (n = 462 genes) — TOR/PKA-inhibited state"
)

print(fig9_categorised)
```

```{r figure-10-pc2-categorised, fig.width=28, fig.height=32}
fig10_categorised <- make_combined_categorised_figure(
  go_pc2_top, go_pc2_bottom,
  label_top    = "Top 10% PC2 (n = 462 genes) — upregulated in large cells",
  label_bottom = "Bottom 10% PC2 (n = 462 genes) — upregulated in small cells"
)

print(fig10_categorised)
```

```{r save-outputs}
ggsave(file.path(figures_dir, "figure9_go_clusterprofiler_pc1_categorised.png"),
       plot = fig9_categorised, width = 28, height = 32, dpi = 300)
ggsave(file.path(figures_dir, "figure10_go_clusterprofiler_pc2_categorised.png"),
       plot = fig10_categorised, width = 28, height = 32, dpi = 300)

cat("Done - categorised figures saved to figures/\n")
```
