---
title: "GO Enrichment by Expression-Volume Slope, and TF-GO Overlap Corroboration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this file's location
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)
library(readxl)
library(clusterProfiler)
library(org.Sc.sgd.db)
library(GO.db)
library(AnnotationDbi)
library(patchwork)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Part 1: GO enrichment using expression-volume slope (GSEA)

Independent of the PC1/PC2-based GO enrichment (scripts 01-04), this uses
the per-gene Spearman rho (expression vs cell volume) as the GSEA ranking
metric instead of a PCA loading, as a further check on the PC2 GSEA
results (script 03).

```{r load-slope-data}
gene_volume_reg <- read_tsv(
  file.path(output_dir, "gene_expression_vs_cell_volume.txt"),
  show_col_types = FALSE
)

cat("Genes with expression-volume regression results:",
    nrow(gene_volume_reg), "\n")
```

```{r build-ranked-list-slope}
# Ranked list for GSEA: spearman_rho used as the ranking statistic.
# Genes with NA rho are dropped.
ranked_slope <- gene_volume_reg %>%
  filter(!is.na(spearman_rho)) %>%
  arrange(desc(spearman_rho)) %>%
  dplyr::select(gene_id, spearman_rho) %>%
  deframe()

cat("Ranked list for GSEA:", length(ranked_slope), "genes, range [",
    round(min(ranked_slope), 3), ",", round(max(ranked_slope), 3), "]\n")
```

```{r run-gsea-slope}
# gseGO on the slope/rho-ranked list, same parameters as script 03
set.seed(42)
gsea_slope <- gseGO(
  geneList      = ranked_slope,
  OrgDb         = org.Sc.sgd.db,
  keyType       = "ORF",
  ont           = "ALL",
  minGSSize     = 10,
  maxGSSize     = 500,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE,
  nPermSimple   = 1000
)

cat("GSEA (slope-ranked) - enriched terms:",
    nrow(as.data.frame(gsea_slope)), "\n")
```

```{r plot-gsea-slope, fig.width=10, fig.height=18}
# Same two-panel positive/negative NES layout as script 03
plot_gsea_panel <- function(gsea_result, sign, panel_label, direction_label,
                            n_top = 20) {
  df <- as.data.frame(gsea_result)

  df_sign <- df %>%
    filter(if (sign == "activated") NES > 0 else NES < 0) %>%
    arrange(p.adjust) %>%
    slice_head(n = n_top) %>%
    mutate(
      Description_short = truncate_wrap_description(Description),
      Description_short = fct_reorder(Description_short, abs(NES)),
      ONTOLOGY = factor(ONTOLOGY, levels = c("BP", "MF", "CC"))
    )

  if (nrow(df_sign) == 0) return(NULL)

  ggplot(df_sign, aes(x = NES, y = Description_short,
                      size = setSize, color = ONTOLOGY)) +
    geom_point(alpha = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", color = project_colours[["grey"]]) +
    scale_color_manual(values = c(
      "BP" = project_colours[["blue"]],
      "MF" = project_colours[["red"]],
      "CC" = project_colours[["purple"]]
    )) +
    labs(
      tag   = panel_label,
      x     = "Normalised Enrichment Score (NES)",
      y     = direction_label,
      size  = "Gene set size",
      color = "GO namespace"
    ) +
    theme_bw() +
    theme(
      axis.text.y  = element_text(size = 8),
      axis.title.y = element_text(size = 8, face = "italic", color = "grey40"),
      plot.tag     = element_text(face = "bold", size = 12)
    )
}

p_pos <- plot_gsea_panel(gsea_slope, "activated", "A",
                         "Positive NES = expression increases with cell volume")
p_neg <- plot_gsea_panel(gsea_slope, "suppressed", "B",
                         "Negative NES = expression decreases with cell volume")

if (!is.null(p_pos) & !is.null(p_neg)) {
  fig_gsea_slope <- p_pos / p_neg +
    plot_annotation(
      title = paste0(
        "GO enrichment of expression-volume slope (GSEA)\n",
        "Ranked by per-gene Spearman rho (log2 TPMsc vs mean cell volume) | ",
        "independent of PCA"
      ),
      theme = theme(plot.title = element_text(face = "bold", size = 11, hjust = 0))
    )
  print(fig_gsea_slope)
}
```

```{r compare-slope-vs-pc2-gsea}
# Compare top terms between slope-ranked GSEA and PC2 GSEA (script 03)
go_gsea_pc2_path <- file.path(output_dir, "go_gsea_pc2.txt")

if (file.exists(go_gsea_pc2_path)) {
  gsea_pc2_saved <- read_tsv(go_gsea_pc2_path, show_col_types = FALSE)

  top_slope_pos <- as.data.frame(gsea_slope) %>%
    filter(NES > 0) %>% arrange(p.adjust) %>% slice_head(n = 15) %>% pull(Description)
  top_pc2_pos <- gsea_pc2_saved %>%
    filter(NES > 0) %>% arrange(p.adjust) %>% slice_head(n = 15) %>% pull(Description)

  overlap_pos <- intersect(top_slope_pos, top_pc2_pos)

  cat("Top 15 positive-NES terms overlapping between slope-GSEA and",
      "PC2-GSEA:", length(overlap_pos), "of 15\n")
  cat("Overlapping terms:\n")
  print(overlap_pos)
} else {
  cat("go_gsea_pc2.txt not found in output_tables/ - run script 03 first",
      "to enable this comparison.\n")
}
```

```{r save-part1-outputs}
write_tsv(as.data.frame(gsea_slope),
          file.path(output_dir, "go_gsea_expression_volume_slope.txt"))

if (!is.null(p_pos) & !is.null(p_neg)) {
  ggsave(file.path(figures_dir, "figure_gsea_expression_volume_slope.png"),
         plot = fig_gsea_slope, width = 10, height = 18, dpi = 300)
}

cat("Part 1 done.\n")
```

# Part 2: Do TF and GO enrichment results corroborate each other?

Tests whether genes in top GO terms from the PC1/PC2 analysis overlap
with direct targets of the top TFs from the same analysis, using a
hypergeometric test. GO terms are matched by exact GO ID.

```{r rebuild-go-genesets}
# Rebuild GO gene sets (BP terms), keyed by GO ID rather than label.
#
# LIMITATION: AnnotationDbi::select() returns only direct gene-to-GO
# annotations, no hierarchy propagation. This understates broad
# "umbrella" terms like "cellular respiration" (only 14 genes) - that
# term is dropped from the Part 2 overlap test below. Ribosome
# biogenesis, cytoplasmic translation, and TCA cycle set sizes were
# checked and look reasonable.
build_go_genesets_by_id <- function(ontology) {
  go_map <- AnnotationDbi::select(
    org.Sc.sgd.db,
    keys    = keys(org.Sc.sgd.db, keytype = "ORF"),
    columns = c("GO", "ONTOLOGY"),
    keytype = "ORF"
  ) %>%
    filter(ONTOLOGY == ontology, !is.na(GO)) %>%
    dplyr::select(ORF, GO)

  go_map %>%
    group_by(GO) %>%
    summarise(genes = list(unique(ORF)), .groups = "drop") %>%
    deframe()
}

go_genesets_bp_by_id <- build_go_genesets_by_id("BP")
cat("GO:BP gene sets built:", length(go_genesets_bp_by_id), "\n")

# Lookup table of GO ID -> human-readable term name, for labelling
# results and for the manual verification step below.
go_id_to_term <- AnnotationDbi::select(
  GO.db,
  keys    = names(go_genesets_bp_by_id),
  columns = "TERM",
  keytype = "GOID"
) %>%
  dplyr::rename(GO = GOID, term_name = TERM)
```

```{r select-top-go-terms}
# Target GO:BP terms: ribosome biogenesis, cytoplasmic translation
# (PC1 axis), tricarboxylic acid cycle (PC2 axis). "Cellular
# respiration" tested and dropped (see note above).
target_go_ids <- c(
  "ribosome biogenesis"       = "GO:0042254",
  "cytoplasmic translation"   = "GO:0002181",
  "tricarboxylic acid cycle"  = "GO:0006099"
)

matched_go_check <- tibble(
  go_label_intended = names(target_go_ids),
  go_id             = unname(target_go_ids)
) %>%
  left_join(go_id_to_term, by = c("go_id" = "GO")) %>%
  mutate(
    gene_set_size = map_int(go_id, ~length(go_genesets_bp_by_id[[.x]]) %||% NA_integer_),
    id_found_in_genesets = go_id %in% names(go_genesets_bp_by_id)
  )

cat("GO term verification (check term_name matches go_label_intended,",
    "and gene_set_size looks reasonable):\n")
print(matched_go_check, n = Inf)

# stop early if any target GO ID wasn't found or looks implausibly small
if (any(!matched_go_check$id_found_in_genesets)) {
  stop("One or more target GO IDs not found in go_genesets_bp_by_id - ",
       "check target_go_ids against current GO.db annotations before proceeding.")
}
if (any(matched_go_check$gene_set_size < 15, na.rm = TRUE)) {
  warning("At least one target GO term has a gene set size < 15 - ",
          "this flagged the cellular respiration issue previously; ",
          "verify this term's flat-annotation size is genuinely correct ",
          "(not a hierarchy-propagation artefact) before trusting its ",
          "overlap test result.")
}
```

```{r rebuild-tf-genesets-table-e}
# Rebuild Table E direct-target gene sets for CBF1, IFH1, SFP1
# (PC1) and ERT1, RTG1, RTG3 (PC2).
supp_path <- file.path(tables_dir, "Supp Table 3 updated.xlsx")
table_e   <- read_excel(supp_path, sheet = "Table-S3e")

tf_cols_e    <- colnames(table_e)[-1]
table_e_long <- table_e %>%
  pivot_longer(-gene_id, names_to = "TF", values_to = "log2FC") %>%
  filter(!is.na(log2FC))

genesets_e_activated <- table_e_long %>%
  filter(log2FC < 0) %>%
  group_by(TF) %>%
  summarise(genes = list(unique(gene_id)), .groups = "drop") %>%
  filter(map_int(genes, length) >= 5) %>%
  deframe() %>%
  set_names(paste0(names(.), "_direct_activated"))

genesets_e_repressed <- table_e_long %>%
  filter(log2FC > 0) %>%
  group_by(TF) %>%
  summarise(genes = list(unique(gene_id)), .groups = "drop") %>%
  filter(map_int(genes, length) >= 5) %>%
  deframe() %>%
  set_names(paste0(names(.), "_direct_repressed"))

genesets_e <- c(genesets_e_activated, genesets_e_repressed)

target_tf_sets <- c("CBF1_direct_activated", "IFH1_direct_activated",
                    "SFP1_direct_activated", "ERT1_direct_repressed",
                    "RTG1_direct_activated", "RTG3_direct_activated")

available_tf_sets <- intersect(target_tf_sets, names(genesets_e))
cat("Target TF gene sets available:", length(available_tf_sets),
    "of", length(target_tf_sets), "\n")
print(available_tf_sets)

if (length(available_tf_sets) < length(target_tf_sets)) {
  cat("\nMissing:", setdiff(target_tf_sets, available_tf_sets), "\n")
}
```

```{r hypergeometric-overlap-test}
# Hypergeometric test for overlap between GO term genes and TF targets,
# background = PCA gene universe.
# P(overlap >= k) = phyper(k - 1, a, N - a, b, lower.tail = FALSE)
pca_genes <- read_tsv(file.path(tables_dir, "PCA_genes_AC.txt"),
                      show_col_types = FALSE)
universe <- pca_genes$gene_id
n_universe <- length(universe)

run_overlap_test <- function(go_label, go_genes, tf_name, tf_genes, universe) {
  go_genes_u <- intersect(go_genes, universe)
  tf_genes_u <- intersect(tf_genes, universe)
  overlap_genes <- intersect(go_genes_u, tf_genes_u)
  k <- length(overlap_genes)
  a <- length(go_genes_u)
  b <- length(tf_genes_u)
  n <- length(universe)

  p_value <- phyper(k - 1, a, n - a, b, lower.tail = FALSE)
  expected <- (a * b) / n

  tibble(
    go_term          = go_label,
    tf_set           = tf_name,
    go_set_size      = a,
    tf_set_size      = b,
    overlap          = k,
    overlap_genes    = list(overlap_genes),
    expected_overlap = round(expected, 2),
    fold_enrichment  = if_else(expected > 0, round(k / expected, 2), NA_real_),
    p_value          = p_value
  )
}

overlap_results <- expand_grid(
  go_label = names(target_go_ids),
  tf_name  = available_tf_sets
) %>%
  pmap_dfr(function(go_label, tf_name) {
    go_id <- target_go_ids[[go_label]]
    run_overlap_test(
      go_label, go_genesets_bp_by_id[[go_id]],
      tf_name, genesets_e[[tf_name]],
      universe
    )
  })

overlap_results <- overlap_results %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  arrange(p_adjust)

cat("GO-TF overlap tests:", nrow(overlap_results), "(expected:",
    length(target_go_ids) * length(available_tf_sets), ")\n")
cat("Significant overlaps (FDR < 0.05):",
    sum(overlap_results$p_adjust < 0.05), "\n\n")

print(overlap_results %>% dplyr::select(-overlap_genes), n = Inf)
```

```{r plot-overlap-heatmap, fig.width=9, fig.height=4.5}
# Heatmap of fold enrichment across GO terms x TFs, significance marked
overlap_results_plot <- overlap_results %>%
  mutate(sig_label = if_else(p_adjust < 0.05, "*", ""))

p_overlap <- ggplot(overlap_results_plot,
                    aes(x = tf_set, y = go_term, fill = fold_enrichment)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(overlap, sig_label)), size = 4) +
  scale_fill_gradient(low = "white", high = project_colours[["red"]],
                      na.value = "grey90") +
  labs(
    x = "TF direct-target gene set",
    y = "GO:BP term",
    fill = "Fold\nenrichment"
  ) +
  theme_bw() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1)
  )

print(p_overlap)
```

```{r annotate-overlap-genes-with-names}
# Add gene names alongside systematic IDs for the overlap gene lists
gene_id_to_name <- pca_genes %>%
  dplyr::select(gene_id, any_of("gene_name")) %>%
  distinct()

if (!"gene_name" %in% colnames(gene_id_to_name)) {
  cat("gene_name column not found in PCA_genes_AC.txt - falling back to",
      "df_sc_final.txt for the ID-to-name lookup.\n")
  gene_id_to_name <- read_tsv(file.path(tables_dir, "df_sc_final.txt"),
                              show_col_types = FALSE) %>%
    dplyr::select(gene_id, gene_name) %>%
    distinct()
}

sig_overlaps <- overlap_results %>% filter(p_adjust < 0.05)

if (nrow(sig_overlaps) > 0) {
  cat("Significant GO-TF overlaps and their gene lists (with names):\n\n")
  for (i in seq_len(nrow(sig_overlaps))) {
    genes_named <- tibble(gene_id = sig_overlaps$overlap_genes[[i]]) %>%
      left_join(gene_id_to_name, by = "gene_id")

    cat(sig_overlaps$go_term[i], "x", sig_overlaps$tf_set[i],
        "(", sig_overlaps$overlap[i], "genes ):\n")
    print(genes_named, n = Inf)
    cat("\n")
  }
} else {
  cat("No significant overlaps at FDR < 0.05.\n")
}
```

---
title: "GO Enrichment by Expression-Volume Slope, and TF-GO Overlap Corroboration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this file's location
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)
library(readxl)
library(clusterProfiler)
library(org.Sc.sgd.db)
library(GO.db)
library(AnnotationDbi)
library(patchwork)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Part 1: GO enrichment using expression-volume slope (GSEA)

Independent of the PC1/PC2-based GO enrichment (scripts 01-04), this uses
the per-gene Spearman rho (expression vs cell volume) as the GSEA ranking
metric instead of a PCA loading, as a further check on the PC2 GSEA
results (script 03).

```{r load-slope-data}
gene_volume_reg <- read_tsv(
  file.path(output_dir, "gene_expression_vs_cell_volume.txt"),
  show_col_types = FALSE
)

cat("Genes with expression-volume regression results:",
    nrow(gene_volume_reg), "\n")
```

```{r build-ranked-list-slope}
# Ranked list for GSEA: spearman_rho used as the ranking statistic.
# Genes with NA rho are dropped.
ranked_slope <- gene_volume_reg %>%
  dplyr::filter(!is.na(spearman_rho)) %>%
  arrange(desc(spearman_rho)) %>%
  dplyr::select(gene_id, spearman_rho) %>%
  deframe()

cat("Ranked list for GSEA:", length(ranked_slope), "genes, range [",
    round(min(ranked_slope), 3), ",", round(max(ranked_slope), 3), "]\n")
```

```{r run-gsea-slope}
# gseGO on the slope/rho-ranked list, same parameters as script 03
set.seed(42)
gsea_slope <- gseGO(
  geneList      = ranked_slope,
  OrgDb         = org.Sc.sgd.db,
  keyType       = "ORF",
  ont           = "ALL",
  minGSSize     = 10,
  maxGSSize     = 500,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE,
  nPermSimple   = 1000
)

cat("GSEA (slope-ranked) - enriched terms:",
    nrow(as.data.frame(gsea_slope)), "\n")
```

```{r plot-gsea-slope, fig.width=10, fig.height=18}
# Same two-panel positive/negative NES layout as script 03
plot_gsea_panel <- function(gsea_result, sign, panel_label, direction_label,
                            n_top = 20) {
  df <- as.data.frame(gsea_result)

  df_sign <- df %>%
    dplyr::filter(if (sign == "activated") NES > 0 else NES < 0) %>%
    arrange(p.adjust) %>%
    slice_head(n = n_top) %>%
    mutate(
      Description_short = truncate_wrap_description(Description),
      Description_short = fct_reorder(Description_short, abs(NES)),
      ONTOLOGY = factor(ONTOLOGY, levels = c("BP", "MF", "CC"))
    )

  if (nrow(df_sign) == 0) return(NULL)

  ggplot(df_sign, aes(x = NES, y = Description_short,
                      size = setSize, color = ONTOLOGY)) +
    geom_point(alpha = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", color = project_colours[["grey"]]) +
    scale_color_manual(values = c(
      "BP" = project_colours[["blue"]],
      "MF" = project_colours[["red"]],
      "CC" = project_colours[["purple"]]
    )) +
    labs(
      tag   = panel_label,
      x     = "Normalised Enrichment Score (NES)",
      y     = direction_label,
      size  = "Gene set size",
      color = "GO namespace"
    ) +
    theme_bw() +
    theme(
      axis.text.y  = element_text(size = 8),
      axis.title.y = element_text(size = 8, face = "italic", color = "grey40"),
      plot.tag     = element_text(face = "bold", size = 12)
    )
}

p_pos <- plot_gsea_panel(gsea_slope, "activated", "A",
                         "Positive NES = expression increases with cell volume")
p_neg <- plot_gsea_panel(gsea_slope, "suppressed", "B",
                         "Negative NES = expression decreases with cell volume")

if (!is.null(p_pos) & !is.null(p_neg)) {
  fig_gsea_slope <- p_pos / p_neg +
    plot_annotation(
      title = paste0(
        "GO enrichment of expression-volume slope (GSEA)\n",
        "Ranked by per-gene Spearman rho (log2 TPMsc vs mean cell volume) | ",
        "independent of PCA"
      ),
      theme = theme(plot.title = element_text(face = "bold", size = 11, hjust = 0))
    )
  print(fig_gsea_slope)
}
```

```{r compare-slope-vs-pc2-gsea}
# Compare top terms between slope-ranked GSEA and PC2 GSEA (script 03)
go_gsea_pc2_path <- file.path(output_dir, "go_gsea_pc2.txt")

if (file.exists(go_gsea_pc2_path)) {
  gsea_pc2_saved <- read_tsv(go_gsea_pc2_path, show_col_types = FALSE)

  top_slope_pos <- as.data.frame(gsea_slope) %>%
    dplyr::filter(NES > 0) %>% arrange(p.adjust) %>% slice_head(n = 15) %>% pull(Description)
  top_pc2_pos <- gsea_pc2_saved %>%
    dplyr::filter(NES > 0) %>% arrange(p.adjust) %>% slice_head(n = 15) %>% pull(Description)

  overlap_pos <- intersect(top_slope_pos, top_pc2_pos)

  cat("Top 15 positive-NES terms overlapping between slope-GSEA and",
      "PC2-GSEA:", length(overlap_pos), "of 15\n")
  cat("Overlapping terms:\n")
  print(overlap_pos)
} else {
  cat("go_gsea_pc2.txt not found in output_tables/ - run script 03 first",
      "to enable this comparison.\n")
}
```

```{r save-part1-outputs}
write_tsv(as.data.frame(gsea_slope),
          file.path(output_dir, "go_gsea_expression_volume_slope.txt"))

if (!is.null(p_pos) & !is.null(p_neg)) {
  ggsave(file.path(figures_dir, "figure_gsea_expression_volume_slope.png"),
         plot = fig_gsea_slope, width = 10, height = 18, dpi = 300)
}

cat("Part 1 done.\n")
```

# Part 2: Do TF and GO enrichment results corroborate each other?

Tests whether genes in top GO terms from the PC1/PC2 analysis overlap
with direct targets of the top TFs from the same analysis, using a
hypergeometric test. GO terms are matched by exact GO ID.

```{r rebuild-go-genesets}
# Rebuild GO gene sets (BP terms), keyed by GO ID rather than label.
#
# LIMITATION: AnnotationDbi::select() returns only direct gene-to-GO
# annotations, no hierarchy propagation. This understates broad
# "umbrella" terms like "cellular respiration" (only 14 genes) - that
# term is dropped from the Part 2 overlap test below. Ribosome
# biogenesis, cytoplasmic translation, and TCA cycle set sizes were
# checked and look reasonable.
build_go_genesets_by_id <- function(ontology) {
  go_map <- AnnotationDbi::select(
    org.Sc.sgd.db,
    keys    = keys(org.Sc.sgd.db, keytype = "ORF"),
    columns = c("GO", "ONTOLOGY"),
    keytype = "ORF"
  ) %>%
    dplyr::filter(ONTOLOGY == ontology, !is.na(GO)) %>%
    dplyr::select(ORF, GO)

  go_map %>%
    group_by(GO) %>%
    summarise(genes = list(unique(ORF)), .groups = "drop") %>%
    deframe()
}

go_genesets_bp_by_id <- build_go_genesets_by_id("BP")
cat("GO:BP gene sets built:", length(go_genesets_bp_by_id), "\n")

# Lookup table of GO ID -> human-readable term name, for labelling
# results and for the manual verification step below.
go_id_to_term <- AnnotationDbi::select(
  GO.db,
  keys    = names(go_genesets_bp_by_id),
  columns = "TERM",
  keytype = "GOID"
) %>%
  dplyr::rename(GO = GOID, term_name = TERM)
```

```{r select-top-go-terms}
# Target GO:BP terms: ribosome biogenesis, cytoplasmic translation
# (PC1 axis), tricarboxylic acid cycle (PC2 axis). "Cellular
# respiration" tested and dropped (see note above).
target_go_ids <- c(
  "ribosome biogenesis"       = "GO:0042254",
  "cytoplasmic translation"   = "GO:0002181",
  "tricarboxylic acid cycle"  = "GO:0006099"
)

matched_go_check <- tibble(
  go_label_intended = names(target_go_ids),
  go_id             = unname(target_go_ids)
) %>%
  left_join(go_id_to_term, by = c("go_id" = "GO")) %>%
  mutate(
    gene_set_size = map_int(go_id, ~length(go_genesets_bp_by_id[[.x]]) %||% NA_integer_),
    id_found_in_genesets = go_id %in% names(go_genesets_bp_by_id)
  )

cat("GO term verification (check term_name matches go_label_intended,",
    "and gene_set_size looks reasonable):\n")
print(matched_go_check, n = Inf)

# stop early if any target GO ID wasn't found or looks implausibly small
if (any(!matched_go_check$id_found_in_genesets)) {
  stop("One or more target GO IDs not found in go_genesets_bp_by_id - ",
       "check target_go_ids against current GO.db annotations before proceeding.")
}
if (any(matched_go_check$gene_set_size < 15, na.rm = TRUE)) {
  warning("At least one target GO term has a gene set size < 15 - ",
          "this flagged the cellular respiration issue previously; ",
          "verify this term's flat-annotation size is genuinely correct ",
          "(not a hierarchy-propagation artefact) before trusting its ",
          "overlap test result.")
}
```

```{r rebuild-tf-genesets-table-e}
# Rebuild Table E direct-target gene sets for CBF1, IFH1, SFP1
# (PC1) and ERT1, RTG1, RTG3 (PC2).
supp_path <- file.path(tables_dir, "Supp Table 3 updated.xlsx")
table_e   <- read_excel(supp_path, sheet = "Table-S3e")

tf_cols_e    <- colnames(table_e)[-1]
table_e_long <- table_e %>%
  pivot_longer(-gene_id, names_to = "TF", values_to = "log2FC") %>%
  dplyr::filter(!is.na(log2FC))

genesets_e_activated <- table_e_long %>%
  dplyr::filter(log2FC < 0) %>%
  group_by(TF) %>%
  summarise(genes = list(unique(gene_id)), .groups = "drop") %>%
  dplyr::filter(map_int(genes, length) >= 5) %>%
  deframe() %>%
  set_names(paste0(names(.), "_direct_activated"))

genesets_e_repressed <- table_e_long %>%
  dplyr::filter(log2FC > 0) %>%
  group_by(TF) %>%
  summarise(genes = list(unique(gene_id)), .groups = "drop") %>%
  dplyr::filter(map_int(genes, length) >= 5) %>%
  deframe() %>%
  set_names(paste0(names(.), "_direct_repressed"))

genesets_e <- c(genesets_e_activated, genesets_e_repressed)

target_tf_sets <- c("CBF1_direct_activated", "IFH1_direct_activated",
                    "SFP1_direct_activated", "ERT1_direct_repressed",
                    "RTG1_direct_activated", "RTG3_direct_activated")

available_tf_sets <- intersect(target_tf_sets, names(genesets_e))
cat("Target TF gene sets available:", length(available_tf_sets),
    "of", length(target_tf_sets), "\n")
print(available_tf_sets)

if (length(available_tf_sets) < length(target_tf_sets)) {
  cat("\nMissing:", setdiff(target_tf_sets, available_tf_sets), "\n")
}
```

```{r hypergeometric-overlap-test}
# Hypergeometric test for overlap between GO term genes and TF targets,
# background = PCA gene universe.
# P(overlap >= k) = phyper(k - 1, a, N - a, b, lower.tail = FALSE)
pca_genes <- read_tsv(file.path(tables_dir, "PCA_genes_AC.txt"),
                      show_col_types = FALSE)
universe <- pca_genes$gene_id
n_universe <- length(universe)

run_overlap_test <- function(go_label, go_genes, tf_name, tf_genes, universe) {
  go_genes_u <- intersect(go_genes, universe)
  tf_genes_u <- intersect(tf_genes, universe)
  overlap_genes <- intersect(go_genes_u, tf_genes_u)
  k <- length(overlap_genes)
  a <- length(go_genes_u)
  b <- length(tf_genes_u)
  n <- length(universe)

  p_value <- phyper(k - 1, a, n - a, b, lower.tail = FALSE)
  expected <- (a * b) / n

  tibble(
    go_term          = go_label,
    tf_set           = tf_name,
    go_set_size      = a,
    tf_set_size      = b,
    overlap          = k,
    overlap_genes    = list(overlap_genes),
    expected_overlap = round(expected, 2),
    fold_enrichment  = if_else(expected > 0, round(k / expected, 2), NA_real_),
    p_value          = p_value
  )
}

overlap_results <- expand_grid(
  go_label = names(target_go_ids),
  tf_name  = available_tf_sets
) %>%
  pmap_dfr(function(go_label, tf_name) {
    go_id <- target_go_ids[[go_label]]
    run_overlap_test(
      go_label, go_genesets_bp_by_id[[go_id]],
      tf_name, genesets_e[[tf_name]],
      universe
    )
  })

overlap_results <- overlap_results %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  arrange(p_adjust)

cat("GO-TF overlap tests:", nrow(overlap_results), "(expected:",
    length(target_go_ids) * length(available_tf_sets), ")\n")
cat("Significant overlaps (FDR < 0.05):",
    sum(overlap_results$p_adjust < 0.05), "\n\n")

print(overlap_results %>% dplyr::select(-overlap_genes), n = Inf)
```

```{r plot-overlap-heatmap, fig.width=9, fig.height=4.5}
# Heatmap of fold enrichment across GO terms x TFs, significance marked
overlap_results_plot <- overlap_results %>%
  mutate(sig_label = if_else(p_adjust < 0.05, "*", ""))

p_overlap <- ggplot(overlap_results_plot,
                    aes(x = tf_set, y = go_term, fill = fold_enrichment)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(overlap, sig_label)), size = 4) +
  scale_fill_gradient(low = "white", high = project_colours[["red"]],
                      na.value = "grey90") +
  labs(
    x = "TF direct-target gene set",
    y = "GO:BP term",
    fill = "Fold\nenrichment"
  ) +
  theme_bw() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1)
  )

print(p_overlap)
```

```{r annotate-overlap-genes-with-names}
# Add gene names alongside systematic IDs for the overlap gene lists
gene_id_to_name <- pca_genes %>%
  dplyr::select(gene_id, any_of("gene_name")) %>%
  distinct()

if (!"gene_name" %in% colnames(gene_id_to_name)) {
  cat("gene_name column not found in PCA_genes_AC.txt - falling back to",
      "df_sc_final.txt for the ID-to-name lookup.\n")
  gene_id_to_name <- read_tsv(file.path(tables_dir, "df_sc_final.txt"),
                              show_col_types = FALSE) %>%
    dplyr::select(gene_id, gene_name) %>%
    distinct()
}

sig_overlaps <- overlap_results %>% dplyr::filter(p_adjust < 0.05)

if (nrow(sig_overlaps) > 0) {
  cat("Significant GO-TF overlaps and their gene lists (with names):\n\n")
  for (i in seq_len(nrow(sig_overlaps))) {
    genes_named <- tibble(gene_id = sig_overlaps$overlap_genes[[i]]) %>%
      left_join(gene_id_to_name, by = "gene_id")

    cat(sig_overlaps$go_term[i], "x", sig_overlaps$tf_set[i],
        "(", sig_overlaps$overlap[i], "genes ):\n")
    print(genes_named, n = Inf)
    cat("\n")
  }
} else {
  cat("No significant overlaps at FDR < 0.05.\n")
}
```

```{r save-part2-outputs}
write_tsv(overlap_results %>% dplyr::select(-overlap_genes),
          file.path(output_dir, "go_tf_overlap_corroboration.txt"))

# Save overlap gene lists separately (long format, with names)
overlap_genes_long <- overlap_results %>%
  dplyr::select(go_term, tf_set, overlap_genes) %>%
  tidyr::unnest(overlap_genes) %>%
  dplyr::rename(gene_id = overlap_genes) %>%
  left_join(gene_id_to_name, by = "gene_id")

write_tsv(overlap_genes_long,
          file.path(output_dir, "go_tf_overlap_gene_lists.txt"))

# width/height/dpi set explicitly for the saved file (not just the chunk preview)
ggsave(file.path(figures_dir, "figure_go_tf_overlap_heatmap.png"),
       plot = p_overlap, width = 9, height = 4.5, dpi = 300)

cat("Part 2 done - results saved to output_tables/ and figures/\n")
```