---
title: "RNAPII Occupancy Saturation vs TF Direct Targets"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this Rmd's location (Adriana_analysis/scripts/)
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)
library(readxl)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Follow-up to the RNAPII occupancy vs PC1/PC2 analysis (script 11). The
binned-mean trend of log2(occupancy) vs PC1 rises steadily up to roughly
PC1 ~ 7, then turns over and declines for the highest-PC1 genes - the
visual signature of promoter saturation (occupancy failing to keep pace
with transcriptional demand at the very top of the PC1 axis). This script:

1. Defines the plateau/decline region explicitly from the binned-mean data
   (rather than the coarser quartile-based definition used in script 11),
2. Tests whether genes in that region are enriched among the direct
   targets of the TFs already identified as PC1/PC2 hits (CBF1, IFH1,
   SFP1, ERT1, RTG1, RTG3), to check whether saturation is happening
   specifically to the genes those TFs regulate.

```{r load-occupancy-pc}
occ_pc <- read_tsv(file.path(output_dir, "rnapii_occupancy_pc_joined.txt"),
                   show_col_types = FALSE)

cat("Genes with occupancy + PC1/PC2:", nrow(occ_pc), "\n")
```

```{r identify-plateau-region}
# Recreate the same 30-bin binned mean used in script 11's scatter plot,
# to identify exactly where the trend turns over rather than eyeballing
# it from the figure.
n_bins <- 30

pc1_binned <- occ_pc %>%
  mutate(pc1_bin = ntile(PC1, n_bins)) %>%
  group_by(pc1_bin) %>%
  summarise(
    pc1_mean  = mean(PC1),
    occ_mean  = mean(log2_mean_occupancy),
    n_genes   = n(),
    .groups = "drop"
  ) %>%
  arrange(pc1_bin)

cat("Binned mean occupancy across PC1 (bin 30 = highest PC1):\n")
print(pc1_binned, n = Inf)

# Peak bin = where occ_mean is maximised; plateau region defined as all
# bins from the peak onward (inclusive), since occupancy stops rising
# (and starts falling) there despite PC1 continuing to increase.
peak_bin <- pc1_binned$pc1_bin[which.max(pc1_binned$occ_mean)]
plateau_pc1_threshold <- pc1_binned %>%
  filter(pc1_bin == peak_bin) %>%
  pull(pc1_mean)

cat("\nPeak bin:", peak_bin, "of", n_bins,
    "| occupancy peaks at PC1 ~", round(plateau_pc1_threshold, 2), "\n")
cat("Bins", peak_bin, "to", n_bins, "(PC1 >=", round(plateau_pc1_threshold, 2),
    ") define the plateau/decline region.\n")
```

```{r define-saturation-candidates-refined}
# Refined saturation candidate definition: genes at or beyond the peak
# PC1 bin (i.e. in the plateau/decline region), regardless of their
# occupancy rank - unlike script 11's quartile-based definition, this
# uses the actual empirical turnover point rather than an arbitrary
# quartile split, and doesn't require low occupancy rank (the genes of
# interest here are precisely those with high PC1 but occupancy that
# hasn't risen to match - some may still have moderate/high absolute
# occupancy, just not as high as the PC1 trend up to that point implies).
saturation_candidates_refined <- occ_pc %>%
  filter(PC1 >= plateau_pc1_threshold) %>%
  arrange(desc(PC1)) %>%
  select(gene_id, gene_name, PC1, PC2, mean_occupancy,
         log2_mean_occupancy, rank_mean_occupancy)

cat("Genes in the PC1 plateau/decline region (PC1 >=",
    round(plateau_pc1_threshold, 2), "):",
    nrow(saturation_candidates_refined), "\n")
print(saturation_candidates_refined, n = 30)
```

```{r rebuild-tf-genesets}
# Rebuild Table E direct-target gene sets exactly as in script 05,
# restricted to the headline TFs: CBF1, IFH1, SFP1 (PC1-positive) and
# ERT1, RTG1, RTG3 (PC2-positive).
supp_path <- file.path(tables_dir, "Supp Table 3 updated.xlsx")
table_e   <- read_excel(supp_path, sheet = "Table-S3e")

tf_cols_e    <- colnames(table_e)[-1]
table_e_long <- table_e %>%
  pivot_longer(-gene_id, names_to = "TF", values_to = "log2FC") %>%
  filter(!is.na(log2FC))

genesets_e_activated <- table_e_long %>%
  filter(log2FC < 0) %>%
  group_by(TF) %>%
  summarise(genes = list(unique(gene_id)), .groups = "drop") %>%
  filter(map_int(genes, length) >= 5) %>%
  deframe() %>%
  set_names(paste0(names(.), "_direct_activated"))

genesets_e_repressed <- table_e_long %>%
  filter(log2FC > 0) %>%
  group_by(TF) %>%
  summarise(genes = list(unique(gene_id)), .groups = "drop") %>%
  filter(map_int(genes, length) >= 5) %>%
  deframe() %>%
  set_names(paste0(names(.), "_direct_repressed"))

genesets_e <- c(genesets_e_activated, genesets_e_repressed)

target_tf_sets <- c("CBF1_direct_activated", "IFH1_direct_activated",
                    "SFP1_direct_activated", "ERT1_direct_repressed",
                    "RTG1_direct_activated", "RTG3_direct_activated")

available_tf_sets <- intersect(target_tf_sets, names(genesets_e))
cat("Target TF gene sets available:", length(available_tf_sets),
    "of", length(target_tf_sets), "\n")
```

```{r hypergeometric-overlap-test}
# Hypergeometric test for overlap between the saturation candidate genes
# and each TF's direct target genes, restricted to the common gene
# universe (all genes with both occupancy and PC1/PC2 data) so the
# background matches what the saturation candidates were drawn from.
universe <- occ_pc$gene_id
n_universe <- length(universe)

run_overlap_test <- function(candidate_genes, tf_name, tf_genes, universe) {
  cand_u <- intersect(candidate_genes, universe)
  tf_u   <- intersect(tf_genes, universe)
  overlap_genes <- intersect(cand_u, tf_u)
  k <- length(overlap_genes)
  a <- length(cand_u)
  b <- length(tf_u)
  n <- length(universe)

  p_value <- phyper(k - 1, a, n - a, b, lower.tail = FALSE)
  expected <- (a * b) / n

  tibble(
    tf_set           = tf_name,
    candidate_set_size = a,
    tf_set_size      = b,
    overlap          = k,
    overlap_genes    = list(overlap_genes),
    expected_overlap = round(expected, 2),
    fold_enrichment  = if_else(expected > 0, round(k / expected, 2), NA_real_),
    p_value          = p_value
  )
}

saturation_tf_overlap <- map_dfr(available_tf_sets, function(tf_name) {
  run_overlap_test(
    saturation_candidates_refined$gene_id, tf_name,
    genesets_e[[tf_name]], universe
  )
})

saturation_tf_overlap <- saturation_tf_overlap %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  arrange(p_adjust)

cat("Saturation candidates x TF direct-target overlap:\n")
print(saturation_tf_overlap %>% select(-overlap_genes), n = Inf)
```

```{r plot-overlap-bar, fig.width=8, fig.height=5}
# Bar chart of fold enrichment per TF, since this is a 1D comparison
# (one candidate gene set vs 6 TF sets) rather than the 2D grid used for
# the GO-TF overlap in an earlier script.
overlap_plot_df <- saturation_tf_overlap %>%
  mutate(
    sig_label = if_else(p_adjust < 0.05, "*", ""),
    tf_set = fct_reorder(tf_set, fold_enrichment)
  )

p_saturation_tf <- ggplot(overlap_plot_df,
                          aes(x = fold_enrichment, y = tf_set)) +
  geom_col(fill = project_colours[["red"]], alpha = 0.8) +
  geom_text(aes(label = paste0(overlap, sig_label)),
            hjust = -0.2, size = 3.5) +
  labs(
    title = "Overlap between PC1-plateau saturation candidates and TF direct targets",
    subtitle = paste0(
      nrow(saturation_candidates_refined), " candidate genes (PC1 >= ",
      round(plateau_pc1_threshold, 2), ") | * = FDR < 0.05 | bar label = overlap count"
    ),
    x = "Fold enrichment over expected by chance",
    y = "TF direct-target gene set"
  ) +
  theme_bw() +
  xlim(0, max(overlap_plot_df$fold_enrichment, na.rm = TRUE) * 1.3)

print(p_saturation_tf)
```

```{r print-significant-overlap-genes}
# Named gene lists for any significant overlaps, using gene_name already
# present in occ_pc (carried through from the occupancy file's own
# gene_name column).
gene_id_to_name <- occ_pc %>% select(gene_id, gene_name) %>% distinct()

sig_overlaps <- saturation_tf_overlap %>% filter(p_adjust < 0.05)

if (nrow(sig_overlaps) > 0) {
  cat("Significant saturation-candidate x TF overlaps:\n\n")
  for (i in seq_len(nrow(sig_overlaps))) {
    genes_named <- tibble(gene_id = sig_overlaps$overlap_genes[[i]]) %>%
      left_join(gene_id_to_name, by = "gene_id")

    cat(sig_overlaps$tf_set[i], "(", sig_overlaps$overlap[i], "genes ):\n")
    print(genes_named, n = Inf)
    cat("\n")
  }
} else {
  cat("No significant overlaps at FDR < 0.05.\n")
}
```

```{r highlight-plateau-genes-on-scatter, fig.width=8, fig.height=6}
# Re-draw the PC1 occupancy scatter, highlighting saturation candidate
# genes and colouring them by whether they are a direct target of any of
# the 6 headline TFs - visual check of whether the TF-regulated genes
# cluster specifically in the plateau region.
occ_pc_annotated <- occ_pc %>%
  mutate(
    is_saturation_candidate = gene_id %in% saturation_candidates_refined$gene_id,
    is_tf_target = gene_id %in% unlist(genesets_e[available_tf_sets])
  )

pc1_binned_plot <- pc1_binned  # reuse from earlier chunk

p_highlight <- ggplot(occ_pc_annotated, aes(x = PC1, y = log2_mean_occupancy)) +
  geom_point(data = filter(occ_pc_annotated, !is_saturation_candidate),
             color = "grey80", alpha = 0.3, size = 0.6) +
  geom_point(data = filter(occ_pc_annotated, is_saturation_candidate & !is_tf_target),
             color = project_colours[["blue"]], alpha = 0.6, size = 1.2) +
  geom_point(data = filter(occ_pc_annotated, is_saturation_candidate & is_tf_target),
             color = project_colours[["red"]], alpha = 0.8, size = 1.6) +
  geom_line(data = pc1_binned_plot, aes(x = pc1_mean, y = occ_mean),
            color = "black", linewidth = 0.8) +
  geom_vline(xintercept = plateau_pc1_threshold, linetype = "dashed",
             color = project_colours[["grey"]]) +
  labs(
    title = "RNAPII occupancy vs PC1: saturation candidates and TF direct targets",
    subtitle = paste0(
      "Grey = all genes | Blue = plateau-region genes (PC1 >= ",
      round(plateau_pc1_threshold, 2), ") | Red = plateau-region genes that are ",
      "also a direct target of CBF1/IFH1/SFP1/ERT1/RTG1/RTG3"
    ),
    x = "PC1 gene loading",
    y = "log2(mean RNAPII occupancy + 1)"
  ) +
  theme_bw() +
  theme(plot.subtitle = element_text(size = 8))

print(p_highlight)
```

```{r save-outputs}
write_tsv(saturation_candidates_refined,
          file.path(output_dir, "rnapii_saturation_candidates_refined.txt"))
write_tsv(saturation_tf_overlap %>% select(-overlap_genes),
          file.path(output_dir, "rnapii_saturation_tf_overlap.txt"))

overlap_genes_long <- saturation_tf_overlap %>%
  select(tf_set, overlap_genes) %>%
  unnest(overlap_genes) %>%
  rename(gene_id = overlap_genes) %>%
  left_join(gene_id_to_name, by = "gene_id")

write_tsv(overlap_genes_long,
          file.path(output_dir, "rnapii_saturation_tf_overlap_genes.txt"))

ggsave(file.path(figures_dir, "figure_rnapii_saturation_tf_overlap.png"),
       plot = p_saturation_tf, width = 8, height = 5, dpi = 300)
ggsave(file.path(figures_dir, "figure_rnapii_saturation_scatter_highlighted.png"),
       plot = p_highlight, width = 8, height = 6, dpi = 300)

cat("Done - results saved to output_tables/ and figures/\n")
```
