---
title: "RNAPII Occupancy vs PC1/PC2 (Promoter Saturation Analysis)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this Rmd's location (Adriana_analysis/scripts/)
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Rpb1 (RNA Pol II) ChIP occupancy data provided by Matthew (210601_gene
annotation_RPB_occupancy.txt), covering 6 independent experiments (A, B, C,
EFG, H, J - corresponding to different figures of a published paper).
Per Matthew's guidance (Slack, [date]):
- All 6 experiments should be treated as equivalent/interchangeable for this
  purpose - run the main analysis on the average across all 6, then produce
  a grid summarising each individual experiment's correlation separately.
- Correlate occupancy against PC1 and PC2 gene loadings (not cell volume
  directly), so the TOR/PKA-signalling effect (PC1) can be separated from
  the cell-size effect (PC2).
- Correlate both the log2 of occupancy AND the simple rank of occupancy
  against PC1/PC2.

```{r load-occupancy}
occupancy_path <- file.path(
  tables_dir,
  "210601_gene_annotation_RPB_occupancy.txt"
)

occupancy_raw <- read_tsv(occupancy_path, show_col_types = FALSE)

cat("Occupancy data:", nrow(occupancy_raw), "genes x",
    ncol(occupancy_raw), "columns\n")
```

```{r select-occupancy-columns}
# The 6 experiments (letters as used in Matthew's file, corresponding to
# different figures of the published paper). RPKM_mean_<X> = occupancy,
# rank<X> = pre-computed within-experiment percentile rank (0-1).
occupancy_conditions <- c("A", "B", "C", "EFG", "H", "J")

rpkm_cols <- paste0("RPKM_mean_", occupancy_conditions)
rank_cols <- paste0("rank", occupancy_conditions)

occupancy <- occupancy_raw %>%
  select(systematic_name, gene_name, all_of(rpkm_cols), all_of(rank_cols)) %>%
  rename(gene_id = systematic_name)

cat("Missing values per experiment (RPKM):\n")
print(colSums(is.na(occupancy[rpkm_cols])))
```

```{r compute-average-and-transforms}
# Average occupancy across all 6 experiments, treated as replicates per
# Matthew's instruction. na.rm = TRUE so a gene missing in one experiment
# is averaged over the remaining experiments rather than dropped entirely;
# n_experiments_present tracks how many contributed, for filtering low-
# coverage genes below.
occupancy <- occupancy %>%
  rowwise() %>%
  mutate(
    n_experiments_present = sum(!is.na(c_across(all_of(rpkm_cols)))),
    mean_occupancy         = mean(c_across(all_of(rpkm_cols)), na.rm = TRUE)
  ) %>%
  ungroup()

cat("Genes by number of experiments present (of 6):\n")
print(table(occupancy$n_experiments_present))

# Require at least 4 of 6 experiments present for a gene to be included in
# the averaged analysis - arbitrary but conservative threshold; adjust if
# Matthew has a preference. Genes failing this can still appear in the
# per-experiment grid below where they do have data.
min_experiments <- 4

occupancy_filtered <- occupancy %>%
  filter(n_experiments_present >= min_experiments)

cat("\nGenes retained for averaged analysis (>=", min_experiments,
    "of 6 experiments):", nrow(occupancy_filtered), "of", nrow(occupancy), "\n")

# Two occupancy metrics to correlate against PC1/PC2, per Matthew's
# instruction: log2(occupancy) and rank(occupancy).
#
# INTERPRETIVE CHOICE (flagged for Matthew if this reading is wrong): rank
# here is computed fresh on the averaged RPKM value (rank_mean_occupancy),
# rather than averaging the pre-existing per-experiment rank columns
# (rankA...rankJ) that are already in the file. Both are defensible; this
# was chosen for consistency with how the log2 metric is derived (from the
# same averaged RPKM value), so both metrics describe the same "average
# occupancy" quantity, just transformed two different ways. The
# alternative (mean of provided per-experiment ranks) is computed below
# for comparison and reported alongside.
occupancy_filtered <- occupancy_filtered %>%
  mutate(
    log2_mean_occupancy = log2(mean_occupancy + 1),
    rank_mean_occupancy = rank(mean_occupancy, ties.method = "average"),
    mean_of_provided_ranks = rowMeans(select(., all_of(rank_cols)), na.rm = TRUE)
  )

cat("\nCorrelation between rank(mean occupancy) and mean of provided",
    "per-experiment ranks (sanity check - should be very high if the two",
    "approaches are broadly equivalent):\n")
cat("  Spearman r =",
    round(cor(occupancy_filtered$rank_mean_occupancy,
              occupancy_filtered$mean_of_provided_ranks,
              method = "spearman", use = "complete.obs"), 3), "\n")
```

```{r load-pc-loadings}
# PC1/PC2 gene loadings from the joint 93-sample PCA.
pca_genes <- read_tsv(file.path(tables_dir, "PCA_genes_AC.txt"),
                      show_col_types = FALSE)

pc_loadings <- pca_genes %>%
  select(gene_id, PC1, PC2)

cat("PC1/PC2 loadings available for", nrow(pc_loadings), "genes\n")
```

```{r join-occupancy-pc}
occ_pc <- occupancy_filtered %>%
  inner_join(pc_loadings, by = "gene_id")

cat("Genes with both occupancy (>=", min_experiments,
    "experiments) and PC1/PC2 loading:", nrow(occ_pc), "\n")
```

# Averaged occupancy vs PC1/PC2

```{r correlate-average}
# Pearson on log2(occupancy) (linearised magnitude), and Pearson on
# rank(occupancy) (equivalent to Spearman between occupancy and PC1/PC2,
# since ranking one side of a Pearson correlation and correlating against
# the raw values on the other side is the standard definition of Spearman
# when both variables are monotonically related).
run_correlation <- function(x, y, method = "pearson") {
  ct <- cor.test(x, y, method = method)
  tibble(r = unname(ct$estimate), p_value = ct$p.value)
}

avg_results <- bind_rows(
  run_correlation(occ_pc$log2_mean_occupancy, occ_pc$PC1) %>%
    mutate(metric = "log2_occupancy", pc = "PC1"),
  run_correlation(occ_pc$log2_mean_occupancy, occ_pc$PC2) %>%
    mutate(metric = "log2_occupancy", pc = "PC2"),
  run_correlation(occ_pc$rank_mean_occupancy, occ_pc$PC1) %>%
    mutate(metric = "rank_occupancy", pc = "PC1"),
  run_correlation(occ_pc$rank_mean_occupancy, occ_pc$PC2) %>%
    mutate(metric = "rank_occupancy", pc = "PC2")
) %>%
  select(pc, metric, r, p_value) %>%
  mutate(experiment = "Average", n_genes = nrow(occ_pc))

cat("Averaged occupancy - correlation with PC1/PC2:\n")
print(avg_results)
```

```{r plot-average-scatter, fig.width=11, fig.height=5}
# Scatter of log2(occupancy) vs PC1 and PC2, with binned mean overlay
# matching the visual convention used elsewhere in the dissertation
# (e.g. Figure 2). Useful for visually spotting saturation - genes where
# occupancy plateaus rather than continuing to rise with PC1/PC2.
plot_occupancy_scatter <- function(df, pc_col, pc_label, r_val) {
  df_binned <- df %>%
    mutate(pc_bin = ntile(.data[[pc_col]], 30)) %>%
    group_by(pc_bin) %>%
    summarise(
      pc_mean  = mean(.data[[pc_col]]),
      occ_mean = mean(log2_mean_occupancy),
      .groups = "drop"
    )

  ggplot(df, aes(x = .data[[pc_col]], y = log2_mean_occupancy)) +
    geom_point(color = project_colours[["blue"]], alpha = 0.15, size = 0.8) +
    geom_line(data = df_binned, aes(x = pc_mean, y = occ_mean),
              color = "black", linewidth = 0.8) +
    labs(
      title = paste0("RNAPII occupancy vs ", pc_label),
      subtitle = paste0("Pearson r = ", round(r_val, 3),
                        " | n = ", nrow(df), " genes"),
      x = pc_label,
      y = "log2(mean RNAPII occupancy + 1)"
    ) +
    theme_bw()
}

r_pc1 <- avg_results %>% filter(metric == "log2_occupancy", pc == "PC1") %>% pull(r)
r_pc2 <- avg_results %>% filter(metric == "log2_occupancy", pc == "PC2") %>% pull(r)

p_pc1 <- plot_occupancy_scatter(occ_pc, "PC1", "PC1 gene loading", r_pc1)
p_pc2 <- plot_occupancy_scatter(occ_pc, "PC2", "PC2 gene loading", r_pc2)

fig_occupancy_avg <- p_pc1 + p_pc2

print(fig_occupancy_avg)
```

# Per-experiment grid: does the relationship hold across all 6 experiments?

```{r correlate-per-experiment}
# Same two correlations (log2, rank), repeated independently for each of
# the 6 experiments, using that experiment's own RPKM/rank columns
# (not the averaged occupancy) and requiring only that this experiment's
# value is non-missing for a given gene - so per-experiment n differs
# slightly from the averaged analysis above.
run_per_experiment <- function(cond) {
  rpkm_col <- paste0("RPKM_mean_", cond)
  rank_col <- paste0("rank", cond)

  df <- occupancy %>%
    select(gene_id, all_of(rpkm_col), all_of(rank_col)) %>%
    rename(rpkm = all_of(rpkm_col), provided_rank = all_of(rank_col)) %>%
    filter(!is.na(rpkm)) %>%
    mutate(log2_occ = log2(rpkm + 1)) %>%
    inner_join(pc_loadings, by = "gene_id")

  bind_rows(
    run_correlation(df$log2_occ, df$PC1) %>% mutate(metric = "log2_occupancy", pc = "PC1"),
    run_correlation(df$log2_occ, df$PC2) %>% mutate(metric = "log2_occupancy", pc = "PC2"),
    run_correlation(df$provided_rank, df$PC1) %>% mutate(metric = "rank_occupancy", pc = "PC1"),
    run_correlation(df$provided_rank, df$PC2) %>% mutate(metric = "rank_occupancy", pc = "PC2")
  ) %>%
    mutate(experiment = cond, n_genes = nrow(df))
}

per_experiment_results <- map_dfr(occupancy_conditions, run_per_experiment)

grid_results <- bind_rows(avg_results, per_experiment_results) %>%
  mutate(
    experiment = factor(experiment,
                        levels = c("Average", occupancy_conditions)),
    p_adjust = p.adjust(p_value, method = "BH")
  ) %>%
  arrange(pc, metric, experiment)

cat("Full correlation grid (Average + 6 individual experiments):\n")
print(grid_results, n = Inf)
```

```{r plot-grid-heatmap, fig.width=9, fig.height=5}
# Grid heatmap: rows = experiment (Average + A-J), columns = PC x metric
# combination, fill = correlation coefficient. This is the "grid that
# summarises each of the separate correlations" Matthew asked for -
# lets you see at a glance whether the PC1/PC2 relationship is
# consistent across all 6 experiments or driven by a subset.
grid_results_plot <- grid_results %>%
  mutate(
    panel = paste0(pc, " (", metric, ")"),
    sig_label = if_else(p_adjust < 0.05, "*", "")
  )

p_grid <- ggplot(grid_results_plot,
                 aes(x = panel, y = experiment, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(round(r, 2), sig_label)), size = 3.5) +
  scale_fill_gradient2(
    low = project_colours[["blue"]], mid = "white",
    high = project_colours[["red"]], midpoint = 0
  ) +
  labs(
    title = "RNAPII occupancy vs PC1/PC2: correlation across experiments",
    subtitle = str_wrap(
      "Cell values: correlation coefficient (* = FDR < 0.05) | Average row uses occupancy averaged across all 6 experiments; other rows use each experiment individually",
      width = 75
    ),
    x = NULL,
    y = "Experiment",
    fill = "r"
  ) +
  theme_bw() +
  theme(
    axis.text.x  = element_text(angle = 30, hjust = 1),
    plot.subtitle = element_text(size = 9)
  )

print(p_grid)
```

```{r identify-saturating-genes}
# Candidate "saturating" genes: high PC1 or PC2 loading (i.e. strongly
# part of the transcriptional programme) but low occupancy rank relative
# to that - genes where transcriptional signal is high but RNAPII
# occupancy has not kept pace. Defined here as genes in the top quartile
# of |PC1| or |PC2| but bottom half of occupancy rank, using the averaged
# occupancy metric. This is an exploratory first pass, not a formal
# statistical test - intended as a starting gene list to inspect, not a
# final result.
saturation_candidates <- occ_pc %>%
  mutate(
    pc1_quartile = ntile(abs(PC1), 4),
    pc2_quartile = ntile(abs(PC2), 4),
    occupancy_half = ntile(rank_mean_occupancy, 2)
  ) %>%
  filter((pc1_quartile == 4 | pc2_quartile == 4) & occupancy_half == 1) %>%
  select(gene_id, gene_name, PC1, PC2, mean_occupancy,
         log2_mean_occupancy, rank_mean_occupancy) %>%
  arrange(desc(pmax(abs((.)$PC1), abs((.)$PC2))))

cat("Candidate saturating genes (top quartile |PC1| or |PC2|, bottom half",
    "occupancy rank):", nrow(saturation_candidates), "\n")
print(saturation_candidates, n = 30)
```

```{r save-outputs}
write_tsv(occ_pc, file.path(output_dir, "rnapii_occupancy_pc_joined.txt"))
write_tsv(grid_results, file.path(output_dir, "rnapii_occupancy_pc_correlation_grid.txt"))
write_tsv(saturation_candidates, file.path(output_dir, "rnapii_occupancy_saturation_candidates.txt"))

ggsave(file.path(figures_dir, "figure_rnapii_occupancy_vs_pc_scatter.png"),
       plot = fig_occupancy_avg, width = 11, height = 5, dpi = 300)
ggsave(file.path(figures_dir, "figure_rnapii_occupancy_pc_grid.png"),
       plot = p_grid, width = 9, height = 5, dpi = 300)

cat("Done - results saved to output_tables/ and figures/\n")
```
