---
title: "RNAPII Occupancy Saturation vs TF Direct Targets"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this file's location
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)
library(readxl)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Follow-up to the RNAPII occupancy vs PC1/PC2 analysis (script 11). The
binned-mean trend of log2(occupancy) vs PC1 rises then declines at the
highest-PC1 genes - the signature of promoter saturation. This script:

1. Defines the plateau/decline region from the binned-mean data,
2. Tests whether genes in that region are enriched among the direct
   targets of the TFs already identified as PC1/PC2 hits (CBF1, IFH1,
   SFP1, ERT1, RTG1, RTG3).

Part 2 extends the exclusion set to FHL1 (IFH1's DNA-binding partner)
and Dot6/Tod6 (repressors of the same programme), to see if excluding
them flattens the residual turnover further.

```{r load-occupancy-pc}
occ_pc <- read_tsv(file.path(output_dir, "rnapii_occupancy_pc_joined.txt"),
                   show_col_types = FALSE)

cat("Genes with occupancy + PC1/PC2:", nrow(occ_pc), "\n")
```

```{r identify-plateau-region}
# Recreate the 30-bin binned mean from script 11's scatter plot
n_bins <- 30

pc1_binned <- occ_pc %>%
  mutate(pc1_bin = ntile(PC1, n_bins)) %>%
  group_by(pc1_bin) %>%
  summarise(
    pc1_mean  = mean(PC1),
    occ_mean  = mean(log2_mean_occupancy),
    n_genes   = n(),
    .groups = "drop"
  ) %>%
  arrange(pc1_bin)

cat("Binned mean occupancy across PC1 (bin 30 = highest PC1):\n")
print(pc1_binned, n = Inf)

# Peak bin = where occ_mean is maximised; plateau = bins from peak onward
peak_bin <- pc1_binned$pc1_bin[which.max(pc1_binned$occ_mean)]
plateau_pc1_threshold <- pc1_binned %>%
  filter(pc1_bin == peak_bin) %>%
  pull(pc1_mean)

cat("\nPeak bin:", peak_bin, "of", n_bins,
    "| occupancy peaks at PC1 ~", round(plateau_pc1_threshold, 2), "\n")
cat("Bins", peak_bin, "to", n_bins, "(PC1 >=", round(plateau_pc1_threshold, 2),
    ") define the plateau/decline region.\n")
```

```{r define-saturation-candidates-refined}
# Genes at or beyond the peak PC1 bin (plateau/decline region)
saturation_candidates_refined <- occ_pc %>%
  filter(PC1 >= plateau_pc1_threshold) %>%
  arrange(desc(PC1)) %>%
  select(gene_id, gene_name, PC1, PC2, mean_occupancy,
         log2_mean_occupancy, rank_mean_occupancy)

cat("Genes in the PC1 plateau/decline region (PC1 >=",
    round(plateau_pc1_threshold, 2), "):",
    nrow(saturation_candidates_refined), "\n")
print(saturation_candidates_refined, n = 30)
```

```{r rebuild-tf-genesets}
# Rebuild all Table E direct-target gene sets (not just the 6 headline)
supp_path <- file.path(tables_dir, "Supp Table 3 updated.xlsx")
table_e   <- read_excel(supp_path, sheet = "Table-S3e")

tf_cols_e    <- colnames(table_e)[-1]
table_e_long <- table_e %>%
  pivot_longer(-gene_id, names_to = "TF", values_to = "log2FC") %>%
  filter(!is.na(log2FC))

genesets_e_activated <- table_e_long %>%
  filter(log2FC < 0) %>%
  group_by(TF) %>%
  summarise(genes = list(unique(gene_id)), .groups = "drop") %>%
  filter(map_int(genes, length) >= 5) %>%
  deframe() %>%
  set_names(paste0(names(.), "_direct_activated"))

genesets_e_repressed <- table_e_long %>%
  filter(log2FC > 0) %>%
  group_by(TF) %>%
  summarise(genes = list(unique(gene_id)), .groups = "drop") %>%
  filter(map_int(genes, length) >= 5) %>%
  deframe() %>%
  set_names(paste0(names(.), "_direct_repressed"))

genesets_e <- c(genesets_e_activated, genesets_e_repressed)

target_tf_sets <- c("CBF1_direct_activated", "IFH1_direct_activated",
                    "SFP1_direct_activated", "ERT1_direct_repressed",
                    "RTG1_direct_activated", "RTG3_direct_activated")

available_tf_sets <- intersect(target_tf_sets, names(genesets_e))
cat("Original 6 target TF gene sets available:", length(available_tf_sets),
    "of", length(target_tf_sets), "\n")
```

```{r list-available-tf-columns}
# All TF column names present in Table-S3e, for checking FHL1/Dot6/Tod6 naming
cat("All TF columns present in Table-S3e:\n")
print(sort(tf_cols_e))

cat("\nAll built gene set names (both directions):\n")
print(sort(names(genesets_e)))
```

```{r robustness-check-peak-excluding-tf-targets}
# Is the PC1 turnover point pulled leftward by the IFH1/CBF1/SFP1
# target genes? Recompute peak on genes excluding the original 6 TFs.
all_tf_target_genes <- unique(unlist(genesets_e[available_tf_sets]))

occ_pc_no_tf <- occ_pc %>%
  filter(!gene_id %in% all_tf_target_genes)

cat("Genes remaining after excluding original 6 TFs' direct targets:",
    nrow(occ_pc_no_tf), "of", nrow(occ_pc), "\n\n")

pc1_binned_no_tf <- occ_pc_no_tf %>%
  mutate(pc1_bin = ntile(PC1, n_bins)) %>%
  group_by(pc1_bin) %>%
  summarise(
    pc1_mean = mean(PC1),
    occ_mean = mean(log2_mean_occupancy),
    n_genes  = n(),
    .groups = "drop"
  ) %>%
  arrange(pc1_bin)

peak_bin_no_tf <- pc1_binned_no_tf$pc1_bin[which.max(pc1_binned_no_tf$occ_mean)]
plateau_pc1_threshold_no_tf <- pc1_binned_no_tf %>%
  filter(pc1_bin == peak_bin_no_tf) %>%
  pull(pc1_mean)

cat("With original 6 TF targets included - peak at PC1 ~",
    round(plateau_pc1_threshold, 2), "(bin", peak_bin, "of", n_bins, ")\n")
cat("With original 6 TF targets EXCLUDED  - peak at PC1 ~",
    round(plateau_pc1_threshold_no_tf, 2), "(bin", peak_bin_no_tf, "of", n_bins, ")\n\n")
```

```{r robustness-check-expanded-tf-set}
## PART 2: expanded exclusion set, adding FHL1 and Dot6/Tod6 to the
## original 6, to try flattening the residual turnover further.
##
## Uses str_detect() rather than exact matching since column naming
## isn't confirmed - matched names are printed below for verification.
related_tf_patterns <- c("FHL1", "DOT6", "TOD6")

related_tf_sets <- names(genesets_e)[
  str_detect(str_to_upper(names(genesets_e)),
             paste(related_tf_patterns, collapse = "|"))
]

cat("Related TF gene sets matched (FHL1/Dot6/Tod6, either direction):\n")
print(related_tf_sets)

if (length(related_tf_sets) == 0) {
  cat("\nWARNING: none of FHL1/Dot6/Tod6 matched any column in Table-S3e.",
      "Check the list-available-tf-columns output above.\n")
}

expanded_tf_sets <- unique(c(available_tf_sets, related_tf_sets))
cat("\nExpanded exclusion set:", length(expanded_tf_sets), "gene sets",
    "(original 6 +", length(related_tf_sets), "newly matched):\n")
print(expanded_tf_sets)

expanded_tf_target_genes <- unique(unlist(genesets_e[expanded_tf_sets]))

occ_pc_no_expanded_tf <- occ_pc %>%
  filter(!gene_id %in% expanded_tf_target_genes)

cat("\nGenes remaining after excluding EXPANDED TF set's direct targets:",
    nrow(occ_pc_no_expanded_tf), "of", nrow(occ_pc), "\n")

pc1_binned_no_expanded_tf <- occ_pc_no_expanded_tf %>%
  mutate(pc1_bin = ntile(PC1, n_bins)) %>%
  group_by(pc1_bin) %>%
  summarise(
    pc1_mean = mean(PC1),
    occ_mean = mean(log2_mean_occupancy),
    n_genes  = n(),
    .groups = "drop"
  ) %>%
  arrange(pc1_bin)

peak_bin_expanded <- pc1_binned_no_expanded_tf$pc1_bin[which.max(pc1_binned_no_expanded_tf$occ_mean)]
plateau_pc1_threshold_expanded <- pc1_binned_no_expanded_tf %>%
  filter(pc1_bin == peak_bin_expanded) %>%
  pull(pc1_mean)

cat("With expanded TF targets EXCLUDED - peak at PC1 ~",
    round(plateau_pc1_threshold_expanded, 2), "(bin", peak_bin_expanded,
    "of", n_bins, ")\n")
```

```{r quantify-flattening}
## Flatness metric: decline in binned-mean occupancy from peak bin to
## the final bin, compared across all genes / excluding original 6 /
## excluding expanded set.
compute_decline <- function(df_binned, peak_bin_val) {
  peak_occ  <- df_binned %>% filter(pc1_bin == peak_bin_val) %>% pull(occ_mean)
  final_occ <- df_binned %>% filter(pc1_bin == n_bins) %>% pull(occ_mean)
  peak_occ - final_occ  # positive = decline from peak to final bin
}

decline_all      <- compute_decline(pc1_binned, peak_bin)
decline_orig6    <- compute_decline(pc1_binned_no_tf, peak_bin_no_tf)
decline_expanded <- compute_decline(pc1_binned_no_expanded_tf, peak_bin_expanded)

flatness_summary <- tibble(
  gene_set = c("All genes",
              "Excluding original 6 TFs",
              paste0("Excluding expanded TF set (+", length(related_tf_sets), " related)")),
  peak_to_final_decline = c(decline_all, decline_orig6, decline_expanded)
)

cat("Flatness comparison (smaller decline = flatter curve):\n")
print(flatness_summary)

if (decline_expanded < decline_orig6) {
  cat("\nExpanding the exclusion set REDUCED the residual decline",
      "(curve got flatter).\n")
} else if (length(related_tf_sets) > 0) {
  cat("\nExpanding the exclusion set did NOT reduce the residual decline",
      "- turnover persists even excluding FHL1/Dot6/Tod6 targets.\n")
}
```

```{r plot-robustness-check-expanded, fig.width=10, fig.height=6}
# Three-line comparison: all genes, excluding original 6, excluding expanded set
binned_comparison_expanded <- bind_rows(
  pc1_binned                %>% mutate(gene_set = "All genes"),
  pc1_binned_no_tf          %>% mutate(gene_set = "Excluding original 6 TFs"),
  pc1_binned_no_expanded_tf %>% mutate(gene_set = paste0(
    "Excluding expanded TF set (+FHL1/Dot6/Tod6)"))
)

p_robustness_expanded <- ggplot(binned_comparison_expanded,
                                aes(x = pc1_mean, y = occ_mean, color = gene_set)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = c(
    "All genes" = project_colours[["grey"]],
    "Excluding original 6 TFs" = project_colours[["blue"]],
    "Excluding expanded TF set (+FHL1/Dot6/Tod6)" = project_colours[["red"]]
  )) +
  labs(
    title = str_wrap(
      "Does excluding related TFs (FHL1, Dot6, Tod6) flatten the PC1 occupancy plateau further?",
      width = 55
    ),
    subtitle = "Binned mean log2(occupancy) vs PC1 gene loading",
    x = "PC1 gene loading (binned mean)",
    y = "log2(mean RNAPII occupancy + 1), binned mean",
    color = NULL
  ) +
  theme_bw(base_size = 15) +
  theme(
    plot.title      = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    legend.text     = element_text(size = 11),
    axis.title      = element_text(size = 15),
    axis.text       = element_text(size = 13)
  )

print(p_robustness_expanded)
```

```{r save-outputs}
write_tsv(saturation_candidates_refined,
          file.path(output_dir, "rnapii_saturation_candidates_refined.txt"))
write_tsv(flatness_summary,
          file.path(output_dir, "rnapii_saturation_flatness_comparison.txt"))

# Canvas widened for the wrapped title
ggsave(file.path(figures_dir, "figure_rnapii_saturation_robustness_check_expanded.png"),
       plot = p_robustness_expanded, width = 13, height = 7, dpi = 300)

cat("Done - results saved to output_tables/ and figures/\n")
```
