---
title: "RNAPII Occupancy Follow-up: EFG/H Only, Transcription Rate Validation, Bcy1-AID+PhIAA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this Rmd's location (Adriana_analysis/scripts/)
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Follow-up to scripts 11/12, per Matthew's feedback (Slack):

1. A/B/C vs EFG/H/J is a nutrient-condition difference, not equivalent
   experiments as originally assumed. Adriana's data matches the nutrient
   conditions of EFG and H specifically, so only those two should be
   compared against PC1/PC2.
2. EFG and H are cycling (not arrested) small cells, so cell-cycle genes
   should be filtered out before repeating the correlation.
3. Additional validation: correlate occupancy against published
   transcription rate estimates already in the annotation file
   (miller_synth_rate, neymotin_synth_rate), and against Adriana's own
   unnormalised scTPM at the first (smallest-cell) time point.
4. Key figure: Bcy1-AID + PhIAA data (small vs large cells) plotted
   against transcription rate, to show the occupancy-transcription
   correlation breaks down under direct signalling perturbation - a
   direct test against the promoter-saturation model (Vidal et al.),
   which would predict this correlation should hold even under
   perturbation if saturation were driving transcriptional reprogramming.

ASSUMPTIONS MADE BELOW THAT NEED CHECKING AGAINST WHAT MATTHEW ACTUALLY
MEANT (each flagged again at the relevant point in the script):
- "Adriana's first time point" = condition "OD 0.1", arrest "1hr G1...",
  exp_group_3 "A", strain MS69 (WT) - the earliest/smallest sample in
  what looks like a separate OD-timecourse series, distinct from the
  drug-perturbation panel (exp_group_3 "B").
- Bcy1-AID + PhIAA = strain_id MS756 / MS756c2, strain_info "AID-Bcy1",
  condition "PhIAA" vs "DMSO" (matched vehicle control), at "1hr G1..."
  (small) vs "Large G..." (large) arrest.
- MS756 and MS756c2 are treated as biological replicates of the same
  strain and averaged together.
- "Plot vs transcription rate ... to show loss of correlation" is built
  as: occupancy (fixed, from the published ChIP data) vs the
  perturbation-induced transcriptional change (log2FC PhIAA/DMSO),
  separately in small and large cells. An alternative, simpler reading
  (occupancy vs absolute TPM level, not log2FC) is also included as a
  secondary check.

```{r load-occupancy-raw}
occupancy_path <- file.path(
  tables_dir,
  "210601_gene_annotation_RPB_occupancy.txt"
)

occupancy_raw <- read_tsv(occupancy_path, show_col_types = FALSE)

cat("Occupancy data:", nrow(occupancy_raw), "genes x",
    ncol(occupancy_raw), "columns\n")
```

# Part A: Occupancy vs PC1/PC2, restricted to EFG and H, cell-cycle genes excluded

```{r select-efg-h-and-cellcycle-flags}
occupancy_efgh <- occupancy_raw %>%
  select(systematic_name, gene_name,
         RPKM_mean_EFG, RPKM_mean_H, rankEFG, rankH,
         cell_cycle_high_confidence, cell_cycle_low_confidence,
         Campbell_2020_cell_cycle_phase, Spellman_1998_cell_cycle_phase,
         Lichtenberg_2004_cell_cycle_phase) %>%
  rename(gene_id = systematic_name) %>%
  distinct(gene_id, .keep_all = TRUE)  # YLR352W, YMR035W, YOR194C were
                                        # duplicated in the raw annotation
                                        # file (confirmed identical/rounding-
                                        # noise-only rows, not real
                                        # isoform/ORF differences).
                                        # Keeps first occurrence. Clears the
                                        # many-to-many join warnings that
                                        # otherwise fire on every downstream
                                        # join against occ_efgh_pc (Parts
                                        # A/B/D and Figure 18).

cat("Cell-cycle flag counts (high confidence):\n")
print(table(occupancy_efgh$cell_cycle_high_confidence, useNA = "ifany"))
```

```{r filter-cellcycle-genes}
# Excluding genes flagged TRUE for cell_cycle_high_confidence, per
# Matthew's instruction that EFG/H are cycling cells and cell-cycle
# genes should be removed before correlating. Using the high-confidence
# flag (not low-confidence) as the exclusion criterion - a stricter
# annotation than low-confidence, chosen to avoid removing genes that
# aren't robustly cell-cycle-regulated. If Matthew intended the broader
# low-confidence exclusion instead, swap the filter below.
occupancy_efgh_filtered <- occupancy_efgh %>%
  filter(is.na(cell_cycle_high_confidence) | cell_cycle_high_confidence == FALSE)

cat("Genes retained after excluding high-confidence cell-cycle genes:",
    nrow(occupancy_efgh_filtered), "of", nrow(occupancy_efgh), "\n")
```

```{r compute-efgh-average}
occupancy_efgh_filtered <- occupancy_efgh_filtered %>%
  rowwise() %>%
  mutate(
    n_efgh_present  = sum(!is.na(c(RPKM_mean_EFG, RPKM_mean_H))),
    mean_occupancy_efgh = mean(c(RPKM_mean_EFG, RPKM_mean_H), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(n_efgh_present >= 1) %>%
  mutate(
    log2_occupancy_efgh = log2(mean_occupancy_efgh + 1),
    rank_occupancy_efgh = rank(mean_occupancy_efgh, ties.method = "average")
  )

cat("Genes with at least one of EFG/H present (post cell-cycle filter):",
    nrow(occupancy_efgh_filtered), "\n")
```

```{r join-pc-efgh}
pca_genes <- read_tsv(file.path(tables_dir, "PCA_genes_AC.txt"),
                      show_col_types = FALSE)

pc_loadings <- pca_genes %>% select(gene_id, PC1, PC2)

occ_efgh_pc <- occupancy_efgh_filtered %>%
  inner_join(pc_loadings, by = "gene_id")

cat("Genes with EFG/H occupancy (cell-cycle-filtered) + PC1/PC2:",
    nrow(occ_efgh_pc), "\n")
```

```{r correlate-efgh}
run_correlation <- function(x, y, method = "pearson") {
  ct <- cor.test(x, y, method = method)
  tibble(r = unname(ct$estimate), p_value = ct$p.value)
}

efgh_results <- bind_rows(
  run_correlation(occ_efgh_pc$log2_occupancy_efgh, occ_efgh_pc$PC1) %>%
    mutate(metric = "log2_occupancy", pc = "PC1"),
  run_correlation(occ_efgh_pc$log2_occupancy_efgh, occ_efgh_pc$PC2) %>%
    mutate(metric = "log2_occupancy", pc = "PC2"),
  run_correlation(occ_efgh_pc$rank_occupancy_efgh, occ_efgh_pc$PC1) %>%
    mutate(metric = "rank_occupancy", pc = "PC1"),
  run_correlation(occ_efgh_pc$rank_occupancy_efgh, occ_efgh_pc$PC2) %>%
    mutate(metric = "rank_occupancy", pc = "PC2")
) %>%
  select(pc, metric, r, p_value) %>%
  mutate(n_genes = nrow(occ_efgh_pc))

cat("EFG/H occupancy (cell-cycle genes excluded) vs PC1/PC2:\n")
print(efgh_results)

cat("\nFor comparison, the original all-experiment-average result",
    "(script 11, no cell-cycle filter) was approximately: PC1 log2 r =",
    "0.22, PC2 log2 r = 0.03. Compare against efgh_results above to see",
    "whether restricting to EFG/H and removing cell-cycle genes",
    "strengthens, weakens, or leaves unchanged the PC1 relationship.\n")
```

```{r plot-efgh-scatter, fig.width=15, fig.height=5.5}
# Titles shortened (moved the "RNAPII occupancy (EFG/H, cell-cycle
# genes excluded)" portion to a shared caption instead of repeating it
# in both panel titles) and the combined figure widened, since the two
# original panel titles were long enough to overlap when placed side
# by side at the previous width.
plot_occupancy_scatter <- function(df, pc_col, pc_label, r_val, occ_col) {
  df_binned <- df %>%
    mutate(pc_bin = ntile(.data[[pc_col]], 30)) %>%
    group_by(pc_bin) %>%
    summarise(
      pc_mean  = mean(.data[[pc_col]]),
      occ_mean = mean(.data[[occ_col]]),
      .groups = "drop"
    )

  ggplot(df, aes(x = .data[[pc_col]], y = .data[[occ_col]])) +
    geom_point(color = project_colours[["blue"]], alpha = 0.15, size = 0.8) +
    geom_line(data = df_binned, aes(x = pc_mean, y = occ_mean),
              color = "black", linewidth = 0.8) +
    labs(
      title = paste0("vs ", pc_label),
      subtitle = paste0("Pearson r = ", round(r_val, 3),
                        " | n = ", nrow(df), " genes"),
      x = pc_label,
      y = "log2(mean EFG/H occupancy + 1)"
    ) +
    theme_bw() +
    theme(plot.title = element_text(size = 11))
}

r_pc1_efgh <- efgh_results %>% filter(metric == "log2_occupancy", pc == "PC1") %>% pull(r)
r_pc2_efgh <- efgh_results %>% filter(metric == "log2_occupancy", pc == "PC2") %>% pull(r)

p_efgh_pc1 <- plot_occupancy_scatter(occ_efgh_pc, "PC1", "PC1 gene loading",
                                     r_pc1_efgh, "log2_occupancy_efgh")
p_efgh_pc2 <- plot_occupancy_scatter(occ_efgh_pc, "PC2", "PC2 gene loading",
                                     r_pc2_efgh, "log2_occupancy_efgh")

fig_efgh <- (p_efgh_pc1 + p_efgh_pc2) +
  plot_annotation(
    title = "RNAPII occupancy (EFG/H, cell-cycle genes excluded)",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  )

print(fig_efgh)
```

# Part B: Validate occupancy against published transcription rate annotations

```{r load-synth-rate-columns}
synth_rates <- occupancy_raw %>%
  select(systematic_name, miller_synth_rate, neymotin_synth_rate) %>%
  rename(gene_id = systematic_name)

occ_efgh_synth <- occ_efgh_pc %>%
  left_join(synth_rates, by = "gene_id")

cat("Genes with Miller synth rate:",
    sum(!is.na(occ_efgh_synth$miller_synth_rate)), "\n")
cat("Genes with Neymotin synth rate:",
    sum(!is.na(occ_efgh_synth$neymotin_synth_rate)), "\n")
```

```{r correlate-synth-rates}
synth_rate_results <- bind_rows(
  occ_efgh_synth %>%
    filter(!is.na(miller_synth_rate)) %>%
    { run_correlation(.$log2_occupancy_efgh, .$miller_synth_rate, method = "spearman") } %>%
    mutate(source = "Miller synth rate"),
  occ_efgh_synth %>%
    filter(!is.na(neymotin_synth_rate)) %>%
    { run_correlation(.$log2_occupancy_efgh, .$neymotin_synth_rate, method = "spearman") } %>%
    mutate(source = "Neymotin synth rate")
)

cat("EFG/H occupancy vs published transcription (synthesis) rate estimates",
    "(Spearman, as a validation check that occupancy tracks independently",
    "measured transcription rate):\n")
print(synth_rate_results)
```

# Part C: Validate occupancy against Adriana's raw scTPM at the first time point

```{r identify-first-timepoint-sample}
# "Adriana's first time point": interpreted as the OD-timecourse series
# (exp_group_3 == "A", distinct from the drug-perturbation panel where
# exp_group_3 == "B"), earliest point (OD 0.1, 1hr G1 arrest, WT strain
# MS69). CHECK: confirm this is the correct sample before trusting the
# result below - if wrong, adjust the filter conditions.
df_sc_final <- read_tsv(file.path(tables_dir, "df_sc_final.txt"),
                        show_col_types = FALSE)

first_timepoint_samples <- df_sc_final %>%
  filter(exp_group_3 == "A", condition == "OD 0.1",
         str_detect(arrest, "^1hr")) %>%
  distinct(sample, sample_short, condition, arrest, strain_id, mean_cell_volume)

cat("Sample(s) identified as 'Adriana's first time point':\n")
print(first_timepoint_samples)

if (nrow(first_timepoint_samples) != 1) {
  cat("\nWARNING: expected exactly 1 sample, found", nrow(first_timepoint_samples),
      "- check the filter conditions above before trusting Part C results.\n")
}
```

```{r correlate-first-timepoint-tpm}
first_tp_tpm <- df_sc_final %>%
  filter(sample %in% first_timepoint_samples$sample) %>%
  select(gene_id, TPM) %>%
  distinct()

occ_efgh_firsttp <- occ_efgh_pc %>%
  inner_join(first_tp_tpm, by = "gene_id") %>%
  filter(!is.na(TPM))

cat("Genes with both EFG/H occupancy and first-timepoint raw TPM:",
    nrow(occ_efgh_firsttp), "\n")

first_tp_correlation <- run_correlation(
  occ_efgh_firsttp$log2_occupancy_efgh, occ_efgh_firsttp$TPM,
  method = "spearman"
)

cat("\nEFG/H occupancy vs Adriana's raw (unnormalised) TPM at the first",
    "time point (Spearman):\n")
print(first_tp_correlation)
```

# Part D: Bcy1-AID + PhIAA - does occupancy predict the perturbation response?

```{r identify-bcy1-samples}
# Bcy1-AID + PhIAA vs DMSO, small ("1hr G1...") vs large ("Large G...")
# arrest. MS756 and MS756c2 treated as biological replicates of the same
# AID-Bcy1 strain and averaged. CHECK: confirm arrest string matching
# below actually captures the intended small/large samples (arrest
# column values are truncated in the console output used to identify
# them, so str_detect patterns are a best guess at the full strings).
bcy1_samples <- df_sc_final %>%
  filter(strain_id %in% c("MS756", "MS756c2"),
         condition %in% c("PhIAA", "DMSO")) %>%
  mutate(
    cell_size_group = case_when(
      str_detect(arrest, "^1hr")  ~ "small",
      str_detect(arrest, "^Large") ~ "large",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(cell_size_group))

cat("Bcy1-AID samples identified:\n")
print(bcy1_samples %>%
        distinct(sample, condition, arrest, cell_size_group, strain_id,
                 mean_cell_volume))
```

```{r compute-bcy1-log2fc}
# Mean TPMsc per gene x condition x cell_size_group (averaging replicate
# samples, including across MS756/MS756c2), then log2FC(PhIAA/DMSO)
# within each cell size group separately.
bcy1_mean_tpm <- bcy1_samples %>%
  group_by(gene_id, condition, cell_size_group) %>%
  summarise(mean_tpmsc = mean(TPMsc, na.rm = TRUE), .groups = "drop")

bcy1_log2fc <- bcy1_mean_tpm %>%
  pivot_wider(names_from = condition, values_from = mean_tpmsc) %>%
  filter(!is.na(PhIAA), !is.na(DMSO), PhIAA > 0, DMSO > 0) %>%
  mutate(log2fc_phiaa_dmso = log2(PhIAA / DMSO))

cat("Genes with log2FC(PhIAA/DMSO) computed, by cell size group:\n")
print(table(bcy1_log2fc$cell_size_group))
```

```{r join-bcy1-occupancy}
bcy1_occ <- bcy1_log2fc %>%
  inner_join(
    occ_efgh_pc %>% select(gene_id, log2_occupancy_efgh, PC1, PC2),
    by = "gene_id"
  )

cat("Genes with both Bcy1 log2FC and EFG/H occupancy:", nrow(bcy1_occ), "\n")
```

```{r correlate-bcy1-occupancy}
bcy1_correlation <- bcy1_occ %>%
  group_by(cell_size_group) %>%
  summarise(
    r = cor(log2_occupancy_efgh, log2fc_phiaa_dmso, method = "spearman",
            use = "complete.obs"),
    p_value = cor.test(log2_occupancy_efgh, log2fc_phiaa_dmso,
                       method = "spearman")$p.value,
    n = n(),
    .groups = "drop"
  )

cat("Correlation between static occupancy and PhIAA-induced transcriptional",
    "change (log2FC PhIAA/DMSO), separately in small and large cells.\n")
cat("Under a strict promoter-saturation model, occupancy alone should",
    "largely determine this response. Report both r and r^2 below:",
    "r describes the strength/direction of any relationship, r^2",
    "describes how much of the response variance occupancy actually",
    "explains - a moderate r with a small r^2 (as found here, ~10-12%)",
    "indicates occupancy has *some* relationship to the response but is",
    "far from sufficient to explain it, which is the precise claim to",
    "report rather than 'no correlation':\n")
print(bcy1_correlation %>% mutate(r2_pct = round(r^2 * 100, 1)))
```

```{r plot-bcy1-occupancy, fig.width=11, fig.height=5}
# Title/subtitle corrected to match the actual result: bcy1_correlation
# (computed in the chunk above) shows r ~ 0.31-0.35 in both cell size
# groups, p < 1e-100, n = 4392 - a moderate, highly significant
# correlation, NOT "no correlation" as an earlier version of this figure
# claimed. r^2 (~10-12%) is used in the subtitle since that is the more
# interpretable quantity here: it states plainly how much of the
# transcriptional response variance is (and, more importantly, is NOT)
# explained by static occupancy alone. Values are pulled directly from
# bcy1_correlation rather than hardcoded, so the figure stays accurate
# if the underlying data/filtering changes on a re-run.
#
# NOTE: this panel plots occupancy vs the DRUG effect (log2FC
# PhIAA/DMSO), kept here as a sanity check / secondary reading. Figure
# 18 proper (Part E below) plots the SIZE effect (log2FC large/small)
# instead, per the restructure - see Part E for the primary figure.
bcy1_corr_label <- bcy1_correlation %>%
  mutate(
    r2 = r^2,
    label = paste0(
      str_to_title(cell_size_group), ": r = ", round(r, 2),
      " (r\u00b2 = ", round(r2 * 100, 1), "%)"
    )
  ) %>%
  arrange(cell_size_group) %>%
  pull(label) %>%
  paste(collapse = "  |  ")

p_bcy1 <- ggplot(bcy1_occ, aes(x = log2_occupancy_efgh, y = log2fc_phiaa_dmso)) +
  geom_point(color = project_colours[["blue"]], alpha = 0.3, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = project_colours[["grey"]]) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.7) +
  facet_wrap(~cell_size_group, labeller = labeller(
    cell_size_group = c(small = "Small cells", large = "Large cells")
  )) +
  labs(
    subtitle = paste0(
      "Static occupancy vs log2 FC (PhIAA/DMSO)  |  ", bcy1_corr_label
    ),
    x = "log2(mean RNAPII occupancy + 1)",
    y = "log2 FC (PhIAA/DMSO)"
  ) +
  theme_bw() +
  theme(
    plot.subtitle   = element_text(size = 8.5),
    strip.background = element_blank(),
    strip.text        = element_text(face = "bold", size = 11)
  )

print(p_bcy1)
```

```{r plot-bcy1-occupancy-alternative, fig.width=11, fig.height=5}
# Secondary/alternative reading: occupancy vs absolute TPM level (not
# log2FC) within PhIAA-treated cells only, small vs large - a simpler
# version of "plot occupancy vs transcription rate in small and large
# cells" in case this, rather than the log2FC version above, is what
# Matthew meant.
bcy1_absolute <- bcy1_mean_tpm %>%
  filter(condition == "PhIAA") %>%
  mutate(log2_tpmsc = log2(mean_tpmsc + 1)) %>%
  inner_join(
    occ_efgh_pc %>% select(gene_id, log2_occupancy_efgh),
    by = "gene_id"
  )

bcy1_absolute_correlation <- bcy1_absolute %>%
  group_by(cell_size_group) %>%
  summarise(
    r = cor(log2_occupancy_efgh, log2_tpmsc, method = "spearman",
            use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )

cat("Alternative version: occupancy vs absolute PhIAA-treated expression",
    "level (not log2FC), small vs large cells:\n")
print(bcy1_absolute_correlation)

p_bcy1_absolute <- ggplot(bcy1_absolute, aes(x = log2_occupancy_efgh, y = log2_tpmsc)) +
  geom_point(color = project_colours[["red"]], alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.7) +
  facet_wrap(~cell_size_group, labeller = labeller(
    cell_size_group = c(small = "Small cells", large = "Large cells")
  )) +
  labs(
    subtitle = "Alternative reading: static occupancy vs PhIAA-treated log2(TPMsc), small vs large cells",
    x = "log2(mean RNAPII occupancy + 1)",
    y = "log2(TPMsc + 1), PhIAA-treated"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text        = element_text(face = "bold", size = 11)
  )

print(p_bcy1_absolute)
```

# Part E: Figure 18 -- WT+PhIAA and Bcy1-AID+PhIAA, size effect vs occupancy/TPM/transcription rate

```{r figure18-assumptions-flag}
## FIGURE 18 -- RESOLVED so far (see figure18-diagnose-samples output):
##
## 1. BATCH: "exp250624" is confirmed as a substring of `sample` for
##    both MS69 and MS756 PhIAA rows. CONFIRMED.
## 2. ARREST STRING MAPPING is strain-specific (this was the actual bug
##    in the first version): MS69 uses "1hr G1 arrest"/"4.75hr G1
##    arrest", MS756 uses "1hr G1 arrest"/"Large G1 arrest" for what is
##    the same small/large timepoint pair. Confirmed via
##    mean_cell_volume (~65-69 fL small, ~243-268 fL large, consistent
##    across both strains). CONFIRMED, implemented in
##    figure18-identify-samples below.
## 3. MS756c2 is absent from batch exp250624 entirely - "Bcy1-AID" in
##    Figure 18 means MS756 alone here, unlike Part D above which pools
##    MS756/MS756c2 from a different batch. NOTED, not a bug.
## 4. Each strain x size x PhIAA combination in this batch has exactly
##    one replicate. Thinner than Part D's pooled comparison - flagged
##    as an open question for Matthew (worth pooling in a second batch,
##    or is single-replicate-per-batch fine for this figure?), not
##    blocking.
## 5. NO DMSO NEEDED: the y-axis here is the SIZE effect (log2FC
##    large/small, WITHIN PhIAA only), not the drug effect - so
##    exp250624 lacking a DMSO arm doesn't block this analysis. Part D
##    above correctly still uses the DMSO-paired batch (exp241010) for
##    its own (different) drug-effect figure.
##
## STILL OPEN / NOT YET CONFIRMED:
## 6. TPM BASELINE: defined as each gene's small-cell PhIAA TPMsc (log2)
##    - the starting expression level before the size change, not
##    large-cell or a mean of both. Swap `cell_size_group == "small"`
##    in figure18-join-predictors below if a different baseline is
##    wanted.
## 7. TRANSCRIPTION RATE: Miller and Neymotin synthesis-rate estimates
##    are averaged into one `mean_synth_rate` predictor (same approach
##    already used for EFG/H occupancy) rather than shown as two
##    separate panels. Flag if the two sources should stay separate.
```

```{r figure18-diagnose-samples}
## DIAGNOSTIC - run this first. The previous version of this script
## guessed at both the batch substring ("exp250624" inside `sample`)
## and the arrest strings ("1hr-arrest" / "4hr45m-arrest") without
## being able to see the actual data, and the guess was wrong (empty
## match -> pivot_wider produced no "small" column -> cryptic "object
## 'small' not found" error downstream). This chunk shows the real
## values so the filter below can be built from what's actually there
## instead of guessed again.

cat("All MS69/MS756/MS756c2 + PhIAA rows, regardless of batch guess:\n")
df_sc_final %>%
  filter(strain_id %in% c("MS69", "MS756", "MS756c2"),
         condition == "PhIAA") %>%
  distinct(sample, sample_short, strain_id, condition, arrest,
           mean_cell_volume) %>%
  arrange(strain_id, arrest) %>%
  print(n = 100)

cat("\nDoes any sample string actually contain 'exp250624'?\n")
df_sc_final %>%
  filter(str_detect(sample, "exp250624") |
         str_detect(sample_short, "exp250624")) %>%
  distinct(sample, sample_short) %>%
  print(n = 20)

cat("\nAll distinct arrest strings for MS69 and MS756/MS756c2 (any condition):\n")
df_sc_final %>%
  filter(strain_id %in% c("MS69", "MS756", "MS756c2")) %>%
  distinct(strain_id, arrest) %>%
  arrange(strain_id, arrest) %>%
  print(n = 50)
```

```{r figure18-identify-samples}
## WT (MS69) and Bcy1-AID (MS756) under PhIAA, batch exp250624 -
## confirmed by diagnostic output (see figure18-diagnose-samples above
## and the conversation this was resolved in):
##
## - BATCH: "exp250624" IS a substring of `sample` - the earlier guess
##   was right on this part.
## - AREST STRING MAPPING is strain-specific, not a single shared
##   pattern - this was the actual bug. The `arrest` column uses
##   different labels per strain for what is the same underlying
##   timepoint/batch:
##     MS69:  "1hr G1 arrest" (small) / "4.75hr G1 arrest" (large)
##     MS756: "1hr G1 arrest" (small) / "Large G1 arrest"   (large)
##   Confirmed via mean_cell_volume: ~65-69 fL for all "1hr" rows,
##   ~243-268 fL for all "4.75hr"/"Large" rows in this batch - both
##   strains land on the same small/large volume scale.
## - MS756c2 is NOT present in exp250624 at all (only MS756 is) - so
##   "Bcy1-AID+PhIAA" here means MS756 alone, not the MS756/MS756c2
##   average used in Part D above. This is a real feature of the data,
##   not a bug - noted in the sample count check below.
## - Each strain x size x condition combination in this batch has
##   exactly ONE sample (single replicate) - thinner than Part D's
##   pooled-batch comparison, but consistent with how single-sample
##   TPMsc comparisons are already used elsewhere in this pipeline
##   (e.g. df_ms1039_wt in CW7). Flagged here rather than blocking on
##   it - worth deciding with Matthew whether a second batch should be
##   pooled in if within-batch, single-replicate is judged too thin.

BATCH_STRING <- "exp250624"

fig18_samples <- df_sc_final %>%
  filter(str_detect(sample, BATCH_STRING),
         strain_id %in% c("MS69", "MS756", "MS756c2"),
         condition == "PhIAA") %>%
  mutate(
    strain_group = case_when(
      strain_id == "MS69"                  ~ "WT",
      strain_id %in% c("MS756", "MS756c2") ~ "Bcy1-AID",
      TRUE                                 ~ NA_character_),
    cell_size_group = case_when(
      str_detect(arrest, "^1hr")                                    ~ "small",
      strain_id == "MS69"  & str_detect(arrest, "^4\\.75hr")        ~ "large",
      strain_id != "MS69"  & str_detect(arrest, "^Large")           ~ "large",
      TRUE ~ NA_character_)
  ) %>%
  filter(!is.na(strain_group), !is.na(cell_size_group))

cat("Figure 18 sample identification:\n")
print(fig18_samples %>%
        distinct(sample, strain_id, strain_group, condition, arrest,
                 cell_size_group, mean_cell_volume))

cat("\nSample counts per strain x size (expect exactly 1 each, per the",
    "single-replicate caveat above - MS756c2 will show 0 rows since it",
    "isn't in this batch, which is expected, not an error):\n")
print(fig18_samples %>% count(strain_id, cell_size_group))

if (nrow(fig18_samples) == 0) {
  stop("fig18_samples is empty - re-check BATCH_STRING and the arrest ",
       "patterns above against a fresh figure18-diagnose-samples run.")
}

for (grp in c("WT", "Bcy1-AID")) {
  sizes_present <- fig18_samples %>%
    filter(strain_group == grp) %>%
    pull(cell_size_group) %>%
    unique()
  if (!all(c("small", "large") %in% sizes_present)) {
    cat("\nWARNING:", grp, "is missing one of small/large (has:",
        paste(sizes_present, collapse = ", "),
        ") - the size FC pivot below will fail for this strain.\n")
  }
}
```

```{r figure18-compute-sizefc}
## Per-gene mean TPMsc by strain x cell size (averaging replicate
## samples, including across MS756/MS756c2), then log2FC(large/small)
## within each strain separately - this is the SIZE effect, computed
## entirely within the PhIAA condition.

fig18_mean_tpm <- fig18_samples %>%
  group_by(gene_id, strain_group, cell_size_group) %>%
  summarise(mean_tpmsc = mean(TPMsc, na.rm = TRUE), .groups = "drop")

fig18_wide_check <- fig18_mean_tpm %>%
  pivot_wider(names_from = cell_size_group, values_from = mean_tpmsc)

if (!all(c("small", "large") %in% colnames(fig18_wide_check))) {
  stop("fig18_mean_tpm is missing a 'small' or 'large' column after ",
       "pivoting (has: ", paste(setdiff(colnames(fig18_wide_check),
       c("gene_id", "strain_group")), collapse = ", "), "). This means ",
       "fig18_samples had no rows for that size group for at least one ",
       "strain - go back to figure18-identify-samples and fix ",
       "SMALL_ARREST_STRING/LARGE_ARREST_STRING/BATCH_STRING.")
}

fig18_sizefc <- fig18_wide_check %>%
  filter(!is.na(small), !is.na(large), small > 0, large > 0) %>%
  mutate(log2fc_size = log2(large / small))

cat("Genes with size FC (large/small, PhIAA, exp250624) computed, by strain:\n")
print(table(fig18_sizefc$strain_group))
```

```{r figure18-join-predictors}
## TPM baseline: small-cell PhIAA TPMsc (log2), per strain - see
## assumption 4 above.
fig18_tpm_baseline <- fig18_mean_tpm %>%
  filter(cell_size_group == "small") %>%
  transmute(gene_id, strain_group, log2_tpm_baseline = log2(mean_tpmsc + 1))

## Transcription rate: mean of Miller and Neymotin synthesis-rate
## estimates where both are available - see assumption 5 above.
synth_rate_mean <- occupancy_raw %>%
  select(systematic_name, miller_synth_rate, neymotin_synth_rate) %>%
  rename(gene_id = systematic_name) %>%
  rowwise() %>%
  mutate(mean_synth_rate = mean(c(miller_synth_rate, neymotin_synth_rate),
                                na.rm = TRUE)) %>%
  ungroup() %>%
  filter(is.finite(mean_synth_rate)) %>%
  select(gene_id, mean_synth_rate)

fig18_data <- fig18_sizefc %>%
  left_join(occ_efgh_pc %>% select(gene_id, log2_occupancy_efgh),
            by = "gene_id") %>%
  left_join(fig18_tpm_baseline, by = c("gene_id", "strain_group")) %>%
  left_join(synth_rate_mean, by = "gene_id") %>%
  filter(!is.na(log2_occupancy_efgh) | !is.na(log2_tpm_baseline) |
         !is.na(mean_synth_rate))

cat("Figure 18 joined dataset, genes per strain (at least one predictor present):\n")
print(table(fig18_data$strain_group))
```

```{r figure18-plot-function}
## Generic panel: predictor (x) vs the size effect log2FC (large/small,
## y), with a 30-bin running mean overlay and Spearman R in the
## subtitle - same visual convention used throughout the GBA/eIF4G
## figures (scatter_binned_gba) and the occupancy-vs-PC scatter above
## (plot_occupancy_scatter), applied here to a new set of x-variables.

plot_predictor_vs_sizefc <- function(df, predictor_col, predictor_label,
                                     point_col, tag = NULL) {
  df_plot <- df %>%
    filter(!is.na(.data[[predictor_col]]), !is.na(log2fc_size),
           is.finite(.data[[predictor_col]]), is.finite(log2fc_size))

  r_val <- cor(df_plot[[predictor_col]], df_plot$log2fc_size,
              method = "spearman", use = "complete.obs")

  # Axis trimming for display only, via coord_cartesian (not
  # scale_*_continuous limits) - a handful of genes with near-zero
  # TPM/synthesis-rate denominators produce extreme log2FC and extreme
  # predictor values (e.g. WT panel B/C ran to +/-60 on the y-axis and
  # past 400 on the synthesis-rate axis), crushing the bulk of the data
  # into an unreadable strip near zero. coord_cartesian only changes
  # what's drawn - r_val above and the binned means below are still
  # computed on the FULL df_plot, so neither the reported R nor the
  # binned mean line is affected by this trim.
  x_lim <- quantile(df_plot[[predictor_col]], c(0.01, 0.99), na.rm = TRUE)
  y_lim <- quantile(df_plot$log2fc_size, c(0.01, 0.99), na.rm = TRUE)

  df_binned <- df_plot %>%
    mutate(pred_bin = ntile(.data[[predictor_col]], 30)) %>%
    group_by(pred_bin) %>%
    summarise(
      pred_mean = mean(.data[[predictor_col]]),
      fc_mean   = mean(log2fc_size),
      .groups = "drop")

  p <- ggplot(df_plot, aes(x = .data[[predictor_col]], y = log2fc_size)) +
    geom_point(color = point_col, alpha = 0.2, size = 0.8) +
    geom_line(data = df_binned, aes(x = pred_mean, y = fc_mean),
              color = "black", linewidth = 0.8) +
    coord_cartesian(xlim = x_lim, ylim = y_lim) +
    labs(
      subtitle = paste0("Spearman R = ", round(r_val, 3),
                        "  |  n = ", nrow(df_plot), " genes"),
      x = predictor_label,
      y = "log2 FC (large/small), PhIAA-treated"
    ) +
    theme_bw()

  if (!is.null(tag)) p <- p + labs(tag = tag)
  p
}
```

```{r figure18-build, fig.width=16, fig.height=5.5}
## Two 3-panel figures: WT+PhIAA and Bcy1-AID+PhIAA, each showing the
## size effect against occupancy (A), TPM baseline (B), and
## transcription rate (C). Comparing panel A between the two strains is
## the direct test against promoter saturation: if occupancy predicts
## the size effect in WT but that relationship breaks down under Bcy1
## depletion, that argues the size effect depends on active signalling
## rather than being fully explained by static occupancy alone.

build_figure18_panel_set <- function(strain_label, df_strain) {
  p_occ <- plot_predictor_vs_sizefc(
    df_strain, "log2_occupancy_efgh",
    "log2(mean EFG/H occupancy + 1)",
    project_colours[["blue"]], tag = "A")

  p_tpm <- plot_predictor_vs_sizefc(
    df_strain, "log2_tpm_baseline",
    "log2(TPMsc + 1), small cells (baseline)",
    project_colours[["purple"]], tag = "B")

  p_synth <- plot_predictor_vs_sizefc(
    df_strain, "mean_synth_rate",
    "Mean synthesis rate (Miller/Neymotin)",
    project_colours[["red"]], tag = "C")

  (p_occ | p_tpm | p_synth) +
    plot_annotation(
      title = paste0(strain_label, " + PhIAA: predictors of the size-dependent expression change"),
      theme = theme(plot.title = element_text(face = "bold", size = 13))
    )
}

fig18_wt   <- fig18_data %>% filter(strain_group == "WT")
fig18_bcy1 <- fig18_data %>% filter(strain_group == "Bcy1-AID")

figure18_wt_panel   <- build_figure18_panel_set("WT", fig18_wt)
figure18_bcy1_panel <- build_figure18_panel_set("Bcy1-AID", fig18_bcy1)

print(figure18_wt_panel)
print(figure18_bcy1_panel)
```

```{r figure18-save}
ggsave(file.path(figures_dir, "figure18_WT_PhIAA_size_fc_predictors.png"),
       plot = figure18_wt_panel, width = 16, height = 5.5, dpi = 300)
ggsave(file.path(figures_dir, "figure18_Bcy1AID_PhIAA_size_fc_predictors.png"),
       plot = figure18_bcy1_panel, width = 16, height = 5.5, dpi = 300)

write_tsv(fig18_data, file.path(output_dir, "figure18_size_fc_predictors_joined.txt"))

cat("Saved Figure 18 (WT+PhIAA and Bcy1-AID+PhIAA panels)\n")
```

```{r save-outputs}
write_tsv(occ_efgh_pc, file.path(output_dir, "rnapii_occupancy_efgh_pc_joined.txt"))
write_tsv(efgh_results, file.path(output_dir, "rnapii_occupancy_efgh_correlation.txt"))
write_tsv(synth_rate_results, file.path(output_dir, "rnapii_occupancy_synth_rate_validation.txt"))
write_tsv(first_tp_correlation, file.path(output_dir, "rnapii_occupancy_first_timepoint_validation.txt"))
write_tsv(bcy1_occ, file.path(output_dir, "bcy1_phiaa_occupancy_joined.txt"))
write_tsv(bcy1_correlation, file.path(output_dir, "bcy1_phiaa_occupancy_correlation.txt"))

ggsave(file.path(figures_dir, "figure_rnapii_occupancy_efgh_scatter.png"),
       plot = fig_efgh, width = 15, height = 5.5, dpi = 300)
ggsave(file.path(figures_dir, "figure_bcy1_phiaa_occupancy_log2fc.png"),
       plot = p_bcy1, width = 11, height = 5, dpi = 300)
ggsave(file.path(figures_dir, "figure_bcy1_phiaa_occupancy_absolute.png"),
       plot = p_bcy1_absolute, width = 11, height = 5, dpi = 300)

cat("Done - results saved to output_tables/ and figures/\n")
```
