---
title: "RNAPII Occupancy Follow-up: EFG/H Only, Transcription Rate Validation, Bcy1-AID+PhIAA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this file's location
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Follow-up to scripts 11/12:

1. A/B/C vs EFG/H/J is a nutrient-condition difference. Only EFG and H
   match the single-cell data conditions, so only those two are used.
2. EFG and H are cycling (not arrested) cells, so cell-cycle genes are
   filtered out before correlating.
3. Additional validation: correlate occupancy against published
   transcription rate estimates (miller_synth_rate, neymotin_synth_rate)
   and against unnormalised scTPM at the first (smallest-cell) time point.
4. Key figure: Bcy1-AID + PhIAA data (small vs large cells) vs
   transcription rate, testing whether the occupancy-transcription
   correlation breaks down under signalling perturbation.

ASSUMPTIONS (flagged again at the relevant point in the script):
- "First time point" = condition "OD 0.1", arrest "1hr G1...",
  exp_group_3 "A", strain MS69 (WT).
- Bcy1-AID + PhIAA = strain_id MS756 / MS756c2, condition "PhIAA" vs
  "DMSO", at "1hr G1..." (small) vs "Large G..." (large) arrest.
- MS756 and MS756c2 treated as replicates and averaged.
- "Plot vs transcription rate" = occupancy (fixed) vs perturbation-
  induced change (log2FC PhIAA/DMSO), small vs large cells. An
  occupancy-vs-absolute-TPM version is also included as a check.

```{r load-occupancy-raw}
occupancy_path <- file.path(
  tables_dir,
  "210601_gene_annotation_RPB_occupancy.txt"
)

occupancy_raw <- read_tsv(occupancy_path, show_col_types = FALSE)

cat("Occupancy data:", nrow(occupancy_raw), "genes x",
    ncol(occupancy_raw), "columns\n")
```

# Part A: Occupancy vs PC1/PC2, restricted to EFG and H, cell-cycle genes excluded

```{r select-efg-h-and-cellcycle-flags}
occupancy_efgh <- occupancy_raw %>%
  select(systematic_name, gene_name,
         RPKM_mean_EFG, RPKM_mean_H, rankEFG, rankH,
         cell_cycle_high_confidence, cell_cycle_low_confidence,
         Campbell_2020_cell_cycle_phase, Spellman_1998_cell_cycle_phase,
         Lichtenberg_2004_cell_cycle_phase) %>%
  rename(gene_id = systematic_name) %>%
  distinct(gene_id, .keep_all = TRUE)  # a few genes duplicated in the
                                        # raw file (identical rows) - keeps
                                        # first occurrence

cat("Cell-cycle flag counts (high confidence):\n")
print(table(occupancy_efgh$cell_cycle_high_confidence, useNA = "ifany"))
```

```{r filter-cellcycle-genes}
# Exclude high-confidence cell-cycle genes (EFG/H are cycling cells)
occupancy_efgh_filtered <- occupancy_efgh %>%
  filter(is.na(cell_cycle_high_confidence) | cell_cycle_high_confidence == FALSE)

cat("Genes retained after excluding high-confidence cell-cycle genes:",
    nrow(occupancy_efgh_filtered), "of", nrow(occupancy_efgh), "\n")
```

```{r compute-efgh-average}
occupancy_efgh_filtered <- occupancy_efgh_filtered %>%
  rowwise() %>%
  mutate(
    n_efgh_present  = sum(!is.na(c(RPKM_mean_EFG, RPKM_mean_H))),
    mean_occupancy_efgh = mean(c(RPKM_mean_EFG, RPKM_mean_H), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(n_efgh_present >= 1) %>%
  mutate(
    log2_occupancy_efgh = log2(mean_occupancy_efgh + 1),
    rank_occupancy_efgh = rank(mean_occupancy_efgh, ties.method = "average")
  )

cat("Genes with at least one of EFG/H present (post cell-cycle filter):",
    nrow(occupancy_efgh_filtered), "\n")
```

```{r join-pc-efgh}
pca_genes <- read_tsv(file.path(tables_dir, "PCA_genes_AC.txt"),
                      show_col_types = FALSE)

pc_loadings <- pca_genes %>% select(gene_id, PC1, PC2)

occ_efgh_pc <- occupancy_efgh_filtered %>%
  inner_join(pc_loadings, by = "gene_id")

cat("Genes with EFG/H occupancy (cell-cycle-filtered) + PC1/PC2:",
    nrow(occ_efgh_pc), "\n")
```

```{r correlate-efgh}
run_correlation <- function(x, y, method = "pearson") {
  ct <- cor.test(x, y, method = method)
  tibble(r = unname(ct$estimate), p_value = ct$p.value)
}

efgh_results <- bind_rows(
  run_correlation(occ_efgh_pc$log2_occupancy_efgh, occ_efgh_pc$PC1) %>%
    mutate(metric = "log2_occupancy", pc = "PC1"),
  run_correlation(occ_efgh_pc$log2_occupancy_efgh, occ_efgh_pc$PC2) %>%
    mutate(metric = "log2_occupancy", pc = "PC2"),
  run_correlation(occ_efgh_pc$rank_occupancy_efgh, occ_efgh_pc$PC1) %>%
    mutate(metric = "rank_occupancy", pc = "PC1"),
  run_correlation(occ_efgh_pc$rank_occupancy_efgh, occ_efgh_pc$PC2) %>%
    mutate(metric = "rank_occupancy", pc = "PC2")
) %>%
  select(pc, metric, r, p_value) %>%
  mutate(n_genes = nrow(occ_efgh_pc))

cat("EFG/H occupancy (cell-cycle genes excluded) vs PC1/PC2:\n")
print(efgh_results)

cat("\nFor comparison, the original all-experiment-average result",
    "(script 11, no cell-cycle filter): PC1 log2 r = 0.22, PC2 log2 r =",
    "0.03.\n")
```

```{r plot-efgh-scatter, fig.width=15, fig.height=5.5}
# Titles shortened, common caption used, figure widened to fit side by side
plot_occupancy_scatter <- function(df, pc_col, pc_label, r_val, occ_col) {
  df_binned <- df %>%
    mutate(pc_bin = ntile(.data[[pc_col]], 30)) %>%
    group_by(pc_bin) %>%
    summarise(
      pc_mean  = mean(.data[[pc_col]]),
      occ_mean = mean(.data[[occ_col]]),
      .groups = "drop"
    )

  ggplot(df, aes(x = .data[[pc_col]], y = .data[[occ_col]])) +
    geom_point(color = project_colours[["blue"]], alpha = 0.15, size = 0.8) +
    geom_line(data = df_binned, aes(x = pc_mean, y = occ_mean),
              color = "black", linewidth = 0.8) +
    labs(
      title = paste0("vs ", pc_label),
      subtitle = paste0("Pearson r = ", round(r_val, 3),
                        " | n = ", nrow(df), " genes"),
      x = pc_label,
      y = "log2(mean EFG/H occupancy + 1)"
    ) +
    theme_bw() +
    theme(plot.title = element_text(size = 11))
}

r_pc1_efgh <- efgh_results %>% filter(metric == "log2_occupancy", pc == "PC1") %>% pull(r)
r_pc2_efgh <- efgh_results %>% filter(metric == "log2_occupancy", pc == "PC2") %>% pull(r)

p_efgh_pc1 <- plot_occupancy_scatter(occ_efgh_pc, "PC1", "PC1 gene loading",
                                     r_pc1_efgh, "log2_occupancy_efgh")
p_efgh_pc2 <- plot_occupancy_scatter(occ_efgh_pc, "PC2", "PC2 gene loading",
                                     r_pc2_efgh, "log2_occupancy_efgh")

fig_efgh <- (p_efgh_pc1 + p_efgh_pc2) +
  plot_annotation(
    title = "RNAPII occupancy (EFG/H, cell-cycle genes excluded)",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  )

print(fig_efgh)
```

# Part B: Validate occupancy against published transcription rate annotations

```{r load-synth-rate-columns}
synth_rates <- occupancy_raw %>%
  select(systematic_name, miller_synth_rate, neymotin_synth_rate) %>%
  rename(gene_id = systematic_name)

occ_efgh_synth <- occ_efgh_pc %>%
  left_join(synth_rates, by = "gene_id")

cat("Genes with Miller synth rate:",
    sum(!is.na(occ_efgh_synth$miller_synth_rate)), "\n")
cat("Genes with Neymotin synth rate:",
    sum(!is.na(occ_efgh_synth$neymotin_synth_rate)), "\n")
```

```{r correlate-synth-rates}
synth_rate_results <- bind_rows(
  occ_efgh_synth %>%
    filter(!is.na(miller_synth_rate)) %>%
    { run_correlation(.$log2_occupancy_efgh, .$miller_synth_rate, method = "spearman") } %>%
    mutate(source = "Miller synth rate"),
  occ_efgh_synth %>%
    filter(!is.na(neymotin_synth_rate)) %>%
    { run_correlation(.$log2_occupancy_efgh, .$neymotin_synth_rate, method = "spearman") } %>%
    mutate(source = "Neymotin synth rate")
)

cat("EFG/H occupancy vs published transcription (synthesis) rate estimates",
    "(Spearman, as a validation check that occupancy tracks independently",
    "measured transcription rate):\n")
print(synth_rate_results)
```

# Part C: Validate occupancy against raw scTPM at the first time point

```{r identify-first-timepoint-sample}
# "First time point": OD-timecourse series (exp_group_3 == "A"),
# earliest point (OD 0.1, 1hr G1 arrest, WT strain MS69).
df_sc_final <- read_tsv(file.path(tables_dir, "df_sc_final.txt"),
                        show_col_types = FALSE)

first_timepoint_samples <- df_sc_final %>%
  filter(exp_group_3 == "A", condition == "OD 0.1",
         str_detect(arrest, "^1hr")) %>%
  distinct(sample, sample_short, condition, arrest, strain_id, mean_cell_volume)

cat("Sample(s) identified as the first time point:\n")
print(first_timepoint_samples)

if (nrow(first_timepoint_samples) != 1) {
  cat("\nWARNING: expected exactly 1 sample, found", nrow(first_timepoint_samples), "\n")
}
```

```{r correlate-first-timepoint-tpm}
first_tp_tpm <- df_sc_final %>%
  filter(sample %in% first_timepoint_samples$sample) %>%
  select(gene_id, TPM) %>%
  distinct()

occ_efgh_firsttp <- occ_efgh_pc %>%
  inner_join(first_tp_tpm, by = "gene_id") %>%
  filter(!is.na(TPM))

cat("Genes with both EFG/H occupancy and first-timepoint raw TPM:",
    nrow(occ_efgh_firsttp), "\n")

first_tp_correlation <- run_correlation(
  occ_efgh_firsttp$log2_occupancy_efgh, occ_efgh_firsttp$TPM,
  method = "spearman"
)

cat("\nEFG/H occupancy vs raw (unnormalised) TPM at the first",
    "time point (Spearman):\n")
print(first_tp_correlation)
```

# Part D: Bcy1-AID + PhIAA - does occupancy predict the perturbation response?

```{r identify-bcy1-samples}
# Bcy1-AID + PhIAA vs DMSO, small ("1hr G1...") vs large ("Large G...")
# arrest. MS756/MS756c2 treated as replicates and averaged.
bcy1_samples <- df_sc_final %>%
  filter(strain_id %in% c("MS756", "MS756c2"),
         condition %in% c("PhIAA", "DMSO")) %>%
  mutate(
    cell_size_group = case_when(
      str_detect(arrest, "^1hr")  ~ "small",
      str_detect(arrest, "^Large") ~ "large",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(cell_size_group))

cat("Bcy1-AID samples identified:\n")
print(bcy1_samples %>%
        distinct(sample, condition, arrest, cell_size_group, strain_id,
                 mean_cell_volume))
```

```{r compute-bcy1-log2fc}
# Mean TPMsc per gene x condition x cell_size_group, then log2FC(PhIAA/DMSO)
bcy1_mean_tpm <- bcy1_samples %>%
  group_by(gene_id, condition, cell_size_group) %>%
  summarise(mean_tpmsc = mean(TPMsc, na.rm = TRUE), .groups = "drop")

bcy1_log2fc <- bcy1_mean_tpm %>%
  pivot_wider(names_from = condition, values_from = mean_tpmsc) %>%
  filter(!is.na(PhIAA), !is.na(DMSO), PhIAA > 0, DMSO > 0) %>%
  mutate(log2fc_phiaa_dmso = log2(PhIAA / DMSO))

cat("Genes with log2FC(PhIAA/DMSO) computed, by cell size group:\n")
print(table(bcy1_log2fc$cell_size_group))
```

```{r join-bcy1-occupancy}
bcy1_occ <- bcy1_log2fc %>%
  inner_join(
    occ_efgh_pc %>% select(gene_id, log2_occupancy_efgh, PC1, PC2),
    by = "gene_id"
  )

cat("Genes with both Bcy1 log2FC and EFG/H occupancy:", nrow(bcy1_occ), "\n")
```

```{r correlate-bcy1-occupancy}
bcy1_correlation <- bcy1_occ %>%
  group_by(cell_size_group) %>%
  summarise(
    r = cor(log2_occupancy_efgh, log2fc_phiaa_dmso, method = "spearman",
            use = "complete.obs"),
    p_value = cor.test(log2_occupancy_efgh, log2fc_phiaa_dmso,
                       method = "spearman")$p.value,
    n = n(),
    .groups = "drop"
  )

cat("Correlation between static occupancy and PhIAA-induced transcriptional",
    "change (log2FC PhIAA/DMSO), small vs large cells.\n")
cat("Reporting both r and r^2: a moderate r with small r^2 (~10-12%)",
    "means occupancy has some relationship to the response but is far",
    "from sufficient to explain it.\n")
print(bcy1_correlation %>% mutate(r2_pct = round(r^2 * 100, 1)))
```

```{r plot-bcy1-occupancy, fig.width=11, fig.height=5}
# r/r^2 values pulled directly from bcy1_correlation, not hardcoded.
#
# NOTE: this panel plots occupancy vs the DRUG effect (log2FC
# PhIAA/DMSO). Figure 18 (Part E below) plots the SIZE effect
# (log2FC large/small) instead - see Part E for the primary figure.
bcy1_corr_label <- bcy1_correlation %>%
  mutate(
    r2 = r^2,
    label = paste0(
      str_to_title(cell_size_group), ": r = ", round(r, 2),
      " (r\u00b2 = ", round(r2 * 100, 1), "%)"
    )
  ) %>%
  arrange(cell_size_group) %>%
  pull(label) %>%
  paste(collapse = "  |  ")

p_bcy1 <- ggplot(bcy1_occ, aes(x = log2_occupancy_efgh, y = log2fc_phiaa_dmso)) +
  geom_point(color = project_colours[["blue"]], alpha = 0.3, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = project_colours[["grey"]]) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.7) +
  facet_wrap(~cell_size_group, labeller = labeller(
    cell_size_group = c(small = "Small cells", large = "Large cells")
  )) +
  labs(
    subtitle = paste0(
      "Static occupancy vs log2 FC (PhIAA/DMSO)  |  ", bcy1_corr_label
    ),
    x = "log2(mean RNAPII occupancy + 1)",
    y = "log2 FC (PhIAA/DMSO)"
  ) +
  theme_bw() +
  theme(
    plot.subtitle   = element_text(size = 8.5),
    strip.background = element_blank(),
    strip.text        = element_text(face = "bold", size = 11)
  )

print(p_bcy1)
```

```{r plot-bcy1-occupancy-alternative, fig.width=11, fig.height=5}
# Secondary reading: occupancy vs absolute TPM level (not log2FC)
# within PhIAA-treated cells only, small vs large.
bcy1_absolute <- bcy1_mean_tpm %>%
  filter(condition == "PhIAA") %>%
  mutate(log2_tpmsc = log2(mean_tpmsc + 1)) %>%
  inner_join(
    occ_efgh_pc %>% select(gene_id, log2_occupancy_efgh),
    by = "gene_id"
  )

bcy1_absolute_correlation <- bcy1_absolute %>%
  group_by(cell_size_group) %>%
  summarise(
    r = cor(log2_occupancy_efgh, log2_tpmsc, method = "spearman",
            use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )

cat("Alternative version: occupancy vs absolute PhIAA-treated expression",
    "level (not log2FC), small vs large cells:\n")
print(bcy1_absolute_correlation)

p_bcy1_absolute <- ggplot(bcy1_absolute, aes(x = log2_occupancy_efgh, y = log2_tpmsc)) +
  geom_point(color = project_colours[["red"]], alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.7) +
  facet_wrap(~cell_size_group, labeller = labeller(
    cell_size_group = c(small = "Small cells", large = "Large cells")
  )) +
  labs(
    subtitle = "Alternative reading: static occupancy vs PhIAA-treated log2(TPMsc), small vs large cells",
    x = "log2(mean RNAPII occupancy + 1)",
    y = "log2(TPMsc + 1), PhIAA-treated"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text        = element_text(face = "bold", size = 11)
  )

print(p_bcy1_absolute)
```

# Part E: Figure 18 -- WT+PhIAA and Bcy1-AID+PhIAA, size effect vs occupancy/TPM/transcription rate

```{r figure18-assumptions-flag}
## FIGURE 18 -- notes on sample identification:
## 1. BATCH "exp250624" confirmed as a substring of `sample` for both
##    MS69 and MS756 PhIAA rows.
## 2. ARREST STRING MAPPING is strain-specific: MS69 uses "1hr G1
##    arrest"/"4.75hr G1 arrest", MS756 uses "1hr G1 arrest"/"Large G1
##    arrest" for the same small/large pair (confirmed via
##    mean_cell_volume).
## 3. MS756c2 is absent from batch exp250624 - "Bcy1-AID" here means
##    MS756 alone, unlike Part D which pools MS756/MS756c2.
## 4. Each strain x size x PhIAA combination has exactly one replicate
##    in this batch (thinner than Part D's pooled comparison).
## 5. No DMSO needed: the y-axis is the SIZE effect (log2FC
##    large/small, within PhIAA only), not the drug effect.
##
## STILL OPEN:
## 6. TPM BASELINE: each gene's small-cell PhIAA TPMsc (log2). Swap
##    `cell_size_group == "small"` below for a different baseline.
## 7. TRANSCRIPTION RATE: Miller and Neymotin estimates averaged into
##    one `mean_synth_rate` predictor, not shown as two panels.
```

```{r figure18-diagnose-samples}
## Diagnostic: print actual sample/arrest values before filtering,
## rather than guessing at batch/arrest strings blind.

cat("All MS69/MS756/MS756c2 + PhIAA rows, regardless of batch guess:\n")
df_sc_final %>%
  filter(strain_id %in% c("MS69", "MS756", "MS756c2"),
         condition == "PhIAA") %>%
  distinct(sample, sample_short, strain_id, condition, arrest,
           mean_cell_volume) %>%
  arrange(strain_id, arrest) %>%
  print(n = 100)

cat("\nDoes any sample string actually contain 'exp250624'?\n")
df_sc_final %>%
  filter(str_detect(sample, "exp250624") |
         str_detect(sample_short, "exp250624")) %>%
  distinct(sample, sample_short) %>%
  print(n = 20)

cat("\nAll distinct arrest strings for MS69 and MS756/MS756c2 (any condition):\n")
df_sc_final %>%
  filter(strain_id %in% c("MS69", "MS756", "MS756c2")) %>%
  distinct(strain_id, arrest) %>%
  arrange(strain_id, arrest) %>%
  print(n = 50)
```

```{r figure18-identify-samples}
## WT (MS69) and Bcy1-AID (MS756) under PhIAA, batch exp250624:
## - AREST STRING MAPPING is strain-specific: MS69 uses "1hr G1
##   arrest" (small) / "4.75hr G1 arrest" (large); MS756 uses "1hr G1
##   arrest" (small) / "Large G1 arrest" (large). Confirmed via
##   mean_cell_volume.
## - MS756c2 is NOT present in exp250624 - "Bcy1-AID+PhIAA" here means
##   MS756 alone, unlike Part D which pools MS756/MS756c2.
## - Each strain x size x condition combination has exactly one sample
##   in this batch.

BATCH_STRING <- "exp250624"

fig18_samples <- df_sc_final %>%
  filter(str_detect(sample, BATCH_STRING),
         strain_id %in% c("MS69", "MS756", "MS756c2"),
         condition == "PhIAA") %>%
  mutate(
    strain_group = case_when(
      strain_id == "MS69"                  ~ "WT",
      strain_id %in% c("MS756", "MS756c2") ~ "Bcy1-AID",
      TRUE                                 ~ NA_character_),
    cell_size_group = case_when(
      str_detect(arrest, "^1hr")                                    ~ "small",
      strain_id == "MS69"  & str_detect(arrest, "^4\\.75hr")        ~ "large",
      strain_id != "MS69"  & str_detect(arrest, "^Large")           ~ "large",
      TRUE ~ NA_character_)
  ) %>%
  filter(!is.na(strain_group), !is.na(cell_size_group))

cat("Figure 18 sample identification:\n")
print(fig18_samples %>%
        distinct(sample, strain_id, strain_group, condition, arrest,
                 cell_size_group, mean_cell_volume))

cat("\nSample counts per strain x size (MS756c2 shows 0 rows - expected,",
    "not in this batch):\n")
print(fig18_samples %>% count(strain_id, cell_size_group))

if (nrow(fig18_samples) == 0) {
  stop("fig18_samples is empty - re-check BATCH_STRING and the arrest ",
       "patterns above against a fresh figure18-diagnose-samples run.")
}

for (grp in c("WT", "Bcy1-AID")) {
  sizes_present <- fig18_samples %>%
    filter(strain_group == grp) %>%
    pull(cell_size_group) %>%
    unique()
  if (!all(c("small", "large") %in% sizes_present)) {
    cat("\nWARNING:", grp, "is missing one of small/large (has:",
        paste(sizes_present, collapse = ", "),
        ") - the size FC pivot below will fail for this strain.\n")
  }
}
```

```{r figure18-compute-sizefc}
## Per-gene mean TPMsc by strain x cell size, then log2FC(large/small)
## within each strain, within the PhIAA condition.

fig18_mean_tpm <- fig18_samples %>%
  group_by(gene_id, strain_group, cell_size_group) %>%
  summarise(mean_tpmsc = mean(TPMsc, na.rm = TRUE), .groups = "drop")

fig18_wide_check <- fig18_mean_tpm %>%
  pivot_wider(names_from = cell_size_group, values_from = mean_tpmsc)

if (!all(c("small", "large") %in% colnames(fig18_wide_check))) {
  stop("fig18_mean_tpm is missing a 'small' or 'large' column after ",
       "pivoting (has: ", paste(setdiff(colnames(fig18_wide_check),
       c("gene_id", "strain_group")), collapse = ", "), "). This means ",
       "fig18_samples had no rows for that size group for at least one ",
       "strain - go back to figure18-identify-samples and fix ",
       "SMALL_ARREST_STRING/LARGE_ARREST_STRING/BATCH_STRING.")
}

fig18_sizefc <- fig18_wide_check %>%
  filter(!is.na(small), !is.na(large), small > 0, large > 0) %>%
  mutate(log2fc_size = log2(large / small))

cat("Genes with size FC (large/small, PhIAA, exp250624) computed, by strain:\n")
print(table(fig18_sizefc$strain_group))
```

```{r figure18-join-predictors}
## TPM baseline: small-cell PhIAA TPMsc (log2), per strain
fig18_tpm_baseline <- fig18_mean_tpm %>%
  filter(cell_size_group == "small") %>%
  transmute(gene_id, strain_group, log2_tpm_baseline = log2(mean_tpmsc + 1))

## Transcription rate: mean of Miller and Neymotin synthesis-rate estimates
synth_rate_mean <- occupancy_raw %>%
  select(systematic_name, miller_synth_rate, neymotin_synth_rate) %>%
  rename(gene_id = systematic_name) %>%
  rowwise() %>%
  mutate(mean_synth_rate = mean(c(miller_synth_rate, neymotin_synth_rate),
                                na.rm = TRUE)) %>%
  ungroup() %>%
  filter(is.finite(mean_synth_rate)) %>%
  select(gene_id, mean_synth_rate)

fig18_data <- fig18_sizefc %>%
  left_join(occ_efgh_pc %>% select(gene_id, log2_occupancy_efgh),
            by = "gene_id") %>%
  left_join(fig18_tpm_baseline, by = c("gene_id", "strain_group")) %>%
  left_join(synth_rate_mean, by = "gene_id") %>%
  filter(!is.na(log2_occupancy_efgh) | !is.na(log2_tpm_baseline) |
         !is.na(mean_synth_rate))

cat("Figure 18 joined dataset, genes per strain (at least one predictor present):\n")
print(table(fig18_data$strain_group))
```

```{r figure18-plot-function}
## Generic panel: predictor (x) vs size effect log2FC (large/small, y),
## with a 30-bin running mean overlay and Spearman R in the subtitle.

plot_predictor_vs_sizefc <- function(df, predictor_col, predictor_label,
                                     point_col, tag = NULL) {
  df_plot <- df %>%
    filter(!is.na(.data[[predictor_col]]), !is.na(log2fc_size),
           is.finite(.data[[predictor_col]]), is.finite(log2fc_size))

  r_val <- cor(df_plot[[predictor_col]], df_plot$log2fc_size,
              method = "spearman", use = "complete.obs")

  # Axis trimming for display only, via coord_cartesian - a few genes
  # with near-zero denominators produce extreme values that crush the
  # bulk of the data. r_val and binned means are still computed on the
  # full (untrimmed) data.
  x_lim <- quantile(df_plot[[predictor_col]], c(0.01, 0.99), na.rm = TRUE)
  y_lim <- quantile(df_plot$log2fc_size, c(0.01, 0.99), na.rm = TRUE)

  df_binned <- df_plot %>%
    mutate(pred_bin = ntile(.data[[predictor_col]], 30)) %>%
    group_by(pred_bin) %>%
    summarise(
      pred_mean = mean(.data[[predictor_col]]),
      fc_mean   = mean(log2fc_size),
      .groups = "drop")

  p <- ggplot(df_plot, aes(x = .data[[predictor_col]], y = log2fc_size)) +
    geom_point(color = point_col, alpha = 0.2, size = 0.8) +
    geom_line(data = df_binned, aes(x = pred_mean, y = fc_mean),
              color = "black", linewidth = 0.8) +
    coord_cartesian(xlim = x_lim, ylim = y_lim) +
    labs(
      subtitle = paste0("Spearman R = ", round(r_val, 3),
                        "  |  n = ", nrow(df_plot), " genes"),
      x = predictor_label,
      y = "log2 FC (large/small), PhIAA-treated"
    ) +
    theme_bw()

  if (!is.null(tag)) p <- p + labs(tag = tag)
  p
}
```

```{r figure18-build, fig.width=16, fig.height=5.5}
## Two 3-panel figures: WT+PhIAA and Bcy1-AID+PhIAA, each showing the
## size effect against occupancy (A), TPM baseline (B), and
## transcription rate (C). Comparing panel A between strains tests
## whether occupancy predicts the size effect equally under Bcy1
## depletion.

build_figure18_panel_set <- function(strain_label, df_strain) {
  p_occ <- plot_predictor_vs_sizefc(
    df_strain, "log2_occupancy_efgh",
    "log2(mean EFG/H occupancy + 1)",
    project_colours[["blue"]], tag = "A")

  p_tpm <- plot_predictor_vs_sizefc(
    df_strain, "log2_tpm_baseline",
    "log2(TPMsc + 1), small cells (baseline)",
    project_colours[["purple"]], tag = "B")

  p_synth <- plot_predictor_vs_sizefc(
    df_strain, "mean_synth_rate",
    "Mean synthesis rate (Miller/Neymotin)",
    project_colours[["red"]], tag = "C")

  (p_occ | p_tpm | p_synth) +
    plot_annotation(
      title = paste0(strain_label, " + PhIAA: predictors of the size-dependent expression change"),
      theme = theme(plot.title = element_text(face = "bold", size = 13))
    )
}

fig18_wt   <- fig18_data %>% filter(strain_group == "WT")
fig18_bcy1 <- fig18_data %>% filter(strain_group == "Bcy1-AID")

figure18_wt_panel   <- build_figure18_panel_set("WT", fig18_wt)
figure18_bcy1_panel <- build_figure18_panel_set("Bcy1-AID", fig18_bcy1)

print(figure18_wt_panel)
print(figure18_bcy1_panel)
```

```{r figure18-save}
ggsave(file.path(figures_dir, "figure18_WT_PhIAA_size_fc_predictors.png"),
       plot = figure18_wt_panel, width = 16, height = 5.5, dpi = 300)
ggsave(file.path(figures_dir, "figure18_Bcy1AID_PhIAA_size_fc_predictors.png"),
       plot = figure18_bcy1_panel, width = 16, height = 5.5, dpi = 300)

write_tsv(fig18_data, file.path(output_dir, "figure18_size_fc_predictors_joined.txt"))

cat("Saved Figure 18 (WT+PhIAA and Bcy1-AID+PhIAA panels)\n")
```

```{r save-outputs}
write_tsv(occ_efgh_pc, file.path(output_dir, "rnapii_occupancy_efgh_pc_joined.txt"))
write_tsv(efgh_results, file.path(output_dir, "rnapii_occupancy_efgh_correlation.txt"))
write_tsv(synth_rate_results, file.path(output_dir, "rnapii_occupancy_synth_rate_validation.txt"))
write_tsv(first_tp_correlation, file.path(output_dir, "rnapii_occupancy_first_timepoint_validation.txt"))
write_tsv(bcy1_occ, file.path(output_dir, "bcy1_phiaa_occupancy_joined.txt"))
write_tsv(bcy1_correlation, file.path(output_dir, "bcy1_phiaa_occupancy_correlation.txt"))

ggsave(file.path(figures_dir, "figure_rnapii_occupancy_efgh_scatter.png"),
       plot = fig_efgh, width = 15, height = 5.5, dpi = 300)
ggsave(file.path(figures_dir, "figure_bcy1_phiaa_occupancy_log2fc.png"),
       plot = p_bcy1, width = 11, height = 5, dpi = 300)
ggsave(file.path(figures_dir, "figure_bcy1_phiaa_occupancy_absolute.png"),
       plot = p_bcy1_absolute, width = 11, height = 5, dpi = 300)

cat("Done - results saved to output_tables/ and figures/\n")
```
