---
title: "RNAPII Occupancy Follow-up: EFG/H Only, Transcription Rate Validation, Bcy1-AID+PhIAA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this Rmd's location (Adriana_analysis/scripts/)
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Follow-up to scripts 11/12, per Matthew's feedback (Slack):

1. A/B/C vs EFG/H/J is a nutrient-condition difference, not equivalent
   experiments as originally assumed. Adriana's data matches the nutrient
   conditions of EFG and H specifically, so only those two should be
   compared against PC1/PC2.
2. EFG and H are cycling (not arrested) small cells, so cell-cycle genes
   should be filtered out before repeating the correlation.
3. Additional validation: correlate occupancy against published
   transcription rate estimates already in the annotation file
   (miller_synth_rate, neymotin_synth_rate), and against Adriana's own
   unnormalised scTPM at the first (smallest-cell) time point.
4. Key figure: Bcy1-AID + PhIAA data (small vs large cells) plotted
   against transcription rate, to show the occupancy-transcription
   correlation breaks down under direct signalling perturbation - a
   direct test against the promoter-saturation model (Vidal et al.),
   which would predict this correlation should hold even under
   perturbation if saturation were driving transcriptional reprogramming.

ASSUMPTIONS MADE BELOW THAT NEED CHECKING AGAINST WHAT MATTHEW ACTUALLY
MEANT (each flagged again at the relevant point in the script):
- "Adriana's first time point" = condition "OD 0.1", arrest "1hr G1...",
  exp_group_3 "A", strain MS69 (WT) - the earliest/smallest sample in
  what looks like a separate OD-timecourse series, distinct from the
  drug-perturbation panel (exp_group_3 "B").
- Bcy1-AID + PhIAA = strain_id MS756 / MS756c2, strain_info "AID-Bcy1",
  condition "PhIAA" vs "DMSO" (matched vehicle control), at "1hr G1..."
  (small) vs "Large G..." (large) arrest.
- MS756 and MS756c2 are treated as biological replicates of the same
  strain and averaged together.
- "Plot vs transcription rate ... to show loss of correlation" is built
  as: occupancy (fixed, from the published ChIP data) vs the
  perturbation-induced transcriptional change (log2FC PhIAA/DMSO),
  separately in small and large cells. An alternative, simpler reading
  (occupancy vs absolute TPM level, not log2FC) is also included as a
  secondary check.

```{r load-occupancy-raw}
occupancy_path <- file.path(
  tables_dir,
  "210601_gene_annotation_RPB_occupancy.txt"
)

occupancy_raw <- read_tsv(occupancy_path, show_col_types = FALSE)

cat("Occupancy data:", nrow(occupancy_raw), "genes x",
    ncol(occupancy_raw), "columns\n")
```

# Part A: Occupancy vs PC1/PC2, restricted to EFG and H, cell-cycle genes excluded

```{r select-efg-h-and-cellcycle-flags}
occupancy_efgh <- occupancy_raw %>%
  select(systematic_name, gene_name,
         RPKM_mean_EFG, RPKM_mean_H, rankEFG, rankH,
         cell_cycle_high_confidence, cell_cycle_low_confidence,
         Campbell_2020_cell_cycle_phase, Spellman_1998_cell_cycle_phase,
         Lichtenberg_2004_cell_cycle_phase) %>%
  rename(gene_id = systematic_name)

cat("Cell-cycle flag counts (high confidence):\n")
print(table(occupancy_efgh$cell_cycle_high_confidence, useNA = "ifany"))
```

```{r filter-cellcycle-genes}
# Excluding genes flagged TRUE for cell_cycle_high_confidence, per
# Matthew's instruction that EFG/H are cycling cells and cell-cycle
# genes should be removed before correlating. Using the high-confidence
# flag (not low-confidence) as the exclusion criterion - a stricter
# annotation than low-confidence, chosen to avoid removing genes that
# aren't robustly cell-cycle-regulated. If Matthew intended the broader
# low-confidence exclusion instead, swap the filter below.
occupancy_efgh_filtered <- occupancy_efgh %>%
  filter(is.na(cell_cycle_high_confidence) | cell_cycle_high_confidence == FALSE)

cat("Genes retained after excluding high-confidence cell-cycle genes:",
    nrow(occupancy_efgh_filtered), "of", nrow(occupancy_efgh), "\n")
```

```{r compute-efgh-average}
occupancy_efgh_filtered <- occupancy_efgh_filtered %>%
  rowwise() %>%
  mutate(
    n_efgh_present  = sum(!is.na(c(RPKM_mean_EFG, RPKM_mean_H))),
    mean_occupancy_efgh = mean(c(RPKM_mean_EFG, RPKM_mean_H), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(n_efgh_present >= 1) %>%
  mutate(
    log2_occupancy_efgh = log2(mean_occupancy_efgh + 1),
    rank_occupancy_efgh = rank(mean_occupancy_efgh, ties.method = "average")
  )

cat("Genes with at least one of EFG/H present (post cell-cycle filter):",
    nrow(occupancy_efgh_filtered), "\n")
```

```{r join-pc-efgh}
pca_genes <- read_tsv(file.path(tables_dir, "PCA_genes_AC.txt"),
                      show_col_types = FALSE)

pc_loadings <- pca_genes %>% select(gene_id, PC1, PC2)

occ_efgh_pc <- occupancy_efgh_filtered %>%
  inner_join(pc_loadings, by = "gene_id")

cat("Genes with EFG/H occupancy (cell-cycle-filtered) + PC1/PC2:",
    nrow(occ_efgh_pc), "\n")
```

```{r correlate-efgh}
run_correlation <- function(x, y, method = "pearson") {
  ct <- cor.test(x, y, method = method)
  tibble(r = unname(ct$estimate), p_value = ct$p.value)
}

efgh_results <- bind_rows(
  run_correlation(occ_efgh_pc$log2_occupancy_efgh, occ_efgh_pc$PC1) %>%
    mutate(metric = "log2_occupancy", pc = "PC1"),
  run_correlation(occ_efgh_pc$log2_occupancy_efgh, occ_efgh_pc$PC2) %>%
    mutate(metric = "log2_occupancy", pc = "PC2"),
  run_correlation(occ_efgh_pc$rank_occupancy_efgh, occ_efgh_pc$PC1) %>%
    mutate(metric = "rank_occupancy", pc = "PC1"),
  run_correlation(occ_efgh_pc$rank_occupancy_efgh, occ_efgh_pc$PC2) %>%
    mutate(metric = "rank_occupancy", pc = "PC2")
) %>%
  select(pc, metric, r, p_value) %>%
  mutate(n_genes = nrow(occ_efgh_pc))

cat("EFG/H occupancy (cell-cycle genes excluded) vs PC1/PC2:\n")
print(efgh_results)

cat("\nFor comparison, the original all-experiment-average result",
    "(script 11, no cell-cycle filter) was approximately: PC1 log2 r =",
    "0.22, PC2 log2 r = 0.03. Compare against efgh_results above to see",
    "whether restricting to EFG/H and removing cell-cycle genes",
    "strengthens, weakens, or leaves unchanged the PC1 relationship.\n")
```

```{r plot-efgh-scatter, fig.width=15, fig.height=5.5}
# Titles shortened (moved the "RNAPII occupancy (EFG/H, cell-cycle
# genes excluded)" portion to a shared caption instead of repeating it
# in both panel titles) and the combined figure widened, since the two
# original panel titles were long enough to overlap when placed side
# by side at the previous width.
plot_occupancy_scatter <- function(df, pc_col, pc_label, r_val, occ_col) {
  df_binned <- df %>%
    mutate(pc_bin = ntile(.data[[pc_col]], 30)) %>%
    group_by(pc_bin) %>%
    summarise(
      pc_mean  = mean(.data[[pc_col]]),
      occ_mean = mean(.data[[occ_col]]),
      .groups = "drop"
    )

  ggplot(df, aes(x = .data[[pc_col]], y = .data[[occ_col]])) +
    geom_point(color = project_colours[["blue"]], alpha = 0.15, size = 0.8) +
    geom_line(data = df_binned, aes(x = pc_mean, y = occ_mean),
              color = "black", linewidth = 0.8) +
    labs(
      title = paste0("vs ", pc_label),
      subtitle = paste0("Pearson r = ", round(r_val, 3),
                        " | n = ", nrow(df), " genes"),
      x = pc_label,
      y = "log2(mean EFG/H occupancy + 1)"
    ) +
    theme_bw() +
    theme(plot.title = element_text(size = 11))
}

r_pc1_efgh <- efgh_results %>% filter(metric == "log2_occupancy", pc == "PC1") %>% pull(r)
r_pc2_efgh <- efgh_results %>% filter(metric == "log2_occupancy", pc == "PC2") %>% pull(r)

p_efgh_pc1 <- plot_occupancy_scatter(occ_efgh_pc, "PC1", "PC1 gene loading",
                                     r_pc1_efgh, "log2_occupancy_efgh")
p_efgh_pc2 <- plot_occupancy_scatter(occ_efgh_pc, "PC2", "PC2 gene loading",
                                     r_pc2_efgh, "log2_occupancy_efgh")

fig_efgh <- (p_efgh_pc1 + p_efgh_pc2) +
  plot_annotation(
    title = "RNAPII occupancy (EFG/H, cell-cycle genes excluded)",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  )

print(fig_efgh)
```

# Part B: Validate occupancy against published transcription rate annotations

```{r load-synth-rate-columns}
synth_rates <- occupancy_raw %>%
  select(systematic_name, miller_synth_rate, neymotin_synth_rate) %>%
  rename(gene_id = systematic_name)

occ_efgh_synth <- occ_efgh_pc %>%
  left_join(synth_rates, by = "gene_id")

cat("Genes with Miller synth rate:",
    sum(!is.na(occ_efgh_synth$miller_synth_rate)), "\n")
cat("Genes with Neymotin synth rate:",
    sum(!is.na(occ_efgh_synth$neymotin_synth_rate)), "\n")
```

```{r correlate-synth-rates}
synth_rate_results <- bind_rows(
  occ_efgh_synth %>%
    filter(!is.na(miller_synth_rate)) %>%
    { run_correlation(.$log2_occupancy_efgh, .$miller_synth_rate, method = "spearman") } %>%
    mutate(source = "Miller synth rate"),
  occ_efgh_synth %>%
    filter(!is.na(neymotin_synth_rate)) %>%
    { run_correlation(.$log2_occupancy_efgh, .$neymotin_synth_rate, method = "spearman") } %>%
    mutate(source = "Neymotin synth rate")
)

cat("EFG/H occupancy vs published transcription (synthesis) rate estimates",
    "(Spearman, as a validation check that occupancy tracks independently",
    "measured transcription rate):\n")
print(synth_rate_results)
```

# Part C: Validate occupancy against Adriana's raw scTPM at the first time point

```{r identify-first-timepoint-sample}
# "Adriana's first time point": interpreted as the OD-timecourse series
# (exp_group_3 == "A", distinct from the drug-perturbation panel where
# exp_group_3 == "B"), earliest point (OD 0.1, 1hr G1 arrest, WT strain
# MS69). CHECK: confirm this is the correct sample before trusting the
# result below - if wrong, adjust the filter conditions.
df_sc_final <- read_tsv(file.path(tables_dir, "df_sc_final.txt"),
                        show_col_types = FALSE)

first_timepoint_samples <- df_sc_final %>%
  filter(exp_group_3 == "A", condition == "OD 0.1",
         str_detect(arrest, "^1hr")) %>%
  distinct(sample, sample_short, condition, arrest, strain_id, mean_cell_volume)

cat("Sample(s) identified as 'Adriana's first time point':\n")
print(first_timepoint_samples)

if (nrow(first_timepoint_samples) != 1) {
  cat("\nWARNING: expected exactly 1 sample, found", nrow(first_timepoint_samples),
      "- check the filter conditions above before trusting Part C results.\n")
}
```

```{r correlate-first-timepoint-tpm}
first_tp_tpm <- df_sc_final %>%
  filter(sample %in% first_timepoint_samples$sample) %>%
  select(gene_id, TPM) %>%
  distinct()

occ_efgh_firsttp <- occ_efgh_pc %>%
  inner_join(first_tp_tpm, by = "gene_id") %>%
  filter(!is.na(TPM))

cat("Genes with both EFG/H occupancy and first-timepoint raw TPM:",
    nrow(occ_efgh_firsttp), "\n")

first_tp_correlation <- run_correlation(
  occ_efgh_firsttp$log2_occupancy_efgh, occ_efgh_firsttp$TPM,
  method = "spearman"
)

cat("\nEFG/H occupancy vs Adriana's raw (unnormalised) TPM at the first",
    "time point (Spearman):\n")
print(first_tp_correlation)
```

# Part D: Bcy1-AID + PhIAA - does occupancy predict the perturbation response?

```{r identify-bcy1-samples}
# Bcy1-AID + PhIAA vs DMSO, small ("1hr G1...") vs large ("Large G...")
# arrest. MS756 and MS756c2 treated as biological replicates of the same
# AID-Bcy1 strain and averaged. CHECK: confirm arrest string matching
# below actually captures the intended small/large samples (arrest
# column values are truncated in the console output used to identify
# them, so str_detect patterns are a best guess at the full strings).
bcy1_samples <- df_sc_final %>%
  filter(strain_id %in% c("MS756", "MS756c2"),
         condition %in% c("PhIAA", "DMSO")) %>%
  mutate(
    cell_size_group = case_when(
      str_detect(arrest, "^1hr")  ~ "small",
      str_detect(arrest, "^Large") ~ "large",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(cell_size_group))

cat("Bcy1-AID samples identified:\n")
print(bcy1_samples %>%
        distinct(sample, condition, arrest, cell_size_group, strain_id,
                 mean_cell_volume))
```

```{r compute-bcy1-log2fc}
# Mean TPMsc per gene x condition x cell_size_group (averaging replicate
# samples, including across MS756/MS756c2), then log2FC(PhIAA/DMSO)
# within each cell size group separately.
bcy1_mean_tpm <- bcy1_samples %>%
  group_by(gene_id, condition, cell_size_group) %>%
  summarise(mean_tpmsc = mean(TPMsc, na.rm = TRUE), .groups = "drop")

bcy1_log2fc <- bcy1_mean_tpm %>%
  pivot_wider(names_from = condition, values_from = mean_tpmsc) %>%
  filter(!is.na(PhIAA), !is.na(DMSO), PhIAA > 0, DMSO > 0) %>%
  mutate(log2fc_phiaa_dmso = log2(PhIAA / DMSO))

cat("Genes with log2FC(PhIAA/DMSO) computed, by cell size group:\n")
print(table(bcy1_log2fc$cell_size_group))
```

```{r join-bcy1-occupancy}
bcy1_occ <- bcy1_log2fc %>%
  inner_join(
    occ_efgh_pc %>% select(gene_id, log2_occupancy_efgh, PC1, PC2),
    by = "gene_id"
  )

cat("Genes with both Bcy1 log2FC and EFG/H occupancy:", nrow(bcy1_occ), "\n")
```

```{r correlate-bcy1-occupancy}
bcy1_correlation <- bcy1_occ %>%
  group_by(cell_size_group) %>%
  summarise(
    r = cor(log2_occupancy_efgh, log2fc_phiaa_dmso, method = "spearman",
            use = "complete.obs"),
    p_value = cor.test(log2_occupancy_efgh, log2fc_phiaa_dmso,
                       method = "spearman")$p.value,
    n = n(),
    .groups = "drop"
  )

cat("Correlation between static occupancy and PhIAA-induced transcriptional",
    "change (log2FC PhIAA/DMSO), separately in small and large cells.\n")
cat("Under a strict promoter-saturation model, occupancy alone should",
    "largely determine this response. Report both r and r^2 below:",
    "r describes the strength/direction of any relationship, r^2",
    "describes how much of the response variance occupancy actually",
    "explains - a moderate r with a small r^2 (as found here, ~10-12%)",
    "indicates occupancy has *some* relationship to the response but is",
    "far from sufficient to explain it, which is the precise claim to",
    "report rather than 'no correlation':\n")
print(bcy1_correlation %>% mutate(r2_pct = round(r^2 * 100, 1)))
```

```{r plot-bcy1-occupancy, fig.width=11, fig.height=5}
# Title/subtitle corrected to match the actual result: bcy1_correlation
# (computed in the chunk above) shows r ~ 0.31-0.35 in both cell size
# groups, p < 1e-100, n = 4392 - a moderate, highly significant
# correlation, NOT "no correlation" as an earlier version of this figure
# claimed. r^2 (~10-12%) is used in the subtitle since that is the more
# interpretable quantity here: it states plainly how much of the
# transcriptional response variance is (and, more importantly, is NOT)
# explained by static occupancy alone. Values are pulled directly from
# bcy1_correlation rather than hardcoded, so the figure stays accurate
# if the underlying data/filtering changes on a re-run.
bcy1_corr_label <- bcy1_correlation %>%
  mutate(
    r2 = r^2,
    label = paste0(
      str_to_title(cell_size_group), ": r = ", round(r, 2),
      " (r\u00b2 = ", round(r2 * 100, 1), "%)"
    )
  ) %>%
  arrange(cell_size_group) %>%
  pull(label) %>%
  paste(collapse = "  |  ")

p_bcy1 <- ggplot(bcy1_occ, aes(x = log2_occupancy_efgh, y = log2fc_phiaa_dmso)) +
  geom_point(color = project_colours[["blue"]], alpha = 0.3, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = project_colours[["grey"]]) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.7) +
  facet_wrap(~cell_size_group, labeller = labeller(
    cell_size_group = c(small = "Small cells", large = "Large cells")
  )) +
  labs(
    title = "RNAPII occupancy explains only a small fraction of the Bcy1-AID + PhIAA transcriptional response",
    subtitle = paste0(
      "Static occupancy (EFG/H) vs log2FC(PhIAA/DMSO)  |  ", bcy1_corr_label,
      "  |  most of the transcriptional response is not explained by baseline occupancy"
    ),
    x = "log2(mean EFG/H occupancy + 1)",
    y = "log2FC (PhIAA / DMSO)"
  ) +
  theme_bw() +
  theme(plot.subtitle = element_text(size = 8.5))

print(p_bcy1)
```

```{r plot-bcy1-occupancy-alternative, fig.width=11, fig.height=5}
# Secondary/alternative reading: occupancy vs absolute TPM level (not
# log2FC) within PhIAA-treated cells only, small vs large - a simpler
# version of "plot occupancy vs transcription rate in small and large
# cells" in case this, rather than the log2FC version above, is what
# Matthew meant.
bcy1_absolute <- bcy1_mean_tpm %>%
  filter(condition == "PhIAA") %>%
  mutate(log2_tpmsc = log2(mean_tpmsc + 1)) %>%
  inner_join(
    occ_efgh_pc %>% select(gene_id, log2_occupancy_efgh),
    by = "gene_id"
  )

bcy1_absolute_correlation <- bcy1_absolute %>%
  group_by(cell_size_group) %>%
  summarise(
    r = cor(log2_occupancy_efgh, log2_tpmsc, method = "spearman",
            use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )

cat("Alternative version: occupancy vs absolute PhIAA-treated expression",
    "level (not log2FC), small vs large cells:\n")
print(bcy1_absolute_correlation)

p_bcy1_absolute <- ggplot(bcy1_absolute, aes(x = log2_occupancy_efgh, y = log2_tpmsc)) +
  geom_point(color = project_colours[["red"]], alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.7) +
  facet_wrap(~cell_size_group, labeller = labeller(
    cell_size_group = c(small = "Small cells", large = "Large cells")
  )) +
  labs(
    title = "RNAPII occupancy vs absolute expression under PhIAA treatment",
    subtitle = "Alternative reading: static occupancy vs PhIAA-treated log2(TPMsc), small vs large cells",
    x = "log2(mean EFG/H occupancy + 1)",
    y = "log2(TPMsc + 1), PhIAA-treated"
  ) +
  theme_bw()

print(p_bcy1_absolute)
```

```{r save-outputs}
write_tsv(occ_efgh_pc, file.path(output_dir, "rnapii_occupancy_efgh_pc_joined.txt"))
write_tsv(efgh_results, file.path(output_dir, "rnapii_occupancy_efgh_correlation.txt"))
write_tsv(synth_rate_results, file.path(output_dir, "rnapii_occupancy_synth_rate_validation.txt"))
write_tsv(first_tp_correlation, file.path(output_dir, "rnapii_occupancy_first_timepoint_validation.txt"))
write_tsv(bcy1_occ, file.path(output_dir, "bcy1_phiaa_occupancy_joined.txt"))
write_tsv(bcy1_correlation, file.path(output_dir, "bcy1_phiaa_occupancy_correlation.txt"))

ggsave(file.path(figures_dir, "figure_rnapii_occupancy_efgh_scatter.png"),
       plot = fig_efgh, width = 15, height = 5.5, dpi = 300)
ggsave(file.path(figures_dir, "figure_bcy1_phiaa_occupancy_log2fc.png"),
       plot = p_bcy1, width = 11, height = 5, dpi = 300)
ggsave(file.path(figures_dir, "figure_bcy1_phiaa_occupancy_absolute.png"),
       plot = p_bcy1_absolute, width = 11, height = 5, dpi = 300)

cat("Done - results saved to output_tables/ and figures/\n")
```
