---
title: "Figure 16 -- RNAPII Occupancy (EFG/H average), TPM, and Transcription Rate vs PC1/PC2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this Rmd's location (Adriana_analysis/scripts/)
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Simplified rebuild of the original 13a occupancy-vs-PC1/PC2 grid.

Per the latest revision request, conditions A/B/C are dropped entirely
(previously kept as individual rows in an earlier version of this
script) - they are a different nutrient condition from EFG/H and were
never meant to be pooled with it, and are not needed as a separate
comparison here. The grid is now just three rows:

1. EFG/H AVERAGE (cell-cycle genes excluded): reuses the same
   `mean_occupancy_efgh` logic already built for script 14/Figure 18 -
   average of EFG+H RPKM, with high-confidence cell-cycle genes
   excluded (Matthew's stated reason: EFG/H are cycling, not arrested,
   cells).
2. TPM: script 14's "first timepoint" reference sample (condition "OD
   0.1", arrest "1hr G1...", exp_group_3 "A", strain MS69) - the same
   general-purpose TPM reference occupancy was already validated
   against in script 14 Part C, NOT Figure 18's PhIAA/strain-specific
   baseline. FLAG if a different TPM source is intended here.
3. TRANSCRIPTION RATE: mean of Miller and Neymotin synthesis-rate
   estimates, same definition as script 14/Figure 18.

Metric: Pearson r on log2-transformed predictor, matching the original
script's convention. `rank_occupancy` was dropped in an earlier
revision (highly correlated with the log2 metric, so redundant) and
stays dropped here.

```{r load-occupancy-raw}
occupancy_path <- file.path(
  tables_dir,
  "210601_gene_annotation_RPB_occupancy.txt"
)

occupancy_raw <- read_tsv(occupancy_path, show_col_types = FALSE)

cat("Occupancy data:", nrow(occupancy_raw), "genes x",
    ncol(occupancy_raw), "columns\n")
```

```{r load-pc-loadings}
pca_genes <- read_tsv(file.path(tables_dir, "PCA_genes_AC.txt"),
                      show_col_types = FALSE)

pc_loadings <- pca_genes %>%
  select(gene_id, PC1, PC2)

cat("PC1/PC2 loadings available for", nrow(pc_loadings), "genes\n")
```

```{r run-correlation-function}
run_correlation <- function(x, y, method = "pearson") {
  ct <- cor.test(x, y, method = method)
  tibble(r = unname(ct$estimate), p_value = ct$p.value)
}
```

# Part 1: EFG/H average (cell-cycle genes excluded)

```{r compute-efgh-average}
# Same logic as script 14's Part A: average EFG+H, exclude high-
# confidence cell-cycle genes (EFG/H are cycling, not arrested, cells).
occupancy_efgh <- occupancy_raw %>%
  select(systematic_name, gene_name,
         RPKM_mean_EFG, RPKM_mean_H, cell_cycle_high_confidence) %>%
  rename(gene_id = systematic_name) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  filter(is.na(cell_cycle_high_confidence) | cell_cycle_high_confidence == FALSE)

occupancy_efgh <- occupancy_efgh %>%
  rowwise() %>%
  mutate(
    n_efgh_present      = sum(!is.na(c(RPKM_mean_EFG, RPKM_mean_H))),
    mean_occupancy_efgh = mean(c(RPKM_mean_EFG, RPKM_mean_H), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(n_efgh_present >= 1) %>%
  mutate(log2_occupancy_efgh = log2(mean_occupancy_efgh + 1))

cat("Genes with EFG/H occupancy (cell-cycle-filtered):",
    nrow(occupancy_efgh), "\n")
```

```{r correlate-efgh}
efgh_pc <- occupancy_efgh %>%
  inner_join(pc_loadings, by = "gene_id")

efgh_results <- bind_rows(
  run_correlation(efgh_pc$log2_occupancy_efgh, efgh_pc$PC1) %>% mutate(pc = "PC1"),
  run_correlation(efgh_pc$log2_occupancy_efgh, efgh_pc$PC2) %>% mutate(pc = "PC2")
) %>%
  mutate(predictor = "EFG/H average", n_genes = nrow(efgh_pc))

cat("EFG/H average (cell-cycle excluded) - correlation with PC1/PC2:\n")
print(efgh_results)
```

# Part 2: TPM and transcription rate

```{r load-tpm-reference}
## TPM: reusing script 14's "first timepoint" reference sample (condition
## "OD 0.1", arrest "1hr G1...", exp_group_3 "A", strain MS69) - the same
## general-purpose TPM reference occupancy was already validated against
## in script 14 Part C, NOT Figure 18's PhIAA/strain-specific baseline.
## FLAG if a different TPM source is intended for Figure 16 specifically.
df_sc_final <- read_tsv(file.path(tables_dir, "df_sc_final.txt"),
                        show_col_types = FALSE)

first_timepoint_samples <- df_sc_final %>%
  filter(exp_group_3 == "A", condition == "OD 0.1",
         str_detect(arrest, "^1hr")) %>%
  distinct(sample)

tpm_reference <- df_sc_final %>%
  filter(sample %in% first_timepoint_samples$sample) %>%
  select(gene_id, TPM) %>%
  distinct() %>%
  filter(!is.na(TPM), TPM > 0) %>%
  mutate(log2_tpm = log2(TPM + 1))

cat("Genes with TPM reference value:", nrow(tpm_reference), "\n")
```

```{r load-synth-rate}
## Transcription rate: mean of Miller and Neymotin synthesis-rate
## estimates, same definition as script 14/Figure 18.
synth_rate_mean <- occupancy_raw %>%
  select(systematic_name, miller_synth_rate, neymotin_synth_rate) %>%
  rename(gene_id = systematic_name) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  rowwise() %>%
  mutate(mean_synth_rate = mean(c(miller_synth_rate, neymotin_synth_rate),
                                na.rm = TRUE)) %>%
  ungroup() %>%
  filter(is.finite(mean_synth_rate))

cat("Genes with mean synthesis rate:", nrow(synth_rate_mean), "\n")
```

```{r correlate-tpm-synth}
tpm_pc <- tpm_reference %>% inner_join(pc_loadings, by = "gene_id")
synth_pc <- synth_rate_mean %>% inner_join(pc_loadings, by = "gene_id")

tpm_results <- bind_rows(
  run_correlation(tpm_pc$log2_tpm, tpm_pc$PC1, method = "pearson") %>% mutate(pc = "PC1"),
  run_correlation(tpm_pc$log2_tpm, tpm_pc$PC2, method = "pearson") %>% mutate(pc = "PC2")
) %>%
  mutate(predictor = "TPM (first timepoint)", n_genes = nrow(tpm_pc))

synth_results <- bind_rows(
  run_correlation(synth_pc$mean_synth_rate, synth_pc$PC1, method = "pearson") %>% mutate(pc = "PC1"),
  run_correlation(synth_pc$mean_synth_rate, synth_pc$PC2, method = "pearson") %>% mutate(pc = "PC2")
) %>%
  mutate(predictor = "Transcription rate (Miller/Neymotin)", n_genes = nrow(synth_pc))

cat("TPM - correlation with PC1/PC2:\n")
print(tpm_results)
cat("\nTranscription rate - correlation with PC1/PC2:\n")
print(synth_results)
```

# Combined grid (EFG/H average, TPM, transcription rate)

```{r build-simplified-grid, fig.width=10, fig.height=6}
# A/B/C rows removed - grid_results is now only 3 predictors x 2 PCs
# = 6 cells, down from 6 predictors x 2 PCs = 12.
grid_results <- bind_rows(
  efgh_results,
  tpm_results,
  synth_results
) %>%
  mutate(
    predictor = factor(predictor,
                       levels = c("EFG/H average",
                                 "TPM (first timepoint)",
                                 "Transcription rate (Miller/Neymotin)")),
    p_adjust  = p.adjust(p_value, method = "BH"),
    sig_label = if_else(p_adjust < 0.05, "*", "")
  ) %>%
  arrange(pc, predictor)

cat("Figure 16 grid (log2 metric, Pearson):\n")
print(grid_results, n = Inf)

# Label sizes increased across the board - axis text/title, tile
# labels, subtitle - previous version was noticeably small on a normal
# screen without zooming in.
p_grid16 <- ggplot(grid_results, aes(x = pc, y = predictor, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(round(r, 2), sig_label)), size = 8) +
  scale_fill_gradient2(
    low = project_colours[["blue"]], mid = "white",
    high = project_colours[["red"]], midpoint = 0
  ) +
  labs(
    title = "RNAPII occupancy (EFG/H), TPM, and transcription rate vs PC1/PC2",
    subtitle = str_wrap(
      "Cell values: Pearson r on log2-transformed predictor (* = FDR < 0.05) | EFG/H average uses cell-cycle-excluded occupancy; TPM is the first-timepoint reference sample; transcription rate is the Miller/Neymotin mean",
      width = 70
    ),
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme_bw(base_size = 20) +
  theme(
    axis.text        = element_text(size = 18, colour = "black"),
    plot.title       = element_text(size = 18, face = "bold"),
    plot.subtitle    = element_text(size = 13),
    legend.text      = element_text(size = 15),
    legend.title     = element_text(size = 17)
  )

print(p_grid16)
```

```{r save-outputs}
write_tsv(grid_results, file.path(output_dir, "figure16_simplified_grid.txt"))

ggsave(file.path(figures_dir, "figure16_occupancy_tpm_synth_vs_pc_grid.png"),
       plot = p_grid16, width = 11, height = 7, dpi = 300)

cat("Done - Figure 16 results saved to output_tables/ and figures/\n")
```
