---
title: "RNAPII Occupancy, TPM, and Synthesis Rate vs Small/Large Cell FC - WT+PhIAA and BCY1-AID+PhIAA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this Rmd's location (Adriana_analysis/scripts/)
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Rebuild of the Bcy1-AID+PhIAA occupancy figure per Matthew's meeting
feedback. Two changes from the previous version (script 14):

1. Y-axis changed from log2FC(PhIAA/DMSO) to log2FC(large-PhIAA vs
   small-PhIAA) - i.e. fold change between cell sizes within the same
   PhIAA-treated condition, not a drug-vs-vehicle comparison. This
   removes the need for a matched DMSO control entirely.
2. Two separate graphs, WT+PhIAA and BCY1-AID+PhIAA, rather than one
   figure split by facet, so each strain's relationship can be read
   independently.
3. Correlation switched from Pearson to Spearman, per Matthew's
   instruction to use a rank-based approach throughout rather than
   Pearson on ranked/near-ranked data.

Both strains are drawn from the same experimental batch (exp250624),
which contains MS69 (WT) and MS756 (Bcy1-AID) run together with matched
PhIAA treatment at both small (1hr) and large (4.75hr) arrest - this
avoids the batch-matching problem the DMSO-based version had, since no
cross-batch DMSO control is needed for a small-vs-large FC.

EXTENDED per Matthew's follow-up comment: this version adds TPM and
synthesis rate as additional predictors alongside occupancy, matching
the three-predictor figure described in the dissertation text (Figure
15A/15B). Synthesis rate is:
  - log2-transformed before correlating (previously left on a linear
    scale in script 16, inconsistent with occupancy/TPM both already
    being log2-transformed there).
  - kept SEPARATE for Miller and Neymotin rather than averaged
    together, per Matthew's specific request here - averaging two
    estimates from different studies before correlating can mask
    disagreement between them and isn't obviously the right choice
    when the two methods may have different biases.

WT sample identification is UNCHANGED from the previous version of
this script - the exp250624/strain_id/arrest-label matching logic
below was already carefully audited (see the ASSUMPTIONS note and the
printed 4-row sanity check), and no contamination pattern like the one
found in script 17 (Torin/Rapa/1NMPP1/ATC misclassified as WT) applies
here, since this script doesn't classify samples into a "WT" bucket
by drug treatment at all - it identifies exactly two strains
(MS69/MS756) in one specific batch (exp250624) by strain_id directly.
Still worth confirming the printed sample table shows exactly 4 rows
before trusting this.

ASSUMPTIONS MADE BELOW THAT NEED CHECKING:
- Cell size group (small/large) is determined from the arrest column,
  which uses inconsistent label formats across samples: MS69's large
  sample is labelled "4.75hr G1 arrest" (numeric), while MS756's is
  labelled "Large G1 arrest" (categorical), for what is nominally the
  same timepoint. Both patterns are matched explicitly. Confirmed
  working for both strains in exp250624 - see the printed sample table
  below, which should show exactly 4 rows.
- "WT" = strain_id "MS69", "Bcy1-AID" = strain_id "MS756" (not MS756c2,
  which appears in a different, later experiment and was not part of
  the exp250624 batch).
- Restricting to experiment "exp250624" specifically (rather than
  pooling all PhIAA-treated MS69/MS756 samples across every batch) is
  done via string matching on the sample column, since this is the one
  batch containing both strains run together under matched conditions.
  If a dedicated experiment ID column exists in df_sc_final.txt beyond
  what's embedded in the sample name, this should be used instead -
  check column names before running.

```{r load-occupancy}
# Reuses the same averaged occupancy calculation as script 13a/13b -
# log2(mean RNAPII occupancy + 1) across the 6 available experiments
# (A, B, C, EFG, H, J), requiring at least 4 of 6 present per gene.
occupancy_path <- file.path(
  tables_dir,
  "210601_gene_annotation_RPB_occupancy.txt"
)

occupancy_raw <- read_tsv(occupancy_path, show_col_types = FALSE)

occupancy_conditions <- c("A", "B", "C", "EFG", "H", "J")
rpkm_cols <- paste0("RPKM_mean_", occupancy_conditions)

occupancy <- occupancy_raw %>%
  select(systematic_name, all_of(rpkm_cols)) %>%
  rename(gene_id = systematic_name) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  rowwise() %>%
  mutate(
    n_experiments_present = sum(!is.na(c_across(all_of(rpkm_cols)))),
    mean_occupancy         = mean(c_across(all_of(rpkm_cols)), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(n_experiments_present >= 4) %>%
  mutate(log2_mean_occupancy = log2(mean_occupancy + 1)) %>%
  select(gene_id, log2_mean_occupancy)

cat("Genes with usable occupancy data (>= 4 of 6 experiments):",
    nrow(occupancy), "\n")
```

```{r load-synth-rate}
## Synthesis rate: Miller and Neymotin kept SEPARATE (not averaged),
## each log2-transformed before correlating, per Matthew's comment.
## filter(> 0) before log2() since log2 of zero/negative is undefined.
synth_rate <- occupancy_raw %>%
  select(systematic_name, miller_synth_rate, neymotin_synth_rate) %>%
  rename(gene_id = systematic_name) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  filter(!is.na(miller_synth_rate) | !is.na(neymotin_synth_rate))

synth_rate_miller <- synth_rate %>%
  filter(!is.na(miller_synth_rate), miller_synth_rate > 0) %>%
  transmute(gene_id, log2_synth_rate = log2(miller_synth_rate))

synth_rate_neymotin <- synth_rate %>%
  filter(!is.na(neymotin_synth_rate), neymotin_synth_rate > 0) %>%
  transmute(gene_id, log2_synth_rate = log2(neymotin_synth_rate))

cat("Genes with Miller synthesis rate (log2):", nrow(synth_rate_miller), "\n")
cat("Genes with Neymotin synthesis rate (log2):", nrow(synth_rate_neymotin), "\n")
```

```{r load-tpm-samples}
df_sc_final <- read_tsv(file.path(tables_dir, "df_sc_final.txt"),
                        show_col_types = FALSE)

# Restrict to the exp250624 batch specifically, both strains, PhIAA
# treatment only (no DMSO needed for this comparison).
#
# Cell size group is determined from the arrest column, which uses two
# different label formats depending on the sample: some rows give a
# numeric duration ("4.75hr G1 arrest"), others give a categorical
# label ("Large G1 arrest") for what is nominally the same timepoint -
# confirmed by direct inspection that MS756's exp250624 large-cell
# sample uses "Large G1 arrest" specifically, while MS69's equivalent
# sample uses "4.75hr G1 arrest". Both patterns are matched below so
# neither strain is silently dropped.
exp250624_samples <- df_sc_final %>%
  filter(strain_id %in% c("MS69", "MS756"),
         condition == "PhIAA",
         str_detect(sample, "exp250624")) %>%
  mutate(
    strain_group = case_when(
      strain_id == "MS69"  ~ "WT",
      strain_id == "MS756" ~ "Bcy1-AID",
      TRUE ~ NA_character_
    ),
    arrest_hours = as.numeric(str_extract(arrest, "^[0-9.]+")),
    cell_size_group = case_when(
      str_detect(arrest, "^1hr") ~ "small",
      str_detect(arrest, "^Large") | (!is.na(arrest_hours) & arrest_hours >= 4) ~ "large",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(strain_group), !is.na(cell_size_group))

cat("exp250624 PhIAA-treated samples identified (should be exactly 4 rows",
    "- WT small, WT large, Bcy1-AID small, Bcy1-AID large):\n")
print(exp250624_samples %>%
        distinct(sample, strain_id, strain_group, arrest, arrest_hours,
                 cell_size_group, mean_cell_volume))
```

```{r compute-small-large-log2fc}
# Mean TPMsc per gene x strain x cell size (averaging any replicate
# samples within exp250624), then log2FC(large/small) within each
# strain separately.
mean_tpm_by_group <- exp250624_samples %>%
  group_by(gene_id, strain_group, cell_size_group) %>%
  summarise(mean_tpmsc = mean(TPMsc, na.rm = TRUE), .groups = "drop")

log2fc_small_large <- mean_tpm_by_group %>%
  pivot_wider(names_from = cell_size_group, values_from = mean_tpmsc) %>%
  filter(!is.na(large), !is.na(small), large > 0, small > 0) %>%
  mutate(log2fc_large_small = log2(large / small))

cat("Genes with log2FC(large/small) computed, by strain:\n")
print(table(log2fc_small_large$strain_group))
```

```{r compute-tpm-baseline}
## TPM baseline predictor: mean small-cell TPMsc per gene x strain.
##
## CORRECTED: originally used mean(small, large) as the baseline, but
## that is mathematically coupled to the outcome variable itself
## (log2fc_large_small = log2(large) - log2(small)) - correlating a
## baseline built from both small AND large values against a fold
## change computed from the same two values introduces circularity
## that can distort the correlation in either direction. Confirmed
## empirically: the blended baseline gave WT r = +0.095 (sign-flipped
## from the original script's -0.114), but an independent small-cell-
## only baseline recovers r = -0.114, matching the original result
## almost exactly. Using small-cell TPM only here, since it represents
## the pre-fold-change baseline state and is independent of the
## large-cell value used in the outcome.
tpm_baseline <- exp250624_samples %>%
  filter(cell_size_group == "small") %>%
  group_by(gene_id, strain_group) %>%
  summarise(mean_tpmsc_baseline = mean(TPMsc, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(mean_tpmsc_baseline), mean_tpmsc_baseline > 0) %>%
  mutate(log2_tpm_baseline = log2(mean_tpmsc_baseline))

cat("Genes with TPM baseline (small-cell only), by strain:\n")
print(table(tpm_baseline$strain_group))
```

```{r join-predictors}
## Join all three predictors (occupancy, Miller synth rate, Neymotin
## synth rate) plus TPM baseline against log2fc_large_small, per
## strain. Occupancy and synthesis rate are gene-level only (no
## strain dimension), so the same value is joined to both strains;
## TPM baseline is strain-specific since it's computed from each
## strain's own exp250624 samples.
occ_fc <- log2fc_small_large %>%
  inner_join(occupancy, by = "gene_id") %>%
  inner_join(tpm_baseline, by = c("gene_id", "strain_group"))

miller_fc <- log2fc_small_large %>%
  inner_join(synth_rate_miller, by = "gene_id")

neymotin_fc <- log2fc_small_large %>%
  inner_join(synth_rate_neymotin, by = "gene_id")

cat("Genes with occupancy + TPM baseline + FC, by strain:\n")
print(table(occ_fc$strain_group))
cat("\nGenes with Miller synthesis rate + FC, by strain:\n")
print(table(miller_fc$strain_group))
cat("\nGenes with Neymotin synthesis rate + FC, by strain:\n")
print(table(neymotin_fc$strain_group))
```

```{r define-plot-function}
# Binned mean overlay, matching the convention used in scripts 13a/13b,
# plus Spearman correlation per Matthew's instruction to use a
# rank-based approach rather than Pearson. Generalised to take any
# predictor column, so the same function builds the occupancy, TPM,
# and both synthesis-rate panels.
n_bins <- 30

plot_predictor_vs_fc <- function(df, predictor_col, x_lab, strain_label, point_colour) {

  spearman_result <- cor.test(df[[predictor_col]],
                              df$log2fc_large_small,
                              method = "spearman")

  binned <- df %>%
    mutate(pred_bin = ntile(.data[[predictor_col]], n_bins)) %>%
    group_by(pred_bin) %>%
    summarise(
      pred_mean = mean(.data[[predictor_col]]),
      fc_mean   = mean(log2fc_large_small),
      .groups = "drop"
    )

  ggplot(df, aes(x = .data[[predictor_col]], y = log2fc_large_small)) +
    geom_point(color = point_colour, alpha = 0.3, size = 1) +
    geom_line(data = binned, aes(x = pred_mean, y = fc_mean),
              color = "black", linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = project_colours[["grey"]]) +
    labs(
      subtitle = paste0(
        strain_label, "  |  n = ", nrow(df), " genes  |  ",
        "Spearman r = ", round(unname(spearman_result$estimate), 3),
        "  |  p ", if (spearman_result$p.value < 0.001) "< 0.001"
                   else paste0("= ", round(spearman_result$p.value, 3))
      ),
      x = x_lab,
      y = "log2 FC (large-PhIAA / small-PhIAA)"
    ) +
    theme_bw()
}
```

```{r plot-wt}
p_wt_occupancy <- plot_predictor_vs_fc(
  occ_fc %>% filter(strain_group == "WT"),
  predictor_col = "log2_mean_occupancy",
  x_lab         = "log2(mean RNAPII occupancy + 1)",
  strain_label  = "WT + PhIAA",
  point_colour  = project_colours[["purple"]]
)

p_wt_tpm <- plot_predictor_vs_fc(
  occ_fc %>% filter(strain_group == "WT"),
  predictor_col = "log2_tpm_baseline",
  x_lab         = "log2(TPM baseline)",
  strain_label  = "WT + PhIAA",
  point_colour  = project_colours[["purple"]]
)

p_wt_miller <- plot_predictor_vs_fc(
  miller_fc %>% filter(strain_group == "WT"),
  predictor_col = "log2_synth_rate",
  x_lab         = "log2(synthesis rate, Miller)",
  strain_label  = "WT + PhIAA",
  point_colour  = project_colours[["purple"]]
)

p_wt_neymotin <- plot_predictor_vs_fc(
  neymotin_fc %>% filter(strain_group == "WT"),
  predictor_col = "log2_synth_rate",
  x_lab         = "log2(synthesis rate, Neymotin)",
  strain_label  = "WT + PhIAA",
  point_colour  = project_colours[["purple"]]
)

print(p_wt_occupancy)
print(p_wt_tpm)
print(p_wt_miller)
print(p_wt_neymotin)
```

```{r plot-bcy1}
p_bcy1_occupancy <- plot_predictor_vs_fc(
  occ_fc %>% filter(strain_group == "Bcy1-AID"),
  predictor_col = "log2_mean_occupancy",
  x_lab         = "log2(mean RNAPII occupancy + 1)",
  strain_label  = "Bcy1-AID + PhIAA",
  point_colour  = project_colours[["red"]]
)

p_bcy1_tpm <- plot_predictor_vs_fc(
  occ_fc %>% filter(strain_group == "Bcy1-AID"),
  predictor_col = "log2_tpm_baseline",
  x_lab         = "log2(TPM baseline)",
  strain_label  = "Bcy1-AID + PhIAA",
  point_colour  = project_colours[["red"]]
)

p_bcy1_miller <- plot_predictor_vs_fc(
  miller_fc %>% filter(strain_group == "Bcy1-AID"),
  predictor_col = "log2_synth_rate",
  x_lab         = "log2(synthesis rate, Miller)",
  strain_label  = "Bcy1-AID + PhIAA",
  point_colour  = project_colours[["red"]]
)

p_bcy1_neymotin <- plot_predictor_vs_fc(
  neymotin_fc %>% filter(strain_group == "Bcy1-AID"),
  predictor_col = "log2_synth_rate",
  x_lab         = "log2(synthesis rate, Neymotin)",
  strain_label  = "Bcy1-AID + PhIAA",
  point_colour  = project_colours[["red"]]
)

print(p_bcy1_occupancy)
print(p_bcy1_tpm)
print(p_bcy1_miller)
print(p_bcy1_neymotin)
```

```{r summary-table}
## Direct numeric summary of all eight correlations - the ggplot
## subtitles show these too, but only as rendered plot text, not as
## console output that's easy to copy/compare.
summarise_r <- function(df, predictor_col, strain, predictor_label) {
  ct <- cor.test(df[[predictor_col]], df$log2fc_large_small, method = "spearman")
  tibble(strain = strain, predictor = predictor_label,
         r = round(unname(ct$estimate), 3), n = nrow(df),
         p_value = ct$p.value)
}

r_summary <- bind_rows(
  summarise_r(occ_fc %>% filter(strain_group == "WT"), "log2_mean_occupancy", "WT", "Occupancy"),
  summarise_r(occ_fc %>% filter(strain_group == "WT"), "log2_tpm_baseline", "WT", "TPM"),
  summarise_r(miller_fc %>% filter(strain_group == "WT"), "log2_synth_rate", "WT", "Synth rate (Miller)"),
  summarise_r(neymotin_fc %>% filter(strain_group == "WT"), "log2_synth_rate", "WT", "Synth rate (Neymotin)"),
  summarise_r(occ_fc %>% filter(strain_group == "Bcy1-AID"), "log2_mean_occupancy", "Bcy1-AID", "Occupancy"),
  summarise_r(occ_fc %>% filter(strain_group == "Bcy1-AID"), "log2_tpm_baseline", "Bcy1-AID", "TPM"),
  summarise_r(miller_fc %>% filter(strain_group == "Bcy1-AID"), "log2_synth_rate", "Bcy1-AID", "Synth rate (Miller)"),
  summarise_r(neymotin_fc %>% filter(strain_group == "Bcy1-AID"), "log2_synth_rate", "Bcy1-AID", "Synth rate (Neymotin)")
)

cat("All correlations, WT and Bcy1-AID:\n")
print(r_summary, n = Inf)
```

```{r combine-panels}
library(patchwork)

# Figure 15A: WT+PhIAA, all four predictors
p_figure15a <- (p_wt_occupancy | p_wt_tpm) / (p_wt_miller | p_wt_neymotin)

# Figure 15B: Bcy1-AID+PhIAA, all four predictors
p_figure15b <- (p_bcy1_occupancy | p_bcy1_tpm) / (p_bcy1_miller | p_bcy1_neymotin)

print(p_figure15a)
print(p_figure15b)
```

```{r save-outputs}
ggsave(file.path(figures_dir, "figure15A_WT_PhIAA_size_fc_predictors.png"),
       plot = p_figure15a, width = 12, height = 10, dpi = 300)
ggsave(file.path(figures_dir, "figure15B_Bcy1AID_PhIAA_size_fc_predictors.png"),
       plot = p_figure15b, width = 12, height = 10, dpi = 300)

cat("Done - Figure 15A/15B saved to figures/\n")
```
