---
title: "RNAPII Occupancy, TPM, and Synthesis Rate vs Small/Large Cell FC - WT+PhIAA and BCY1-AID+PhIAA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this file's location
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Rebuild of the Bcy1-AID+PhIAA occupancy figure. Changes from script 14:

1. Y-axis changed from log2FC(PhIAA/DMSO) to log2FC(large-PhIAA vs
   small-PhIAA) - fold change between cell sizes within PhIAA-treated
   cells, removing the need for a DMSO control.
2. Two separate graphs, WT+PhIAA and BCY1-AID+PhIAA.
3. Correlation switched from Pearson to Spearman.

Both strains are drawn from the same batch (exp250624): MS69 (WT) and
MS756 (Bcy1-AID), matched PhIAA treatment at small (1hr) and large
(4.75hr) arrest.

EXTENDED: adds TPM and synthesis rate as additional predictors
alongside occupancy (Figure 15A/15B). Synthesis rate is log2-
transformed and kept separate for Miller and Neymotin (not averaged).

ASSUMPTIONS:
- Cell size (small/large) from the arrest column, which uses different
  label formats per strain: MS69's large sample is "4.75hr G1 arrest",
  MS756's is "Large G1 arrest". Both patterns matched explicitly.
- "WT" = strain_id MS69, "Bcy1-AID" = strain_id MS756 (not MS756c2,
  which isn't in this batch).
- Restricted to batch "exp250624" via string matching on the sample
  column.

```{r load-occupancy}
# Same averaged occupancy calculation as script 13a/13b: log2(mean
# occupancy + 1) across 6 experiments, requiring >= 4 of 6 present.
occupancy_path <- file.path(
  tables_dir,
  "210601_gene_annotation_RPB_occupancy.txt"
)

occupancy_raw <- read_tsv(occupancy_path, show_col_types = FALSE)

occupancy_conditions <- c("A", "B", "C", "EFG", "H", "J")
rpkm_cols <- paste0("RPKM_mean_", occupancy_conditions)

occupancy <- occupancy_raw %>%
  select(systematic_name, all_of(rpkm_cols)) %>%
  rename(gene_id = systematic_name) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  rowwise() %>%
  mutate(
    n_experiments_present = sum(!is.na(c_across(all_of(rpkm_cols)))),
    mean_occupancy         = mean(c_across(all_of(rpkm_cols)), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(n_experiments_present >= 4) %>%
  mutate(log2_mean_occupancy = log2(mean_occupancy + 1)) %>%
  select(gene_id, log2_mean_occupancy)

cat("Genes with usable occupancy data (>= 4 of 6 experiments):",
    nrow(occupancy), "\n")
```

```{r load-synth-rate}
## Synthesis rate: Miller and Neymotin kept SEPARATE (not averaged),
## each log2-transformed before correlating.
## filter(> 0) before log2() since log2 of zero/negative is undefined.
synth_rate <- occupancy_raw %>%
  select(systematic_name, miller_synth_rate, neymotin_synth_rate) %>%
  rename(gene_id = systematic_name) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  filter(!is.na(miller_synth_rate) | !is.na(neymotin_synth_rate))

synth_rate_miller <- synth_rate %>%
  filter(!is.na(miller_synth_rate), miller_synth_rate > 0) %>%
  transmute(gene_id, log2_synth_rate = log2(miller_synth_rate))

synth_rate_neymotin <- synth_rate %>%
  filter(!is.na(neymotin_synth_rate), neymotin_synth_rate > 0) %>%
  transmute(gene_id, log2_synth_rate = log2(neymotin_synth_rate))

cat("Genes with Miller synthesis rate (log2):", nrow(synth_rate_miller), "\n")
cat("Genes with Neymotin synthesis rate (log2):", nrow(synth_rate_neymotin), "\n")
```

```{r load-tpm-samples}
df_sc_final <- read_tsv(file.path(tables_dir, "df_sc_final.txt"),
                        show_col_types = FALSE)

# Restrict to exp250624, both strains, PhIAA only.
#
# Cell size from the arrest column: MS756 uses "Large G1 arrest",
# MS69 uses "4.75hr G1 arrest" for the same timepoint. Both patterns
# matched below.
exp250624_samples <- df_sc_final %>%
  filter(strain_id %in% c("MS69", "MS756"),
         condition == "PhIAA",
         str_detect(sample, "exp250624")) %>%
  mutate(
    strain_group = case_when(
      strain_id == "MS69"  ~ "WT",
      strain_id == "MS756" ~ "Bcy1-AID",
      TRUE ~ NA_character_
    ),
    arrest_hours = as.numeric(str_extract(arrest, "^[0-9.]+")),
    cell_size_group = case_when(
      str_detect(arrest, "^1hr") ~ "small",
      str_detect(arrest, "^Large") | (!is.na(arrest_hours) & arrest_hours >= 4) ~ "large",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(strain_group), !is.na(cell_size_group))

cat("exp250624 PhIAA-treated samples identified (expect 4 rows):\n")
print(exp250624_samples %>%
        distinct(sample, strain_id, strain_group, arrest, arrest_hours,
                 cell_size_group, mean_cell_volume))
```

```{r compute-small-large-log2fc}
# Mean TPMsc per gene x strain x cell size, then log2FC(large/small)
mean_tpm_by_group <- exp250624_samples %>%
  group_by(gene_id, strain_group, cell_size_group) %>%
  summarise(mean_tpmsc = mean(TPMsc, na.rm = TRUE), .groups = "drop")

log2fc_small_large <- mean_tpm_by_group %>%
  pivot_wider(names_from = cell_size_group, values_from = mean_tpmsc) %>%
  filter(!is.na(large), !is.na(small), large > 0, small > 0) %>%
  mutate(log2fc_large_small = log2(large / small))

cat("Genes with log2FC(large/small) computed, by strain:\n")
print(table(log2fc_small_large$strain_group))
```

```{r compute-tpm-baseline}
## TPM baseline predictor: mean small-cell TPMsc per gene x strain.
##
## Uses small-cell TPM only, not mean(small, large) - the latter is
## mathematically coupled to the outcome (log2fc_large_small =
## log2(large) - log2(small)), introducing circularity. Small-cell-only
## baseline gives WT r = -0.114; the blended baseline sign-flips to
## +0.095.
tpm_baseline <- exp250624_samples %>%
  filter(cell_size_group == "small") %>%
  group_by(gene_id, strain_group) %>%
  summarise(mean_tpmsc_baseline = mean(TPMsc, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(mean_tpmsc_baseline), mean_tpmsc_baseline > 0) %>%
  mutate(log2_tpm_baseline = log2(mean_tpmsc_baseline))

cat("Genes with TPM baseline (small-cell only), by strain:\n")
print(table(tpm_baseline$strain_group))
```

```{r join-predictors}
## Join predictors (occupancy, Miller/Neymotin synth rate, TPM
## baseline) against log2fc_large_small, per strain. Occupancy and
## synth rate are gene-level only (same value joined to both strains);
## TPM baseline is strain-specific.
occ_fc <- log2fc_small_large %>%
  inner_join(occupancy, by = "gene_id") %>%
  inner_join(tpm_baseline, by = c("gene_id", "strain_group"))

miller_fc <- log2fc_small_large %>%
  inner_join(synth_rate_miller, by = "gene_id")

neymotin_fc <- log2fc_small_large %>%
  inner_join(synth_rate_neymotin, by = "gene_id")

cat("Genes with occupancy + TPM baseline + FC, by strain:\n")
print(table(occ_fc$strain_group))
cat("\nGenes with Miller synthesis rate + FC, by strain:\n")
print(table(miller_fc$strain_group))
cat("\nGenes with Neymotin synthesis rate + FC, by strain:\n")
print(table(neymotin_fc$strain_group))
```

```{r define-plot-function}
# Binned mean overlay, matching the convention used in scripts 13a/13b,
# plus Spearman correlation, using a rank-based approach rather than
# Pearson. Generalised to take any predictor column, so the same
# function builds the occupancy, TPM, and both synthesis-rate panels.
n_bins <- 30

plot_predictor_vs_fc <- function(df, predictor_col, x_lab, strain_label, point_colour) {

  spearman_result <- cor.test(df[[predictor_col]],
                              df$log2fc_large_small,
                              method = "spearman")

  binned <- df %>%
    mutate(pred_bin = ntile(.data[[predictor_col]], n_bins)) %>%
    group_by(pred_bin) %>%
    summarise(
      pred_mean = mean(.data[[predictor_col]]),
      fc_mean   = mean(log2fc_large_small),
      .groups = "drop"
    )

  ggplot(df, aes(x = .data[[predictor_col]], y = log2fc_large_small)) +
    geom_point(color = point_colour, alpha = 0.3, size = 1) +
    geom_line(data = binned, aes(x = pred_mean, y = fc_mean),
              color = "black", linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = project_colours[["grey"]]) +
    labs(
      subtitle = paste0(
        strain_label, "  |  n = ", nrow(df), " genes  |  ",
        "Spearman r = ", round(unname(spearman_result$estimate), 3),
        "  |  p ", if (spearman_result$p.value < 0.001) "< 0.001"
                   else paste0("= ", round(spearman_result$p.value, 3))
      ),
      x = x_lab,
      y = "log2 FC (large-PhIAA / small-PhIAA)"
    ) +
    theme_bw()
}
```

```{r plot-wt}
p_wt_occupancy <- plot_predictor_vs_fc(
  occ_fc %>% filter(strain_group == "WT"),
  predictor_col = "log2_mean_occupancy",
  x_lab         = "log2(mean RNAPII occupancy + 1)",
  strain_label  = "WT + PhIAA",
  point_colour  = project_colours[["purple"]]
)

p_wt_tpm <- plot_predictor_vs_fc(
  occ_fc %>% filter(strain_group == "WT"),
  predictor_col = "log2_tpm_baseline",
  x_lab         = "log2(TPM baseline)",
  strain_label  = "WT + PhIAA",
  point_colour  = project_colours[["purple"]]
)

p_wt_miller <- plot_predictor_vs_fc(
  miller_fc %>% filter(strain_group == "WT"),
  predictor_col = "log2_synth_rate",
  x_lab         = "log2(synthesis rate, Miller)",
  strain_label  = "WT + PhIAA",
  point_colour  = project_colours[["purple"]]
)

p_wt_neymotin <- plot_predictor_vs_fc(
  neymotin_fc %>% filter(strain_group == "WT"),
  predictor_col = "log2_synth_rate",
  x_lab         = "log2(synthesis rate, Neymotin)",
  strain_label  = "WT + PhIAA",
  point_colour  = project_colours[["purple"]]
)

print(p_wt_occupancy)
print(p_wt_tpm)
print(p_wt_miller)
print(p_wt_neymotin)
```

```{r plot-bcy1}
p_bcy1_occupancy <- plot_predictor_vs_fc(
  occ_fc %>% filter(strain_group == "Bcy1-AID"),
  predictor_col = "log2_mean_occupancy",
  x_lab         = "log2(mean RNAPII occupancy + 1)",
  strain_label  = "Bcy1-AID + PhIAA",
  point_colour  = project_colours[["red"]]
)

p_bcy1_tpm <- plot_predictor_vs_fc(
  occ_fc %>% filter(strain_group == "Bcy1-AID"),
  predictor_col = "log2_tpm_baseline",
  x_lab         = "log2(TPM baseline)",
  strain_label  = "Bcy1-AID + PhIAA",
  point_colour  = project_colours[["red"]]
)

p_bcy1_miller <- plot_predictor_vs_fc(
  miller_fc %>% filter(strain_group == "Bcy1-AID"),
  predictor_col = "log2_synth_rate",
  x_lab         = "log2(synthesis rate, Miller)",
  strain_label  = "Bcy1-AID + PhIAA",
  point_colour  = project_colours[["red"]]
)

p_bcy1_neymotin <- plot_predictor_vs_fc(
  neymotin_fc %>% filter(strain_group == "Bcy1-AID"),
  predictor_col = "log2_synth_rate",
  x_lab         = "log2(synthesis rate, Neymotin)",
  strain_label  = "Bcy1-AID + PhIAA",
  point_colour  = project_colours[["red"]]
)

print(p_bcy1_occupancy)
print(p_bcy1_tpm)
print(p_bcy1_miller)
print(p_bcy1_neymotin)
```

```{r summary-table}
## Direct numeric summary of all eight correlations (console output,
## easy to copy/compare vs the plot subtitles)
summarise_r <- function(df, predictor_col, strain, predictor_label) {
  ct <- cor.test(df[[predictor_col]], df$log2fc_large_small, method = "spearman")
  tibble(strain = strain, predictor = predictor_label,
         r = round(unname(ct$estimate), 3), n = nrow(df),
         p_value = ct$p.value)
}

r_summary <- bind_rows(
  summarise_r(occ_fc %>% filter(strain_group == "WT"), "log2_mean_occupancy", "WT", "Occupancy"),
  summarise_r(occ_fc %>% filter(strain_group == "WT"), "log2_tpm_baseline", "WT", "TPM"),
  summarise_r(miller_fc %>% filter(strain_group == "WT"), "log2_synth_rate", "WT", "Synth rate (Miller)"),
  summarise_r(neymotin_fc %>% filter(strain_group == "WT"), "log2_synth_rate", "WT", "Synth rate (Neymotin)"),
  summarise_r(occ_fc %>% filter(strain_group == "Bcy1-AID"), "log2_mean_occupancy", "Bcy1-AID", "Occupancy"),
  summarise_r(occ_fc %>% filter(strain_group == "Bcy1-AID"), "log2_tpm_baseline", "Bcy1-AID", "TPM"),
  summarise_r(miller_fc %>% filter(strain_group == "Bcy1-AID"), "log2_synth_rate", "Bcy1-AID", "Synth rate (Miller)"),
  summarise_r(neymotin_fc %>% filter(strain_group == "Bcy1-AID"), "log2_synth_rate", "Bcy1-AID", "Synth rate (Neymotin)")
)

cat("All correlations, WT and Bcy1-AID:\n")
print(r_summary, n = Inf)
```

```{r combine-panels}
library(patchwork)

# Figure 15A: WT+PhIAA, all four predictors
p_figure15a <- (p_wt_occupancy | p_wt_tpm) / (p_wt_miller | p_wt_neymotin)

# Figure 15B: Bcy1-AID+PhIAA, all four predictors
p_figure15b <- (p_bcy1_occupancy | p_bcy1_tpm) / (p_bcy1_miller | p_bcy1_neymotin)

print(p_figure15a)
print(p_figure15b)
```

```{r save-outputs}
ggsave(file.path(figures_dir, "figure15A_WT_PhIAA_size_fc_predictors.png"),
       plot = p_figure15a, width = 12, height = 10, dpi = 300)
ggsave(file.path(figures_dir, "figure15B_Bcy1AID_PhIAA_size_fc_predictors.png"),
       plot = p_figure15b, width = 12, height = 10, dpi = 300)

cat("Done - Figure 15A/15B saved to figures/\n")
```
