---
title: "RNAPII Occupancy vs Small/Large Cell FC - WT+PhIAA and BCY1-AID+PhIAA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Paths relative to this Rmd's location (Adriana_analysis/scripts/)
project_dir <- ".."
load(file.path(project_dir, "project_functions.rda"))

paths       <- setup_project_paths(project_dir)
tables_dir  <- paths$tables_dir
figures_dir <- paths$figures_dir
output_dir  <- paths$output_dir
```

```{r libraries}
library(tidyverse)

select <- dplyr::select

# Shared colour palette for dissertation figures
load(file.path(project_dir, "project_colours.rda"))
```

# Background

Rebuild of the Bcy1-AID+PhIAA occupancy figure per Matthew's meeting
feedback. Two changes from the previous version (script 14):

1. Y-axis changed from log2FC(PhIAA/DMSO) to log2FC(large-PhIAA vs
   small-PhIAA) - i.e. fold change between cell sizes within the same
   PhIAA-treated condition, not a drug-vs-vehicle comparison. This
   removes the need for a matched DMSO control entirely.
2. Two separate graphs, WT+PhIAA and BCY1-AID+PhIAA, rather than one
   figure split by facet, so each strain's relationship can be read
   independently.
3. Correlation switched from Pearson to Spearman, per Matthew's
   instruction to use a rank-based approach throughout rather than
   Pearson on ranked/near-ranked data.

Both strains are drawn from the same experimental batch (exp250624),
which contains MS69 (WT) and MS756 (Bcy1-AID) run together with matched
PhIAA treatment at both small (1hr) and large (4.75hr) arrest - this
avoids the batch-matching problem the DMSO-based version had, since no
cross-batch DMSO control is needed for a small-vs-large FC.

ASSUMPTIONS MADE BELOW THAT NEED CHECKING:
- Cell size group (small/large) is determined from the arrest column,
  which uses inconsistent label formats across samples: MS69's large
  sample is labelled "4.75hr G1 arrest" (numeric), while MS756's is
  labelled "Large G1 arrest" (categorical), for what is nominally the
  same timepoint. Both patterns are matched explicitly. Confirmed
  working for both strains in exp250624 - see the printed sample table
  below, which should show exactly 4 rows.
- "WT" = strain_id "MS69", "Bcy1-AID" = strain_id "MS756" (not MS756c2,
  which appears in a different, later experiment and was not part of
  the exp250624 batch).
- Restricting to experiment "exp250624" specifically (rather than
  pooling all PhIAA-treated MS69/MS756 samples across every batch) is
  done via string matching on the sample column, since this is the one
  batch containing both strains run together under matched conditions.
  If a dedicated experiment ID column exists in df_sc_final.txt beyond
  what's embedded in the sample name, this should be used instead -
  check column names before running.

```{r load-occupancy}
# Reuses the same averaged occupancy calculation as script 13a/13b -
# log2(mean RNAPII occupancy + 1) across the 6 available experiments
# (A, B, C, EFG, H, J), requiring at least 4 of 6 present per gene.
occupancy_path <- file.path(
  tables_dir,
  "210601_gene_annotation_RPB_occupancy.txt"
)

occupancy_raw <- read_tsv(occupancy_path, show_col_types = FALSE)

occupancy_conditions <- c("A", "B", "C", "EFG", "H", "J")
rpkm_cols <- paste0("RPKM_mean_", occupancy_conditions)

occupancy <- occupancy_raw %>%
  select(systematic_name, all_of(rpkm_cols)) %>%
  rename(gene_id = systematic_name) %>%
  rowwise() %>%
  mutate(
    n_experiments_present = sum(!is.na(c_across(all_of(rpkm_cols)))),
    mean_occupancy         = mean(c_across(all_of(rpkm_cols)), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(n_experiments_present >= 4) %>%
  mutate(log2_mean_occupancy = log2(mean_occupancy + 1)) %>%
  select(gene_id, log2_mean_occupancy)

cat("Genes with usable occupancy data (>= 4 of 6 experiments):",
    nrow(occupancy), "\n")
```

```{r load-tpm-samples}
df_sc_final <- read_tsv(file.path(tables_dir, "df_sc_final.txt"),
                        show_col_types = FALSE)

# Restrict to the exp250624 batch specifically, both strains, PhIAA
# treatment only (no DMSO needed for this comparison).
#
# Cell size group is determined from the arrest column, which uses two
# different label formats depending on the sample: some rows give a
# numeric duration ("4.75hr G1 arrest"), others give a categorical
# label ("Large G1 arrest") for what is nominally the same timepoint -
# confirmed by direct inspection that MS756's exp250624 large-cell
# sample uses "Large G1 arrest" specifically, while MS69's equivalent
# sample uses "4.75hr G1 arrest". Both patterns are matched below so
# neither strain is silently dropped.
exp250624_samples <- df_sc_final %>%
  filter(strain_id %in% c("MS69", "MS756"),
         condition == "PhIAA",
         str_detect(sample, "exp250624")) %>%
  mutate(
    strain_group = case_when(
      strain_id == "MS69"  ~ "WT",
      strain_id == "MS756" ~ "Bcy1-AID",
      TRUE ~ NA_character_
    ),
    arrest_hours = as.numeric(str_extract(arrest, "^[0-9.]+")),
    cell_size_group = case_when(
      str_detect(arrest, "^1hr") ~ "small",
      str_detect(arrest, "^Large") | (!is.na(arrest_hours) & arrest_hours >= 4) ~ "large",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(strain_group), !is.na(cell_size_group))

cat("exp250624 PhIAA-treated samples identified:\n")
print(exp250624_samples %>%
        distinct(sample, strain_id, strain_group, arrest, arrest_hours,
                 cell_size_group, mean_cell_volume))
```

```{r compute-small-large-log2fc}
# Mean TPMsc per gene x strain x cell size (averaging any replicate
# samples within exp250624), then log2FC(large/small) within each
# strain separately.
mean_tpm_by_group <- exp250624_samples %>%
  group_by(gene_id, strain_group, cell_size_group) %>%
  summarise(mean_tpmsc = mean(TPMsc, na.rm = TRUE), .groups = "drop")

log2fc_small_large <- mean_tpm_by_group %>%
  pivot_wider(names_from = cell_size_group, values_from = mean_tpmsc) %>%
  filter(!is.na(large), !is.na(small), large > 0, small > 0) %>%
  mutate(log2fc_large_small = log2(large / small))

cat("Genes with log2FC(large/small) computed, by strain:\n")
print(table(log2fc_small_large$strain_group))
```

```{r join-occupancy}
occ_fc <- log2fc_small_large %>%
  inner_join(occupancy, by = "gene_id")

cat("Genes with both occupancy and log2FC(large/small):\n")
print(table(occ_fc$strain_group))
```

```{r define-plot-function}
# Binned mean overlay, matching the convention used in scripts 13a/13b,
# plus Spearman correlation per Matthew's instruction to use a
# rank-based approach rather than Pearson.
n_bins <- 30

plot_occupancy_vs_fc <- function(df, strain_label, point_colour) {

  spearman_result <- cor.test(df$log2_mean_occupancy,
                              df$log2fc_large_small,
                              method = "spearman")

  binned <- df %>%
    mutate(occ_bin = ntile(log2_mean_occupancy, n_bins)) %>%
    group_by(occ_bin) %>%
    summarise(
      occ_mean = mean(log2_mean_occupancy),
      fc_mean  = mean(log2fc_large_small),
      .groups = "drop"
    )

  ggplot(df, aes(x = log2_mean_occupancy, y = log2fc_large_small)) +
    geom_point(color = point_colour, alpha = 0.3, size = 1) +
    geom_line(data = binned, aes(x = occ_mean, y = fc_mean),
              color = "black", linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = project_colours[["grey"]]) +
    labs(
      subtitle = paste0(
        strain_label, "  |  n = ", nrow(df), " genes  |  ",
        "Spearman r = ", round(unname(spearman_result$estimate), 3),
        "  |  p ", if (spearman_result$p.value < 0.001) "< 0.001"
                   else paste0("= ", round(spearman_result$p.value, 3))
      ),
      x = "log2(mean RNAPII occupancy + 1)",
      y = "log2 FC (large-PhIAA / small-PhIAA)"
    ) +
    theme_bw()
}
```

```{r plot-wt}
p_wt <- plot_occupancy_vs_fc(
  occ_fc %>% filter(strain_group == "WT"),
  strain_label = "WT + PhIAA",
  point_colour = project_colours[["purple"]]
)

print(p_wt)
```

```{r plot-bcy1}
p_bcy1 <- plot_occupancy_vs_fc(
  occ_fc %>% filter(strain_group == "Bcy1-AID"),
  strain_label = "Bcy1-AID + PhIAA",
  point_colour = project_colours[["red"]]
)

print(p_bcy1)
```

```{r save-outputs}
ggsave(file.path(figures_dir, "figure_wt_phiaa_occupancy_vs_size_fc.png"),
       plot = p_wt, width = 7, height = 5.5, dpi = 300)
ggsave(file.path(figures_dir, "figure_bcy1_phiaa_occupancy_vs_size_fc.png"),
       plot = p_bcy1, width = 7, height = 5.5, dpi = 300)

cat("Done - both figures saved to figures/\n")
```
