---
title: "Translation Efficiency and Poly(A) Tail Correlation Analysis"
author: "CW"
date: "2026-05-19"
output: html_document
---

```{r setup}
rm(list=ls())
library(tidyverse)
library(ggplot2)
library(patchwork)
library(hexbin)
library(Hmisc)
load("plotting_functions.rda")
load("correlation_functions_updated.rda")
```

```{r load}
sample_info  <- read_csv("../tables/sample_info_combined.csv", show_col_types = FALSE)
df_TE        <- read_csv("../tables/translation_efficiency.csv", show_col_types = FALSE)
df_ONT       <- read_csv("../tables/gene_fraction&tail_length.csv", show_col_types = FALSE)
verified_ORF <- read_tsv("../tables/2230128_verified_ORF.txt", show_col_types = FALSE) %>%
  filter(species == "S_cerevisiae")
df_hl <- read_tsv("../output_tables/df_fit_output_c1.txt", show_col_types = FALSE) %>%
  left_join(sample_info, join_by(experiment_strain_condition)) %>%
  filter(condition == "WT", std.error_log_k < 1)
## condition == "WT": MS96 WT + MS1039 DMSO control
```

```{r check}
colnames(df_TE)
colnames(df_ONT)
colnames(df_hl)
df_TE %>% dplyr::select(starts_with("TPMsc")) %>% summary()
```

```{r filter_join}
## translation_efficiency.csv:
## TE_WT_PP -- mean of Arava 2003, Park 2011 HP, Park 2011 LP (polysome profiling, mean-normalised)
## TE_WT_RS -- mean of Gerashchenko 2012, Sen et al. (ribo-seq, mean-normalised)
## TE_rRNAdep_MandP/P_small/large_YM -- in-house size-separated polysome profiling
## TPMsc_input/MandP/P_small/large_YM -- used for filtering only
## TE_polyA columns excluded -- near-zero correlation with rRNAdep; input fraction unclear
##
## gene_fraction&tail_length.csv:
## MS69_small/large_mrl100_mean_tail_length -- mean poly(A) tail length, WT strain MS69, Nanopore

df_TE_filtered <- df_TE %>%
  filter(TPMsc_input_small_YM > 1,
         TPMsc_input_large_YM > 1,
         TPMsc_MandP_small_YM > 1,
         TPMsc_MandP_large_YM > 1,
         TPMsc_P_small_YM     > 1,
         TPMsc_P_large_YM     > 1)

df_TE_ONT          <- left_join(df_TE,          df_ONT, join_by(gene_id))
df_TE_ONT_filtered <- left_join(df_TE_filtered, df_ONT, join_by(gene_id))
```

```{r halflife}
## output_tables/df_fit_output_c1.txt: estimate_half_life per gene per condition
## filtered std.error_log_k < 1, averaged across MS96 + MS1039 DMSO replicates

df_hl_wide <- df_hl %>%
  group_by(gene_id, size) %>%
  summarise(mean_half_life = mean(estimate_half_life, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = gene_id,
              names_from  = size,
              values_from = mean_half_life,
              names_glue  = "hl_{size}")

df_combined_hl <- df_hl_wide %>%
  left_join(df_TE,          join_by(gene_id)) %>%
  left_join(df_ONT,         join_by(gene_id))

df_combined_hl_filtered <- df_hl_wide %>%
  left_join(df_TE_filtered, join_by(gene_id)) %>%
  left_join(df_ONT,         join_by(gene_id))

nrow(df_combined_hl)
```

## Section 1 -- Polysome profiling vs ribo-seq TE (bulk WT literature datasets)
## translation_efficiency.csv: TE_WT_PP vs TE_WT_RS

```{r s1}
p1 <- hex_cor(df_TE,
              "TE_WT_PP", "TE_WT_RS",
              "Literature PP TE (log2)",
              "Literature RS TE (log2)",
              "Translation efficiency: polysome profiling vs ribo-seq in bulk WT cells")
p1
save_single(p1, "../figures/TE_PP_vs_RS_WT.png")
```

## Section 2 -- Translation efficiency: small vs large cells
## translation_efficiency.csv: TE_rRNAdep_MandP_small/large_YM, TE_rRNAdep_P_small/large_YM

```{r s2}
p2 <- scatter_cor_pair(df_TE_filtered,
                       "TE_rRNAdep_MandP_small_YM", "TE_rRNAdep_MandP_large_YM",
                       "TE_rRNAdep_P_small_YM",     "TE_rRNAdep_P_large_YM",
                       "TE — small cells (log2)",
                       "TE — large cells (log2)",
                       "Translation efficiency (monosome + polysome): small vs large WT cells",
                       "Translation efficiency (polysome only): small vs large WT cells",
                       show_diagonal = TRUE)
p2
save_pair(p2, "../figures/TE_small_vs_large_filtered.png")
```

## Section 3a -- Literature PP vs lab size-separated TE (same technique, different labs)
## translation_efficiency.csv: TE_WT_PP vs TE_rRNAdep_MandP/P_small_YM

```{r s3a}
p3a <- scatter_cor(df_TE_filtered,
                   "TE_WT_PP", "TE_rRNAdep_MandP_small_YM",
                   "Literature PP TE (log2)",
                   "Lab MandP TE — small cells (log2)",
                   "Bulk WT polysome profiling TE vs size-separated monosome + polysome TE (small cells)")

p3b <- scatter_cor(df_TE_filtered,
                   "TE_WT_PP", "TE_rRNAdep_P_small_YM",
                   "Literature PP TE (log2)",
                   "Lab P only TE — small cells (log2)",
                   "Bulk WT polysome profiling TE vs size-separated polysome only TE (small cells)")

p3a + p3b
save_pair(p3a + p3b, "../figures/TE_litPP_vs_lab_sizesep.png")
```

## Section 3b -- Literature RS vs lab size-separated TE (different techniques)
## translation_efficiency.csv: TE_WT_RS vs TE_rRNAdep_MandP/P_small_YM

```{r s3b}
p3c <- scatter_cor(df_TE_filtered,
                   "TE_WT_RS", "TE_rRNAdep_MandP_small_YM",
                   "Literature RS TE (log2)",
                   "Lab MandP TE — small cells (log2)",
                   "Bulk WT ribo-seq TE vs size-separated monosome + polysome TE (small cells)")

p3d <- scatter_cor(df_TE_filtered,
                   "TE_WT_RS", "TE_rRNAdep_P_small_YM",
                   "Literature RS TE (log2)",
                   "Lab P only TE — small cells (log2)",
                   "Bulk WT ribo-seq TE vs size-separated polysome only TE (small cells)")

p3c + p3d
save_pair(p3c + p3d, "../figures/TE_litRS_vs_lab_sizesep.png")
```

## Section 3c -- Lab MandP vs lab P only (effect of monosome inclusion)
## translation_efficiency.csv: TE_rRNAdep_MandP vs TE_rRNAdep_P small/large

```{r s3c}
p3ef <- scatter_cor_pair(df_TE_filtered,
                         "TE_rRNAdep_MandP_small_YM", "TE_rRNAdep_P_small_YM",
                         "TE_rRNAdep_MandP_large_YM", "TE_rRNAdep_P_large_YM",
                         "Lab MandP TE (log2)",
                         "Lab P only TE (log2)",
                         "Monosome + polysome vs polysome only TE in small WT cells",
                         "Monosome + polysome vs polysome only TE in large WT cells",
                         show_diagonal = TRUE)
p3ef
save_pair(p3ef, "../figures/TE_MandP_vs_Ponly.png")
```

## Section 4 -- Bulk WT translation efficiency vs poly(A) tail length
## translation_efficiency.csv: TE_WT_PP, TE_WT_RS
## gene_fraction&tail_length.csv: MS69_small/large_mrl100_mean_tail_length

```{r s4}
p4 <- hex_cor_pair(df_TE_ONT,
                   "TE_WT_PP", "MS69_small_mrl100_mean_tail_length",
                   "TE_WT_PP", "MS69_large_mrl100_mean_tail_length",
                   "Literature PP TE (log2)",
                   "Mean poly(A) tail length (log2)",
                   "Bulk WT polysome profiling TE vs mean poly(A) tail length (small cells)",
                   "Bulk WT polysome profiling TE vs mean poly(A) tail length (large cells)")
p4
save_pair(p4, "../figures/TE_PP_vs_polyA_tail_small_large.png")

p5 <- hex_cor_pair(df_TE_ONT,
                   "TE_WT_RS", "MS69_small_mrl100_mean_tail_length",
                   "TE_WT_RS", "MS69_large_mrl100_mean_tail_length",
                   "Literature RS TE (log2)",
                   "Mean poly(A) tail length (log2)",
                   "Bulk WT ribo-seq TE vs mean poly(A) tail length (small cells)",
                   "Bulk WT ribo-seq TE vs mean poly(A) tail length (large cells)")
p5
save_pair(p5, "../figures/TE_RS_vs_polyA_tail_small_large.png")
```

## Section 5 -- poly(A) tail length: small vs large cells
## gene_fraction&tail_length.csv: MS69_small/large_mrl100_mean_tail_length
## coord_equal = TRUE -- both axes are the same variable and comparable range

```{r s5}
p6 <- hex_cor(df_TE_ONT,
              "MS69_small_mrl100_mean_tail_length",
              "MS69_large_mrl100_mean_tail_length",
              "Mean poly(A) tail length — small cells (log2)",
              "Mean poly(A) tail length — large cells (log2)",
              "Mean poly(A) tail length: small vs large WT cells",
              show_diagonal = TRUE,
              coord_equal   = TRUE)
p6
save_single(p6, "../figures/polyA_tail_small_vs_large.png")
```

## Section 6 -- Lab size-separated TE vs poly(A) tail length
## translation_efficiency.csv: TE_rRNAdep_MandP/P_small/large_YM
## gene_fraction&tail_length.csv: MS69_small/large_mrl100_mean_tail_length
## Conditions matched: small TE vs small tail, large TE vs large tail
## Binned mean line used -- shows whether weak correlations are consistent across range

```{r s6}
p7 <- scatter_cor_binned_pair(df_TE_ONT_filtered,
                              "TE_rRNAdep_MandP_small_YM", "MS69_small_mrl100_mean_tail_length",
                              "TE_rRNAdep_MandP_large_YM", "MS69_large_mrl100_mean_tail_length",
                              "Lab MandP TE (log2)",
                              "Mean poly(A) tail length (log2)",
                              "Size-separated monosome + polysome TE vs mean poly(A) tail length (small WT cells)",
                              "Size-separated monosome + polysome TE vs mean poly(A) tail length (large WT cells)",
                              bins=30)
p7
save_pair(p7, "../figures/TE_MandP_vs_polyA_small_large_filtered.png")

p8 <- scatter_cor_binned_pair(df_TE_ONT_filtered,
                              "TE_rRNAdep_P_small_YM", "MS69_small_mrl100_mean_tail_length",
                              "TE_rRNAdep_P_large_YM", "MS69_large_mrl100_mean_tail_length",
                              "Lab P only TE (log2)",
                              "Mean poly(A) tail length (log2)",
                              "Size-separated polysome only TE vs mean poly(A) tail length (small WT cells)",
                              "Size-separated polysome only TE vs mean poly(A) tail length (large WT cells)",
                              bins=30)
p8
save_pair(p8, "../figures/TE_P_vs_polyA_small_large_filtered.png")
```

## Section 7 -- mRNA half-life vs bulk WT translation efficiency
## output_tables/df_fit_output_c1.txt: hl_small, hl_large
## translation_efficiency.csv: TE_WT_PP, TE_WT_RS

```{r s7}
p9 <- hex_cor_pair(df_combined_hl,
                   "TE_WT_PP", "hl_small",
                   "TE_WT_PP", "hl_large",
                   "Literature PP TE (log2)",
                   "mRNA half-life (log2)",
                   "Bulk WT polysome profiling TE vs mRNA half-life (small cells)",
                   "Bulk WT polysome profiling TE vs mRNA half-life (large cells)")
p9
save_pair(p9, "../figures/TE_PP_vs_half_life_small_large.png")

p10 <- hex_cor_pair(df_combined_hl,
                    "TE_WT_RS", "hl_small",
                    "TE_WT_RS", "hl_large",
                    "Literature RS TE (log2)",
                    "mRNA half-life (log2)",
                    "Bulk WT ribo-seq TE vs mRNA half-life (small cells)",
                    "Bulk WT ribo-seq TE vs mRNA half-life (large cells)")
p10
save_pair(p10, "../figures/TE_RS_vs_half_life_small_large.png")
```

## Section 8 -- mRNA half-life vs lab size-separated TE
## output_tables/df_fit_output_c1.txt: hl_small, hl_large
## translation_efficiency.csv: TE_rRNAdep_MandP/P_small/large_YM
## Conditions matched: small TE vs small half-life, large TE vs large half-life
## Binned mean line used -- confirms weakening of correlation in large cells is real

```{r s8}
p11 <- scatter_cor_binned_pair(df_combined_hl_filtered,
                               "TE_rRNAdep_MandP_small_YM", "hl_small",
                               "TE_rRNAdep_MandP_large_YM", "hl_large",
                               "Lab MandP TE (log2)",
                               "mRNA half-life (log2)",
                               "Size-separated monosome + polysome TE vs mRNA half-life (small WT cells)",
                               "Size-separated monosome + polysome TE vs mRNA half-life (large WT cells)",
                               bins=30)
p11
save_pair(p11, "../figures/TE_MandP_vs_half_life_small_large.png")

p12 <- scatter_cor_binned_pair(df_combined_hl_filtered,
                               "TE_rRNAdep_P_small_YM", "hl_small",
                               "TE_rRNAdep_P_large_YM", "hl_large",
                               "Lab P only TE (log2)",
                               "mRNA half-life (log2)",
                               "Size-separated polysome only TE vs mRNA half-life (small WT cells)",
                               "Size-separated polysome only TE vs mRNA half-life (large WT cells)",
                               bins=30)
p12
save_pair(p12, "../figures/TE_P_vs_half_life_small_large.png")
```

## Section 9 -- mRNA half-life vs poly(A) tail length
## output_tables/df_fit_output_c1.txt: hl_small, hl_large
## gene_fraction&tail_length.csv: MS69_small/large_mrl100_mean_tail_length
## Conditions matched: small tail vs small half-life, large tail vs large half-life
## Binned mean line used -- confirms negative correlation is consistent across range

```{r s9}
p13 <- scatter_cor_binned_pair(df_combined_hl,
                               "MS69_small_mrl100_mean_tail_length", "hl_small",
                               "MS69_large_mrl100_mean_tail_length", "hl_large",
                               "Mean poly(A) tail length (log2)",
                               "mRNA half-life (log2)",
                               "Mean poly(A) tail length vs mRNA half-life (small WT cells)",
                               "Mean poly(A) tail length vs mRNA half-life (large WT cells)")
p13
save_pair(p13, "../figures/polyA_tail_vs_half_life_small_large.png")
```

## Section 10 -- mRNA half-life: small vs large cells (core result)
## output_tables/df_fit_output_c1.txt: hl_small, hl_large

```{r s10}
p14 <- scatter_cor(df_combined_hl,
                   "hl_small", "hl_large",
                   "mRNA half-life — small cells (log2)",
                   "mRNA half-life — large cells (log2)",
                   "mRNA half-life: small vs large WT cells",
                   show_diagonal = TRUE)
p14
save_single(p14, "../figures/half_life_small_vs_large.png")
```