---
title: "CW12-Cumulative_Frequency_Distributions"
output: html_document
---

# Cumulative frequency distributions of size-dependent half-life fold change
# for annotated gene groups vs background
# Groups: Not1-sensitive, Xrn1-sensitive, NMD-sensitive (Alalam),
#         eIF4E-sensitive (top quartile YM small cells),
#         Pab1-sensitive (top quartile YM large cells)
# Test: Wilcoxon rank-sum test (two-sided)
# Output: figures/CW12_cumfreq_distributions.png (Not1, Xrn1, NMD)
#         figures/CW12_cumfreq_eIF4E_Pab1.png (eIF4E, Pab1)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/caitlinwaterworth/Library/CloudStorage/OneDrive-UniversityofBath/YM_combined_analysis/EU_combined_analysis")
library(tidyverse)
library(patchwork)

col_small <- "#784283"
col_large <- "#ffa8cb"
col_A     <- "#e5000c"
col_eif4e <- "#2196F3"
col_pab1  <- "#FF9800"
```

```{r load}
# Consensus half-life fold change -- SE < 0.75, c1 fitting
df_fc <- read_tsv("output_tables/df_WT_consensus_hl_fc.txt",
                  show_col_types = FALSE)

# Chappleboim sensitivity flags
df_chapp <- read_tsv("tables/df_chappleboim_gba.txt",
                     show_col_types = FALSE)

# Alalam NMD sensitivity flags
df_alalam <- read_tsv("tables/df_alalam_gba.txt",
                      show_col_types = FALSE) %>%
  dplyr::rename(gene_id = name)

# Yuyang RNA-seq -- eIF4E small cells and Pab1 large cells FC
# NOTE: requires CW13 to be run first
df_yuyang <- read_tsv("tables/df_yuyang_gba.txt",
                      show_col_types = FALSE)

cat("Consensus genes:", nrow(df_fc), "\n")
cat("Not1-sensitive:", sum(df_chapp$not1_sensitive, na.rm = TRUE), "\n")
cat("Xrn1-sensitive:", sum(df_chapp$xrn1_sensitive, na.rm = TRUE), "\n")
cat("NMD-sensitive:", sum(df_alalam$nmd_sensitive, na.rm = TRUE), "\n")
```

```{r merge}
# Define eIF4E-sensitive as top quartile of eIF4E-AID FC in small cells (YM)
# Define Pab1-sensitive as top quartile of Pab1-AID FC in large cells (YM)
eif4e_threshold <- quantile(df_yuyang$log2FC_eIF4E_small_YM,
                             0.75, na.rm = TRUE)
pab1_threshold  <- quantile(df_yuyang$log2FC_Pab1_large_YM,
                             0.75, na.rm = TRUE)

cat("eIF4E top quartile threshold:", round(eif4e_threshold, 3), "\n")
cat("Pab1 top quartile threshold:", round(pab1_threshold, 3), "\n")

df_yuyang_flags <- df_yuyang %>%
  mutate(
    eif4e_sensitive = !is.na(log2FC_eIF4E_small_YM) &
                      log2FC_eIF4E_small_YM >= eif4e_threshold,
    pab1_sensitive  = !is.na(log2FC_Pab1_large_YM) &
                      log2FC_Pab1_large_YM >= pab1_threshold
  ) %>%
  dplyr::select(gene_id, eif4e_sensitive, pab1_sensitive)

cat("eIF4E-sensitive genes:", sum(df_yuyang_flags$eif4e_sensitive), "\n")
cat("Pab1-sensitive genes:", sum(df_yuyang_flags$pab1_sensitive), "\n")

# Merge all flags with fold change
df_plot <- df_fc %>%
  left_join(df_chapp %>% dplyr::select(gene_id, not1_sensitive,
                                        xrn1_sensitive),
            by = "gene_id") %>%
  left_join(df_alalam %>% dplyr::select(gene_id, nmd_sensitive),
            by = "gene_id") %>%
  left_join(df_yuyang_flags, by = "gene_id")

cat("Genes after merge:", nrow(df_plot), "\n")
```

```{r wilcoxon}
# Wilcoxon rank-sum tests: sensitive vs non-sensitive genes
groups <- list(
  Not1  = "not1_sensitive",
  Xrn1  = "xrn1_sensitive",
  NMD   = "nmd_sensitive",
  eIF4E = "eif4e_sensitive",
  Pab1  = "pab1_sensitive"
)

wilcox_results <- map_dfr(names(groups), function(grp) {
  col <- groups[[grp]]
  sensitive <- df_plot %>%
    filter(!is.na(.data[[col]]), .data[[col]] == TRUE,
           !is.na(WT_hl_fc), is.finite(WT_hl_fc)) %>%
    pull(WT_hl_fc)
  background <- df_plot %>%
    filter(!is.na(.data[[col]]), .data[[col]] == FALSE,
           !is.na(WT_hl_fc), is.finite(WT_hl_fc)) %>%
    pull(WT_hl_fc)
  w <- wilcox.test(sensitive, background, alternative = "two.sided")
  tibble(
    group             = grp,
    n_sensitive       = length(sensitive),
    n_background      = length(background),
    median_sensitive  = median(sensitive),
    median_background = median(background),
    W                 = w$statistic,
    p_value           = w$p.value,
    p_label           = ifelse(w$p.value < 0.001, "p < 0.001",
                        ifelse(w$p.value < 0.01,  "p < 0.01",
                        ifelse(w$p.value < 0.05,  "p < 0.05",
                               paste0("p = ", round(w$p.value, 3)))))
  )
})

cat("\nWilcoxon results:\n")
print(wilcox_results, width = Inf)
```

```{r plot-function}
make_cumfreq <- function(df, flag_col, grp_label, grp_col, p_label, n_sensitive) {

  df_sens <- df %>%
    filter(!is.na(.data[[flag_col]]), !is.na(WT_hl_fc),
           is.finite(WT_hl_fc))

  sens_label <- paste0(grp_label, "-sensitive (n = ",
                       format(n_sensitive, big.mark = ","), ")")

  df_lines <- df_sens %>%
    mutate(group = ifelse(.data[[flag_col]], sens_label, "Background")) %>%
    mutate(group = factor(group, levels = c(sens_label, "Background"))) %>%
    group_by(group) %>%
    arrange(WT_hl_fc) %>%
    mutate(cumfreq = seq_along(WT_hl_fc) / n()) %>%
    ungroup()

  ggplot(df_lines, aes(x = WT_hl_fc, y = cumfreq,
                       colour = group, linewidth = group)) +
    geom_line() +
    scale_colour_manual(
      values = c(setNames(grp_col, sens_label), "Background" = "grey50"),
      name = NULL) +
    scale_linewidth_manual(
      values = c(setNames(1.2, sens_label), "Background" = 0.8),
      guide = "none") +
    annotate("text", x = -Inf, y = 0.75, hjust = -0.1,
             label = p_label, size = 5, colour = "black") +
    labs(x     = "log2 half-life FC (large / small cells)",
         y     = "Cumulative frequency") +
    theme_classic(base_size = 14) +
    theme(
      axis.text            = element_text(size = 12, colour = "black"),
      axis.title           = element_text(size = 13, colour = "black"),
      legend.position      = c(0.05, 0.95),
      legend.justification = c(0, 1),
      legend.text          = element_text(size = 11),
      plot.background      = element_rect(fill = "white", colour = NA))
}
```

```{r plot}
# Figure 5: published dataset sensitivity groups
p_not1 <- make_cumfreq(df_plot, "not1_sensitive", "Not1", col_small,
                        wilcox_results$p_label[wilcox_results$group == "Not1"],
                        wilcox_results$n_sensitive[wilcox_results$group == "Not1"]) +
  labs(tag = "A")

p_xrn1 <- make_cumfreq(df_plot, "xrn1_sensitive", "Xrn1", col_large,
                        wilcox_results$p_label[wilcox_results$group == "Xrn1"],
                        wilcox_results$n_sensitive[wilcox_results$group == "Xrn1"]) +
  labs(tag = "B")

p_nmd <- make_cumfreq(df_plot, "nmd_sensitive", "NMD", col_A,
                       wilcox_results$p_label[wilcox_results$group == "NMD"],
                       wilcox_results$n_sensitive[wilcox_results$group == "NMD"]) +
  labs(tag = "C")

p_panel <- p_not1 | p_xrn1 | p_nmd

p_panel
ggsave("figures/CW12_cumfreq_distributions.png", p_panel,
       width  = 18,
       height = 8,
       dpi    = 150)
cat("Saved figures/CW12_cumfreq_distributions.png\n")
```

```{r plot-eif4e-pab1}
# Figure 6: in-house RNA-seq sensitivity groups (Yuyang)
p_eif4e <- make_cumfreq(df_plot, "eif4e_sensitive",
                         "eIF4E", col_eif4e,
                         wilcox_results$p_label[wilcox_results$group == "eIF4E"],
                         wilcox_results$n_sensitive[wilcox_results$group == "eIF4E"]) +
  labs(tag = "A")

p_pab1 <- make_cumfreq(df_plot, "pab1_sensitive",
                        "Pab1", col_pab1,
                        wilcox_results$p_label[wilcox_results$group == "Pab1"],
                        wilcox_results$n_sensitive[wilcox_results$group == "Pab1"]) +
  labs(tag = "B")

p_panel2 <- p_eif4e | p_pab1

p_panel2
ggsave("figures/CW12_cumfreq_eIF4E_Pab1.png", p_panel2,
       width  = 12,
       height = 8,
       dpi    = 150)
cat("Saved figures/CW12_cumfreq_eIF4E_Pab1.png\n")
```