---
title: "WT Replicate Reproducibility and Half-Life Fold Change Analysis"
author: "CW"
date: "2026-05-21"
output: html_document
---

```{r setup}
rm(list=ls())
library(tidyverse)
library(ggplot2)
library(patchwork)
load("plotting_functions.rda")
load("fitting_function.rda")
load("correlation_functions_updated.rda")
load("reproducibility_functions.rda")
```

```{r load}
sample_info  <- read_csv("../tables/sample_info_combined.csv", show_col_types = FALSE)
verified_ORF <- read_tsv("../tables/2230128_verified_ORF.txt", show_col_types = FALSE) %>%
  filter(species == "S_cerevisiae")

hl_files <- dir(path = "../output_tables/",
                pattern = ".*output.*\\.txt",
                full.names = TRUE)

df_hl_long <- hl_files %>%
  map(~read_tsv(., show_col_types = FALSE)) %>%
  set_names(substr(basename(hl_files),
                   nchar(basename(hl_files)) - 5,
                   nchar(basename(hl_files)) - 4)) %>%
  bind_rows(.id = "fitting_strategy") %>%
  left_join(sample_info, join_by(experiment_strain_condition)) %>%
  filter(condition == "WT")
```

```{r process}
df_hl_wide <- df_hl_long %>%
  pivot_wider(id_cols     = gene_id,
              names_from  = c(label, fitting_strategy),
              values_from = estimate_half_life,
              names_glue  = "{label}_estimate_half_life_by_{fitting_strategy}")

fitting_strategies <- c("a1", "b1", "b2", "b3", "c1")
replicates         <- c("MS96", "MS1039")

for (fs in fitting_strategies) {
  for (rep in replicates) {
    small_col <- paste0(rep, "_small_estimate_half_life_by_", fs)
    large_col <- paste0(rep, "_large_estimate_half_life_by_", fs)
    fc_col    <- paste0(rep, "_hl_fc_by_", fs)
    if (small_col %in% colnames(df_hl_wide) & large_col %in% colnames(df_hl_wide)) {
      df_hl_wide[[fc_col]] <- log2(df_hl_wide[[large_col]] / df_hl_wide[[small_col]])
    }
  }
}

nrow(df_hl_wide)
colnames(df_hl_wide)
```

## Section 1 -- Replicate reproducibility: raw half-life MS96 vs MS1039

```{r s1}
walk(fitting_strategies, function(fs) {
  p <- plot_hl_pair(df_hl_wide,
                    fitting_strategy = fs,
                    rep_x = "MS96",
                    rep_y = "MS1039",
                    type  = "halflife")
  ## print suppressed -- RStudio preview causing file permission error on OneDrive
  save_pair(p, paste0("../figures/replicate_halflife_", fs, ".png"))
})
```

## Section 2 -- Replicate reproducibility: half-life fold change MS96 vs MS1039

```{r s2}
walk(fitting_strategies, function(fs) {
  p <- plot_hl_pair(df_hl_wide,
                    fitting_strategy = fs,
                    rep_x = "MS96",
                    rep_y = "MS1039",
                    type  = "fc")
  ## print suppressed -- RStudio preview causing file permission error on OneDrive
  save_single(p, paste0("../figures/replicate_fc_", fs, ".png"))
})
```

## Section 3 -- Correlation matrix: Spearman R across all fitting strategies

```{r s3}
r_summary <- map_dfr(fitting_strategies, function(fs) {
  df <- df_hl_wide %>%
    filter(
      !is.na(.data[[paste0("MS96_small_estimate_half_life_by_",   fs)]]),
      !is.na(.data[[paste0("MS1039_small_estimate_half_life_by_", fs)]]),
      !is.na(.data[[paste0("MS96_large_estimate_half_life_by_",   fs)]]),
      !is.na(.data[[paste0("MS1039_large_estimate_half_life_by_", fs)]]),
      !is.na(.data[[paste0("MS96_hl_fc_by_",                      fs)]]),
      !is.na(.data[[paste0("MS1039_hl_fc_by_",                    fs)]]),
      .data[[paste0("MS96_small_estimate_half_life_by_",   fs)]] > 0,
      .data[[paste0("MS1039_small_estimate_half_life_by_", fs)]] > 0,
      .data[[paste0("MS96_large_estimate_half_life_by_",   fs)]] > 0,
      .data[[paste0("MS1039_large_estimate_half_life_by_", fs)]] > 0)

  tibble(
    fitting_strategy = fs,
    R_small_hl = get_spearman_r(df,
                                paste0("MS96_small_estimate_half_life_by_",   fs),
                                paste0("MS1039_small_estimate_half_life_by_", fs)),
    R_large_hl = get_spearman_r(df,
                                paste0("MS96_large_estimate_half_life_by_",   fs),
                                paste0("MS1039_large_estimate_half_life_by_", fs)),
    R_fc       = get_spearman_r(df,
                                paste0("MS96_hl_fc_by_",   fs),
                                paste0("MS1039_hl_fc_by_", fs))
  )
})

print(r_summary)

p_matrix <- plot_reproducibility_matrix(r_summary)
p_matrix
save_single(p_matrix, "../figures/replicate_reproducibility_matrix.png",
            width = 8, height = 4)
```

## Section 4 -- Averaged WT half-life and fold change (c1 only)

```{r s4}
df_hl_consensus <- df_hl_wide %>%
  mutate(
    WT_small_hl = rowMeans(across(c(MS96_small_estimate_half_life_by_c1,
                                    MS1039_small_estimate_half_life_by_c1)),
                           na.rm = TRUE),
    WT_large_hl = rowMeans(across(c(MS96_large_estimate_half_life_by_c1,
                                    MS1039_large_estimate_half_life_by_c1)),
                           na.rm = TRUE),
    WT_hl_fc    = log2(WT_large_hl / WT_small_hl)
  ) %>%
  dplyr::select(gene_id, WT_small_hl, WT_large_hl, WT_hl_fc) %>%
  filter(!is.na(WT_hl_fc), is.finite(WT_hl_fc))

nrow(df_hl_consensus)
summary(df_hl_consensus$WT_hl_fc)

ggplot(df_hl_consensus, aes(x = WT_hl_fc)) +
  geom_histogram(bins = 100, fill = "royalblue3", alpha = 0.7) +
  geom_vline(xintercept = 0, colour = "grey") +
  labs(x     = "log2 half-life fold change (large/small)",
       y     = "Number of genes",
       title = "mRNA half-life fold change in WT cells (large vs small, averaged across replicates)") +
  correlation_theme

ggsave("../figures/WT_hl_fc_distribution.png",
       width = 6, height = 5, dpi = 300)

write_tsv(df_hl_consensus, "../output_tables/df_WT_consensus_hl_fc.txt")
cat("Saved df_WT_consensus_hl_fc.txt --", nrow(df_hl_consensus), "genes\n")
```

## Section 5 -- Effect of SE filter on gene count and reproducibility

```{r s5}
df_c1_raw <- read_tsv("../output_tables/df_fit_output_c1.txt",
                       show_col_types = FALSE) %>%
  left_join(sample_info, join_by(experiment_strain_condition)) %>%
  filter(condition == "WT")

se_thresholds <- c(0.25, 0.5, 0.75, 1.0)

se_summary <- map_dfr(se_thresholds, function(thresh) {
  df <- df_c1_raw %>%
    filter(std.error_log_k < thresh) %>%
    pivot_wider(id_cols     = gene_id,
                names_from  = label,
                values_from = estimate_half_life,
                names_glue  = "{label}_hl") %>%
    mutate(
      MS96_hl_fc   = log2(MS96_large_hl / MS96_small_hl),
      MS1039_hl_fc = log2(MS1039_large_hl / MS1039_small_hl)) %>%
    filter(
      !is.na(MS96_small_hl),   !is.na(MS1039_small_hl),
      !is.na(MS96_large_hl),   !is.na(MS1039_large_hl),
      !is.na(MS96_hl_fc),      !is.na(MS1039_hl_fc),
      MS96_small_hl   > 0,     MS1039_small_hl   > 0,
      MS96_large_hl   > 0,     MS1039_large_hl   > 0,
      is.finite(MS96_hl_fc),   is.finite(MS1039_hl_fc))

  tibble(
    threshold  = thresh,
    filter     = "SE only",
    n_genes    = nrow(df),
    R_small_hl = get_spearman_r(df, "MS96_small_hl",  "MS1039_small_hl"),
    R_large_hl = get_spearman_r(df, "MS96_large_hl",  "MS1039_large_hl"),
    R_fc       = get_spearman_r(df, "MS96_hl_fc",      "MS1039_hl_fc")
  )
})

print(se_summary)

se_long <- se_summary %>%
  pivot_longer(cols      = c(R_small_hl, R_large_hl, R_fc),
               names_to  = "metric",
               values_to = "R") %>%
  mutate(metric = recode(metric,
                         R_small_hl = "Half-life small cells",
                         R_large_hl = "Half-life large cells",
                         R_fc       = "Fold change large/small"))

p_se_r <- ggplot(se_long, aes(x = threshold, y = R, colour = metric)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(
    "Half-life small cells"   = "orange",
    "Half-life large cells"   = "steelblue",
    "Fold change large/small" = "royalblue3")) +
  scale_x_continuous(breaks = se_thresholds) +
  labs(x      = "std.error_log_k threshold",
       y      = "Spearman R (MS96 vs MS1039)",
       title  = "Effect of SE filter on replicate reproducibility - c1 fitting",
       colour = NULL) +
  correlation_theme

p_se_n <- ggplot(se_summary, aes(x = threshold, y = n_genes)) +
  geom_line(linewidth = 0.8, colour = "royalblue3") +
  geom_point(size = 2, colour = "royalblue3") +
  scale_x_continuous(breaks = se_thresholds) +
  labs(x     = "std.error_log_k threshold",
       y     = "Number of genes retained",
       title = "Effect of SE filter on gene coverage - c1 fitting") +
  correlation_theme

p_se_r + p_se_n
save_pair(p_se_r + p_se_n,
          "../figures/SE_filter_effect_on_reproducibility.png")
```

## Section 6 -- Effect of minimum half-life filter on gene count and reproducibility

```{r s6}
min_hl_thresholds <- c(0, 3, 5, 10)

hl_summary <- map_dfr(min_hl_thresholds, function(min_hl) {
  df <- df_c1_raw %>%
    filter(std.error_log_k < 0.75) %>%
    pivot_wider(id_cols     = gene_id,
                names_from  = label,
                values_from = estimate_half_life,
                names_glue  = "{label}_hl") %>%
    mutate(
      MS96_hl_fc   = log2(MS96_large_hl / MS96_small_hl),
      MS1039_hl_fc = log2(MS1039_large_hl / MS1039_small_hl)) %>%
    filter(
      !is.na(MS96_small_hl),   !is.na(MS1039_small_hl),
      !is.na(MS96_large_hl),   !is.na(MS1039_large_hl),
      !is.na(MS96_hl_fc),      !is.na(MS1039_hl_fc),
      MS96_small_hl   > min_hl, MS1039_small_hl > min_hl,
      MS96_large_hl   > min_hl, MS1039_large_hl > min_hl,
      is.finite(MS96_hl_fc),   is.finite(MS1039_hl_fc))

  tibble(
    min_hl_threshold = min_hl,
    n_genes          = nrow(df),
    R_small_hl       = get_spearman_r(df, "MS96_small_hl",  "MS1039_small_hl"),
    R_large_hl       = get_spearman_r(df, "MS96_large_hl",  "MS1039_large_hl"),
    R_fc             = get_spearman_r(df, "MS96_hl_fc",      "MS1039_hl_fc")
  )
})

print(hl_summary)

hl_long <- hl_summary %>%
  pivot_longer(cols      = c(R_small_hl, R_large_hl, R_fc),
               names_to  = "metric",
               values_to = "R") %>%
  mutate(metric = recode(metric,
                         R_small_hl = "Half-life small cells",
                         R_large_hl = "Half-life large cells",
                         R_fc       = "Fold change large/small"))

p_hl_r <- ggplot(hl_long, aes(x = min_hl_threshold, y = R, colour = metric)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(
    "Half-life small cells"   = "orange",
    "Half-life large cells"   = "steelblue",
    "Fold change large/small" = "royalblue3")) +
  scale_x_continuous(breaks = min_hl_thresholds) +
  labs(x      = "Minimum half-life threshold (min)",
       y      = "Spearman R (MS96 vs MS1039)",
       title  = "Effect of minimum half life filter on replicate reproducibility - c1 fitting",
       colour = NULL) +
  correlation_theme

p_hl_n <- ggplot(hl_summary, aes(x = min_hl_threshold, y = n_genes)) +
  geom_line(linewidth = 0.8, colour = "royalblue3") +
  geom_point(size = 2, colour = "royalblue3") +
  scale_x_continuous(breaks = min_hl_thresholds) +
  labs(x     = "Minimum half-life threshold (min)",
       y     = "Number of genes retained",
       title = "Effect of minimum half life filter on gene coverage - c1 fitting") +
  correlation_theme

p_hl_r + p_hl_n
save_pair(p_hl_r + p_hl_n,
          "../figures/minHL_filter_effect_on_reproducibility.png")
```

## Section 7 -- Fold change vs half-life in small and large cells separately

```{r s7}
df_hl_consensus_75 <- df_c1_raw %>%
  filter(std.error_log_k < 0.75) %>%
  pivot_wider(id_cols     = gene_id,
              names_from  = label,
              values_from = estimate_half_life,
              names_glue  = "{label}_hl") %>%
  filter(
    !is.na(MS96_small_hl),  !is.na(MS1039_small_hl),
    !is.na(MS96_large_hl),  !is.na(MS1039_large_hl),
    MS96_small_hl  > 0,     MS1039_small_hl  > 0,
    MS96_large_hl  > 0,     MS1039_large_hl  > 0) %>%
  mutate(
    WT_small_hl = rowMeans(across(c(MS96_small_hl, MS1039_small_hl)), na.rm = TRUE),
    WT_large_hl = rowMeans(across(c(MS96_large_hl, MS1039_large_hl)), na.rm = TRUE),
    WT_hl_fc    = log2(WT_large_hl / WT_small_hl)
  ) %>%
  filter(!is.na(WT_hl_fc), is.finite(WT_hl_fc)) %>%
  dplyr::select(gene_id, WT_small_hl, WT_large_hl, WT_hl_fc)

nrow(df_hl_consensus_75)

p_fc_vs_hl <- scatter_cor_binned_pair(
  data          = df_hl_consensus_75,
  x_small       = "WT_small_hl",
  y_small       = "WT_hl_fc",
  x_large       = "WT_large_hl",
  y_large       = "WT_hl_fc",
  x_lab         = "Consensus half-life (min)",
  y_lab         = "log2 half-life fold change (large/small)",
  title_small   = "Small cells",
  title_large   = "Large cells",
  bins          = 30,
  show_diagonal = FALSE) &
  coord_cartesian(ylim = c(-4, 4))

p_fc_vs_hl
save_pair(p_fc_vs_hl, "../figures/fc_vs_halflife.png")
```

## Section 8 -- Effect of R-squared filter (with SE < 0.75) on reproducibility

```{r s8}
df_c1_rsq <- read_tsv("../output_tables/df_fit_output_c1_rsq.txt",
                       show_col_types = FALSE) %>%
  left_join(sample_info, join_by(experiment_strain_condition)) %>%
  filter(condition == "WT")

rsq_thresholds <- c(0.5, 0.7, 0.8, 0.9)

rsq_se_summary <- map_dfr(rsq_thresholds, function(thresh) {
  df <- df_c1_rsq %>%
    filter(std.error_log_k < 0.75,
           r_squared >= thresh) %>%
    pivot_wider(id_cols     = gene_id,
                names_from  = label,
                values_from = estimate_half_life,
                names_glue  = "{label}_hl") %>%
    mutate(
      MS96_hl_fc   = log2(MS96_large_hl / MS96_small_hl),
      MS1039_hl_fc = log2(MS1039_large_hl / MS1039_small_hl)) %>%
    filter(
      !is.na(MS96_small_hl),  !is.na(MS1039_small_hl),
      !is.na(MS96_large_hl),  !is.na(MS1039_large_hl),
      !is.na(MS96_hl_fc),     !is.na(MS1039_hl_fc),
      MS96_small_hl  > 0,     MS1039_small_hl > 0,
      MS96_large_hl  > 0,     MS1039_large_hl > 0,
      is.finite(MS96_hl_fc),  is.finite(MS1039_hl_fc))

  tibble(
    threshold  = thresh,
    filter     = "SE + R-squared",
    n_genes    = nrow(df),
    R_small_hl = get_spearman_r(df, "MS96_small_hl",  "MS1039_small_hl"),
    R_large_hl = get_spearman_r(df, "MS96_large_hl",  "MS1039_large_hl"),
    R_fc       = get_spearman_r(df, "MS96_hl_fc",      "MS1039_hl_fc")
  )
})

print(rsq_se_summary)

rsq_se_long <- rsq_se_summary %>%
  pivot_longer(cols      = c(R_small_hl, R_large_hl, R_fc),
               names_to  = "metric",
               values_to = "R") %>%
  mutate(metric = recode(metric,
                         R_small_hl = "Half-life small cells",
                         R_large_hl = "Half-life large cells",
                         R_fc       = "Fold change large/small"))

p_rsq_se_r <- ggplot(rsq_se_long, aes(x = threshold, y = R, colour = metric)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(
    "Half-life small cells"   = "orange",
    "Half-life large cells"   = "steelblue",
    "Fold change large/small" = "royalblue3")) +
  scale_x_continuous(breaks = rsq_thresholds) +
  labs(x      = "Minimum R-squared threshold (SE < 0.75 applied)",
       y      = "Spearman R (MS96 vs MS1039)",
       title  = "Effect of R-squared filter (with SE) on reproducibility - c1 fitting",
       colour = NULL) +
  correlation_theme

p_rsq_se_n <- ggplot(rsq_se_summary, aes(x = threshold, y = n_genes)) +
  geom_line(linewidth = 0.8, colour = "royalblue3") +
  geom_point(size = 2, colour = "royalblue3") +
  scale_x_continuous(breaks = rsq_thresholds) +
  labs(x     = "Minimum R-squared threshold (SE < 0.75 applied)",
       y     = "Number of genes retained",
       title = "Effect of R-squared filter (with SE) on gene coverage - c1 fitting") +
  correlation_theme

p_rsq_se_r + p_rsq_se_n
save_pair(p_rsq_se_r + p_rsq_se_n,
          "../figures/Rsq_SE_filter_effect_on_reproducibility.png")
```

## Section 9 -- R-squared only vs SE only vs SE + R-squared comparison

```{r s9}
rsq_only_summary <- map_dfr(rsq_thresholds, function(thresh) {
  df <- df_c1_rsq %>%
    filter(r_squared >= thresh) %>%
    pivot_wider(id_cols     = gene_id,
                names_from  = label,
                values_from = estimate_half_life,
                names_glue  = "{label}_hl") %>%
    mutate(
      MS96_hl_fc   = log2(MS96_large_hl / MS96_small_hl),
      MS1039_hl_fc = log2(MS1039_large_hl / MS1039_small_hl)) %>%
    filter(
      !is.na(MS96_small_hl),  !is.na(MS1039_small_hl),
      !is.na(MS96_large_hl),  !is.na(MS1039_large_hl),
      !is.na(MS96_hl_fc),     !is.na(MS1039_hl_fc),
      MS96_small_hl  > 0,     MS1039_small_hl > 0,
      MS96_large_hl  > 0,     MS1039_large_hl > 0,
      is.finite(MS96_hl_fc),  is.finite(MS1039_hl_fc))

  tibble(
    threshold  = thresh,
    filter     = "R-squared only",
    n_genes    = nrow(df),
    R_small_hl = get_spearman_r(df, "MS96_small_hl",  "MS1039_small_hl"),
    R_large_hl = get_spearman_r(df, "MS96_large_hl",  "MS1039_large_hl"),
    R_fc       = get_spearman_r(df, "MS96_hl_fc",      "MS1039_hl_fc")
  )
})

print(rsq_only_summary)

combined_summary <- bind_rows(
  se_summary,
  rsq_only_summary,
  rsq_se_summary)

p_compare <- ggplot(combined_summary,
                    aes(x = n_genes, y = R_fc,
                        colour = filter,
                        label  = round(threshold, 2))) +
  geom_point(size = 3) +
  geom_line() +
  geom_text(vjust = -0.8, size = 3.5) +
  scale_colour_manual(values = c(
    "SE only"        = "#ffa8cb",
    "R-squared only" = "#784283",
    "SE + R-squared" = "#e5000c")) +
  labs(x      = "Number of genes retained",
       y      = "Spearman R fold change (MS96 vs MS1039)",
       title  = "Filter comparison - fold change reproducibility vs gene coverage",
       colour = NULL) +
  correlation_theme

p_compare
save_single(p_compare,
            "../figures/filter_comparison.png",
            width = 8, height = 6)
```