---
title: "TF-enrichment-analysis"
author: "AC"
date: "2026-05-14"
output: word_document
---

```{r setup}
rm(list=ls())
library(MASS)
library(plotly)
library(broom)
library(corrr)
library(viridis)
library(tidyverse)
library(readxl)
#BiocManager::install("mitch")
library(mitch)
library(ggrepel)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) #assumes you're in scripts folder 
load('../../../../shared_code/r_functions/plotting_functions.rda')
getwd()
```

# Data wrangling
```{r cars}
PCA_genes <- read_tsv("../tables/PCA_genes_AC.txt", show_col_types=FALSE) # read table with principle components scores for all genes

file_path <- "~/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Swaffer lab documents/NGS_metadata/AC_combined_analysis/analysis/tables/Supp Table 3.xlsx"
file.exists(file_path)
excel_sheets(file_path) #Check sheet names

S3A <- read_excel(file_path, sheet = "Table-S3a")
S3C <- read_excel(file_path, sheet = "Table-S3c")
S3E <- read_excel(file_path, sheet = "Table-S3e")


pca_mat <- PCA_genes %>%
  select(gene_id, PC1, PC2) # filter for PC1 and PC2, as they are the components where most of the variation is explained
 
```

# Check gene overlap
```{r}
# Check genes overlap between S3A and PCA

common_genes <- intersect(pca_mat$gene_id, S3A$gene_id)
length(common_genes) #4370 genes overlap

#Filter all tables
pca_mat <- pca_mat %>%
  filter(gene_id %in% common_genes)

# all four data frames (pca_mat, S3A, S3C, and S3E) contain only the genes that were shared between pca_mat and S3A.
S3A <- S3A %>% filter(gene_id %in% common_genes)
S3C <- S3C %>% filter(gene_id %in% common_genes)
S3E <- S3E %>% filter(gene_id %in% common_genes)
```

# Build TF gene sets for enrichment
```{r}
# Table A (binding yes or no)
tf_bound <- S3A %>%
  pivot_longer(-gene_id, names_to = "TF", values_to = "bound") %>% #convert to long format but keep gene_id
  filter(bound == 1) %>% #filter for TFs that bind
  group_by(TF) %>% #group by TF
  summarise(genes = list(unique(gene_id)), .groups = "drop") %>% #For each TF create a list of genes. e TF1 -> c(Gene 1, gene 2) etc
  deframe() # convert from data frame to a list

# Table C, TF-activated genes (TF depletion ↓ expression → TF normally activates gene)
tf_activated <- S3C %>%
  select(-Kmeans_clusters) %>%   # remove annotation columns
  pivot_longer(
    cols = -gene_id,
    names_to = "TF",
    values_to = "log2FC"
  ) %>%
  filter(!is.na(log2FC)) %>%
  filter(log2FC < 0) %>% #TF depleted < 0 equals repressed under this condition. Therefore, TF normally needed to activate gene
  group_by(TF) %>%
  summarise(
    genes = list(unique(gene_id)),
    .groups = "drop"
  ) %>%
  deframe()

# Table C, TF-repressed genes (TF depletion expression goes up → TF normally represses gene)
tf_repressed <- S3C %>%
  select(-Kmeans_clusters) %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "TF",
    values_to = "log2FC"
  ) %>%
  filter(!is.na(log2FC)) %>%
  filter(log2FC > 0) %>%
  group_by(TF) %>%
  summarise(
    genes = list(unique(gene_id)),
    .groups = "drop"
  ) %>%
  deframe()

# Table E, TF-direct activated genes 
tf_direct_activated <- S3E %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "TF",
    values_to = "log2FC"
  ) %>%
  filter(!is.na(log2FC)) %>%
  filter(log2FC < 0) %>%
  group_by(TF) %>%
  summarise(
    genes = list(unique(gene_id)),
    .groups = "drop"
  ) %>%
  deframe()

# Table E, TF-direct repressed genes 
tf_direct_repressed <- S3E %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "TF",
    values_to = "log2FC"
  ) %>%
  filter(!is.na(log2FC)) %>%
  filter(log2FC > 0) %>%
  group_by(TF) %>%
  summarise(
    genes = list(unique(gene_id)),
    .groups = "drop"
  ) %>%
  deframe()
```

#Create PCA matrix for Mitch, run enrichment, extract results for A ONLY
```{r}
ranked_genes <- pca_mat %>%
  column_to_rownames("gene_id") #gene IDs stored as row names (remove gene_id as column title)

#Enrichment for A
res_bound <- mitch_calc(
  ranked_genes,#take ranked genes data frame
  tf_bound, # take tf_bound
  minsetsize = 5, #TF binds min. 5 genes
  cores = 1, #CPU used
  priority = "effect" # rank enrichments based on effect size (alternative is p value)
)

#Extract results 
df_bound <- res_bound$enrichment_result %>%
  rename(TF = set) #rename TF as Set
```

#Plotting A PC1 vs PC2 by filtering according to most sign. p value
```{r}
labels_A <- df_bound %>%
  arrange(p.adjustMANOVA) %>%
  slice_head(n = 20)

Plot <- ggplot(df_bound, aes(x = s.PC1, y = s.PC2)) +

  geom_hline(yintercept = 0, linetype = 2, color = "grey70") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey70") +

  geom_point(aes(size = setSize, color = -log10(p.adjustMANOVA)), alpha = 0.8) +
  
  scale_size_continuous(
    breaks = c(100, 500, 1000, 2000)
  ) +

  geom_text_repel(data = labels_A, aes(label = TF), size = 4) +

  scale_color_viridis_c() +
  theme_bw() +

  labs(
    x = "PC1 enrichment",
    y = "PC2 enrichment",
    size = "No of target genes per TF",
    color = "-log10(Adj. p-value)",
  )
Plot
#ggsave(Plot, filename = "./figures/TFs_enrichment.png", dpi = 600, width = 8, height = 5)

```
# Another plot but here instead of filtering based on most sign. p-value (which is a combination of pC1 and PC2= MANOVA), I filter according to those TFs that drive most of the variation in PC1 and PC2 space. HERE I DON'T CARE ABOUT DIRECTION (pos. or neg) of my PCs. I am just looking at the largest PCs effects 
```{r}
#TFs driving PC1 (irrespective of whether they are positive or negative)
labels_PC1 <- df_bound %>%
  arrange(desc(abs(s.PC1))) %>%
  slice_head(n = 20)

#TFs driving PC2 (irrespective of whether they are positive or negative)
labels_PC2 <- df_bound %>%
  arrange(desc(abs(s.PC2))) %>%
  slice_head(n = 20)

#Combine them
labels_C <- bind_rows(
  labels_PC1,
  labels_PC2
) %>%
  distinct(TF, .keep_all = TRUE)

#Plot PC1 vs PC2
Plot_C <- ggplot(df_bound, aes(x = s.PC1, y = s.PC2)) +

  geom_hline(yintercept = 0, linetype = 2, color = "grey70") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey70") +

  geom_point(
    aes(size = setSize, color = -log10(p.adjustMANOVA)),
    alpha = 0.8
  ) +

  geom_text_repel(
    data = labels_C,
    aes(label = TF),
    size = 2.0
  ) +

  scale_size_continuous(
    breaks = c(100, 500, 1000, 2000)
  ) +

  scale_color_viridis_c() +
  theme_bw() +

  labs(
    x = "PC1 enrichment",
    y = "PC2 enrichment",
    size = "No of target genes per TF",
    color = "-log10(Adj. p-value)"
  )
Plot_C
#ggsave(Plot_C, filename = "./figures/TFs_enrichment.png", dpi = 600, width = 8, height = 5)

```
#Another plot to show main drivers of PC1 and PC2 separatly (top 20) (not based on p value but on s.PC1 and s.PC2). HERE I CARE ABOUT THE DIRECTION OF PC1/2 + and - (used for poster 260625)
```{r}
top_pos_PC1 <- df_bound %>%
  arrange(desc(s.PC1)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 positive")

top_neg_PC1 <- df_bound %>%
  arrange(s.PC1) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 negative")

top_pos_PC2 <- df_bound %>%
  arrange(desc(s.PC2)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 positive")

top_neg_PC2 <- df_bound %>%
  arrange(s.PC2) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 negative")

labels <- bind_rows(      #create dataframe combining all top 20s
  top_pos_PC1,
  top_neg_PC1,
  top_pos_PC2,
  top_neg_PC2
) %>%
  distinct(TF, .keep_all = TRUE)

#create a new table to specify which TFs are regulated by TORC1/PKA
tf_regulation <- tibble::tibble(
  TF = c("IFH1","FHL1","SFP1","MSN2","MSN4","GCN4","RAP1","GLN3",
         "STB1","MIG1","MIG2","ADR1"),
  regulation = c("TORC1/PKA","TORC1/PKA","TORC1/PKA",
                 "TORC1/PKA","TORC1/PKA",
                 "TORC1","TORC1",
                 "TORC1",
                 "TORC1/PKA",
                 "PKA","PKA","PKA")
)

labels_new <- labels %>% 
  left_join(tf_regulation, by = "TF") %>%
  mutate(label = ifelse(is.na(regulation), TF, regulation)) # create a new column called label, where targets of TORC1/PKA are called like we told it to, while the other TFs keep their name

Plot5 <- ggplot(df_bound, aes(x = s.PC1, y = s.PC2)) +

  geom_hline(
    yintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_vline(
    xintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_point(
    aes(
      size = setSize,
      color = -log10(p.adjustMANOVA)
    ),
    alpha = 0.8
  ) +

  geom_text_repel(
    data = labels_new,
    aes(label = regulation),
    size = 3
  ) +

  scale_size_continuous(
    breaks = c(100, 500, 1000, 2000)
  ) +

  scale_color_viridis_c() +

  theme_bw() +

  labs(
    x = "PC1 enrichment",
    y = "PC2 enrichment",
    size = "No. of target genes per TF",
    color = "-log10(Adj. p-value)"
  )
Plot5
ggsave(Plot5, filename = "./figures/TFs_enrichment_2.png", dpi = 600, width = 8, height = 5)
```

# Enrichment for Table C and E
```{r}
#Enrichment for C ACTIVATED 
res_activated <- mitch_calc(
  ranked_genes,
  tf_activated,
  minsetsize = 5, #TF binds min. 5 genes
  cores = 1,
  priority = "effect"
)
#Extract results 
df_activated <- res_activated$enrichment_result %>%
  rename(TF = set)

#Enrichment for C REPRESSED
res_repressed <- mitch_calc(
  ranked_genes,
  tf_repressed,
  minsetsize = 5, #TF binds min. 5 genes
  cores = 1,
  priority = "effect"
)
#Extract results 
df_repressed <- res_repressed$enrichment_result %>%
  rename(TF = set)

#Enrichment for E DIRECT ACTIVATED 
res_direct_activated <- mitch_calc(
  ranked_genes,
  tf_direct_activated,
  minsetsize = 5, #TF binds min. 5 genes
  cores = 1,
  priority = "effect"
)
#Extract results 
df_direct_activated <- res_direct_activated$enrichment_result %>%
  rename(TF = set)

#Enrichment for E DIRECT REPRESSED
res_direct_repressed <- mitch_calc(
  ranked_genes,
  tf_direct_repressed,
  minsetsize = 5, #TF binds min. 5 genes
  cores = 1,
  priority = "effect"
)
#Extract results 
df_direct_repressed <- res_direct_repressed$enrichment_result %>%
  rename(TF = set)
```

#Plotting E df_direct activated
```{r}
top_pos_PC1_E <- df_direct_activated %>%
  arrange(desc(s.PC1)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 positive")

top_neg_PC1_E <- df_direct_activated %>%
  arrange(s.PC1) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 negative")

top_pos_PC2_E <- df_direct_activated %>%
  arrange(desc(s.PC2)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 positive")

top_neg_PC2_E <- df_direct_activated %>%
  arrange(s.PC2) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 negative")

labels_1 <- bind_rows(      #create dataframe combining all top 20s
  top_pos_PC1_E,
  top_neg_PC1_E,
  top_pos_PC2_E,
  top_neg_PC2_E
) %>%
  distinct(TF, .keep_all = TRUE)

Plot6 <- ggplot(df_direct_activated, aes(x = s.PC1, y = s.PC2)) +

  geom_hline(
    yintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_vline(
    xintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_point(
    aes(
      size = setSize,
      color = -log10(p.adjustMANOVA)
    ),
    alpha = 0.8
  ) +

  geom_text_repel(
    data = labels_1,
    aes(label = TF),
    size = 3
  ) +

  scale_size_continuous(
    breaks = c(10, 25, 50, 100)
  ) +

  scale_color_viridis_c() +

  theme_bw() +

  labs(
    x = "PC1 enrichment",
    y = "PC2 enrichment",
    size = "No. of target genes per TF",
    color = "-log10(Adj. p-value)"
  )
Plot6
ggsave(Plot6, filename = "./figures/TFs_enrichment_E_A.png", dpi = 600, width = 8, height = 5)
```

#Plotting E df_direct repressed
```{r}
top_pos_PC1_E_2 <- df_direct_repressed %>%
  arrange(desc(s.PC1)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 positive")

top_neg_PC1_E_2 <- df_direct_repressed %>%
  arrange(s.PC1) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 negative")

top_pos_PC2_E_2 <- df_direct_repressed %>%
  arrange(desc(s.PC2)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 positive")

top_neg_PC2_E_2 <- df_direct_repressed %>%
  arrange(s.PC2) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 negative")

labels_2 <- bind_rows(      #create dataframe combining all top 20s
  top_pos_PC1_E_2,
  top_neg_PC1_E_2,
  top_pos_PC2_E_2,
  top_neg_PC2_E_2
) %>%
  distinct(TF, .keep_all = TRUE)

Plot7 <- ggplot(df_direct_repressed, aes(x = s.PC1, y = s.PC2)) +

  geom_hline(
    yintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_vline(
    xintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_point(
    aes(
      size = setSize,
      color = -log10(p.adjustMANOVA)
    ),
    alpha = 0.8
  ) +

  geom_text_repel(
    data = labels_2,
    aes(label = TF),
    size = 3
  ) +

  scale_size_continuous(
    breaks = c(10, 25, 50, 100)
  ) +

  scale_color_viridis_c() +

  theme_bw() +

  labs(
    x = "PC1 enrichment",
    y = "PC2 enrichment",
    size = "No. of target genes per TF",
    color = "-log10(Adj. p-value)"
  )
Plot7
ggsave(Plot7, filename = "./figures/TFs_enrichment_E_R.png", dpi = 600, width = 8, height = 5)

```
#Plotting C df_activated
```{r}
#Plotting C df_activated

top_pos_PC1_C <- df_activated %>%
  arrange(desc(s.PC1)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 positive")

top_neg_PC1_C <- df_activated %>%
  arrange(s.PC1) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 negative")

top_pos_PC2_C <- df_activated  %>%
  arrange(desc(s.PC2)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 positive")

top_neg_PC2_C <- df_activated %>%
  arrange(s.PC2) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 negative")

labels_C <- bind_rows(      #create dataframe combining all top 20s
  top_pos_PC1_C,
  top_neg_PC1_C,
  top_pos_PC2_C,
  top_neg_PC2_C
) %>%
  distinct(TF, .keep_all = TRUE)

Plot8 <- ggplot(df_activated, aes(x = s.PC1, y = s.PC2)) +

  geom_hline(
    yintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_vline(
    xintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_point(
    aes(
      size = setSize,
      color = -log10(p.adjustMANOVA)
    ),
    alpha = 0.8
  ) +

  geom_text_repel(
    data = labels_C,
    aes(label = TF),
    size = 3
  ) +

  scale_size_continuous(
    breaks = c(10, 25, 50, 100)
  ) +

  scale_color_viridis_c() +

  theme_bw() +

  labs(
    x = "PC1 enrichment",
    y = "PC2 enrichment",
    size = "No. of target genes per TF",
    color = "-log10(Adj. p-value)"
  )
Plot8
ggsave(Plot8, filename = "./figures/TFs_enrichment_C_A.png", dpi = 600, width = 8, height = 5)
```
#Plotting C df_repressed
```{r}
#Plotting C df_repressed

top_pos_PC1_C_2 <- df_repressed %>%
  arrange(desc(s.PC1)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 positive")

top_neg_PC1_C_2 <- df_repressed %>%
  arrange(s.PC1) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC1 negative")

top_pos_PC2_C_2 <- df_repressed  %>%
  arrange(desc(s.PC2)) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 positive")

top_neg_PC2_C_2 <- df_repressed %>%
  arrange(s.PC2) %>%
  slice_head(n = 20) %>%
  mutate(driver = "PC2 negative")

labels_C_2 <- bind_rows(      #create dataframe combining all top 20s
  top_pos_PC1_C_2,
  top_neg_PC1_C_2,
  top_pos_PC2_C_2,
  top_neg_PC2_C_2
) %>%
  distinct(TF, .keep_all = TRUE)

Plot9 <- ggplot(df_repressed, aes(x = s.PC1, y = s.PC2)) +

  geom_hline(
    yintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_vline(
    xintercept = 0,
    linetype = 2,
    color = "grey70"
  ) +

  geom_point(
    aes(
      size = setSize,
      color = -log10(p.adjustMANOVA)
    ),
    alpha = 0.8
  ) +

  geom_text_repel(
    data = labels_C_2,
    aes(label = TF),
    size = 3
  ) +

  scale_size_continuous(
    breaks = c(10, 25, 50, 100)
  ) +

  scale_color_viridis_c() +

  theme_bw() +

  labs(
    x = "PC1 enrichment",
    y = "PC2 enrichment",
    size = "No. of target genes per TF",
    color = "-log10(Adj. p-value)"
  )
Plot9
ggsave(Plot9, filename = "./figures/TFs_enrichment_C_R.png", dpi = 600, width = 8, height = 5)
```
